### 2024-02-21 article.tex
Initial Overleaf Import

\documentclass[pdflatex]{sn-jnl}
\jyear{2024}
\usepackage{multibib}
\newcites{m}{Methods References}
\usepackage[superscript]{cite}
\usepackage{caption}
\bibliographystyle{unsrt}
\bibliographystylem{unsrt} 
\raggedbottom

% Remove numbering from sections and subsections, as requested in decision email.
\setcounter{secnumdepth}{0}

\newcommand{\yohai}[1]{{\textcolor{red}{#1}}}
\newcommand{\neri}[1]{{\textcolor{cyan}{#1}}}

\begin{document}

\title[Running Title]{Using past seismicity to predict the magnitude of future earthquakes \newline or \newline Does an earthquake “know” how big it’s going to be? A neural-net aided study}

\author[1,2]{\fnm{Neri} \sur{Berman}}\email{neriberman@gmail.com}
\author[2]{\fnm{Oleg} \sur{Zlydenko}}
\author[2]{\fnm{Oren} \sur{Gilon}}
\author[1]{\fnm{Yohai} \sur{Bar-Sinai}}\email{ybarsinai@gmail.com}

\affil[1]{\orgdiv{Department of Physics}, \orgname{Tel-Aviv University}, \orgaddress{\city{Tel-Aviv}, \country{Israel}}}
\affil[2]{\orgdiv{Google Research}, \orgname{Google}, \orgaddress{\city{Tel-Aviv}, \country{Israel}}}

\abstract{
Earthquake occurrence is notoriously difficult to predict. Salient features of the statistical properties of the timing and location of earthquakes, i.e. spatiotemporal clustering, can be relatively well captured by point-process models. However, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts, both due to the lack of information about the fault condition and to the inherent complexity of rupture dynamics. Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events, besides the time-independent Gutenberg Richter (GR) distribution, which describes the marginal distribution over wide regions and long times. This approach implicitly assumes that earthquake magnitudes are independent of seismic history and are identically distributed. In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history. We present a neural network-based model for probabilistic forecasting of future magnitudes based on cataloged properties: hypocentral locations, occurrence times and magnitudes of past earthquakes. Our history-dependent model outperforms stationary and quasi-stationary state of the art GR-based benchmarks, in real catalogs in Southern California, Japan and New-Zealand.  This demonstrates that earthquake catalogs contain information about the magnitude of future earthquakes, prior to their occurrence. We conclude by proposing methods to apply the model in characterization of the preparatory phase of earthquakes, and in operational hazard alert and earthquake forecasting systems.
}

\keywords{}

\maketitle

\neri{comment by Neri} \newline
\yohai{comment by Yohai}

\section{Introduction} \label{sec:introduction}

Earthquake forecasting is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes \cite{geller_earthquakes_1997, bernard_earthquake_1999} (cite nature debate on earthquake prediction? EQs cannot be predicted Kagen?). Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes. This includes various characteristics of spatial and temporal clustering (cite Omori, Kagan, benzion \& Zaliapin, many others, ), which can be incorporated into stochastic prediction schemes (ETAS papers, Hardebeck etal Ann Rev 2024, Meade on NN aftershock location, some CFS paper). These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence. These spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability (cite and read Rathbun, 1993; Schoenberg, 2004). That is, while seismic history is used to inform the location and rate of future earthquakes, i.e.~by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events is determined by drawing values from the Gutenberg-Richter (GR) distribution or its variants (Gutenberg and Richter, 1944;  a kagen tapered GR, some spatiotemporal variation of GR, Ogata's magnitude prediction variational GR paper as well?). Importantly, this distribution does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events. %Such stochastic models, coupled with the separability assumption, imply that earthquake magnitude emergence is a stochastic process, independent of seismic history and spatial distribution, suggesting

\neri{This entire paragraph going on about SOC, perhaps to long and should be inserted as a sentence somewhere?}
This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice. On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff). These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics. In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).

While it is intelectually compelling, the assumption of separability is not necessarily true. Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field, which in turn depend on slip history in a deterministic manner, albeit intangibly complicated. Therefore, the assumption of magnitude separability is not a priori justified, and its validity should be tested empirically. Information about an earthquake magnitude exists before its initiation, but may be inaccessible using our current experimental or analytic tools. In this work we directly address this question, and ask whether some information about the magnitude of a given earthquake can be extracted from regional seismic history.

\neri{(short literature review)}

Here we address this question directly by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and log term seismic history prior to its occurrence. Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model. Thus, we separate out the task of predicting the nucleation of events and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of an earthquake is extractable from cataloged properties alone.

This contribution has two important consequences. From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable. Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction. This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.

% paradigm of earthquake  This both   . We assert that the standard hypocenter catalog alone contains sufficient information to predict future earthquake magnitudes with better-than-random accuracy. This contradicts the common view that such catalogs represent a stochastic process of earthquake magnitude occurrence (cite, perhaps Beroza etal Nat Comm 2021? their claim is that you need a ML generated catalog better predict earthquake properties). The presented results demonstrate that incorporating spatial features and seismic history into magnitude forecasting yields an information gain, thereby answering the question posed above.

We construct an encoder-decoder neural network (NN). Our model uses as input a hypocenter catalog to construct a latent representation of the seismic state at a specific timestamp. Then, the latent representation is passed, together with  a space-time coordinate query of a future earthquake, into a decoder that outputs a parameterized probability density function (pdf) of the earthquake. See \ref{fig:architecture} and caption for detailed description of the neural network's architecture. We optimize the model parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
\begin{equation}
    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i} \left( m_i \right) \right) } \rangle\ ,
    \label{eq:loss_function}
\end{equation}
where $p_{\textbf{x}_i, t_i}$ is the probability function returned by the model for the space-time coordinates $\textbf{x}_i, t_i$ (i.e. the ``query coordinates''), $m_i$ the corresponding to the actual magnitude of the $i^{th}$ earthquake and $\langle \cdot\rangle$ stands for the empirical average over the data set. We use the \yohai{kamawarakskakaskasasi} (cite, refer to methods/SI) parameterized family of probability distributions, which is defined by \yohai{N} parameters. This family smoothly can interpolate between exponential distribution, like GR, and localized distributions whose mass is concentrated around a specific value. This allows the model to output both an ``ignorant'' prediction, essentially resembling the GR distribution, and more confident predictions localized around a given magnitude.

We train the model on four distinct earthquake catalogs to assess the performance across diverse seismogenic regions: Hauksson et al. (cite) and SCSN (cite) for Southern California, GeoNet (cite) for New Zealand, and the JMA catalog (cite) for [Region 1, Region 2...] (refer to Supplementary Figure and Table for details). While all four catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.? cite?) and exhibit varying data quality. A separate model is trained for each region, with identical loss function and parameterization of the pdf.

Model performance is evaluated over a time span that was not included in training. The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence. \neri{I think this is a good place to add that we conclude that there is some preparatory phase.}
%This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation. Examining these results across multiple catalogs and regions highlights the robustness of our methodology and the generalizability of our conclusion, paving the way for further exploration of earthquake predictability.

\begin{figure}[h!]
	\centering
        \includegraphics[width=1\textwidth]{figures/intro_fig.pdf}
	\caption{
 Architecture of the magnitude prediction neural network model. \textbf{a} The input catalog is encoded with a set of predefined feature functions at multiple timestamps. These features are then used as an input to an LSTM network to create an encoded latent representation of the catalog. The latent representation is combined with a space-time query and fed into a fully connected neural network for decoding the data. The decoding network outputs a set of parameters defining a probability density function for each time-space query coordinates. \textbf{b} The pdfs from 100 random samples from the test set of the Hauksson catalog, color coded by the magnitude of the earthquake that has occurred in the corresponding query coordinates. The black dashed line shows Gutenberg Richter distribution, fitted for the train set.
}
\label{fig:architecture}
\end{figure}

\section{Results} \label{sec:results}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/raw_results_hauksson.pdf}
    \caption{
        Caption.
    }
    \label{fig:architecture}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/info_over_time_hauksson.pdf}
    \caption{
        Caption.
    }
    \label{fig:architecture}
\end{figure}

\begin{figure}[h!]
    \centering
    \fbox{
        \rule{1\textwidth}{0pt}\rule{0pt}{8ex}
        FIGURE OF SPATIAL DISTRIBUTION OF GAIN.
    }
    \caption{
        Caption.
    }
    \label{fig:architecture}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.4\textwidth]{figures/ROC_PR_hauksson.pdf}
    \caption{
        Caption.
    }
    \label{fig:architecture}
\end{figure}

% There are two bibliographies in Nature papers. One for the main text, and one for the Methods and Extended Data sections. The numbering is sequential, meaning that the reference section for the Methods and Extended Data section starts after the last number from the main reference section. References should not appear in both sections. Any reference used in Methods or Extended Data that also appears in the main bibliography should *only* appear in the main bibliography.
\let\oldbibliography\thebibliography
\renewcommand{\thebibliography}[1]{%
  \oldbibliography{#1}%
  \setlength{\itemsep}{10pt}%
}
% \bibliography{bibliography}
% \bibliography{bib-article.bib}
\newpage
\bibliography{Magnitude_prediction_paper}

\let\oldthebibliography=\thebibliography
\let\oldendthebibliography=\endthebibliography
\renewenvironment{thebibliography}[1]{
    \oldthebibliography{#1}
    % The number here (34 is an example) is the number of references in the main bibliography, and thus defines the starting number of the first reference in the Methods and Extended Data bibliography.
    \setcounter{enumiv}{34}
}{\oldendthebibliography}

% Figure legends appear after the text, not placed in the text. Do not include the actual image files in the article. Images are submitted as separate files. During the first submission to the journal, you can include the images in the article file for readability, but if you pass the reviews, they will want the images removed from the main article.
\newpage
\unnumbered

\unnumbered
\section{Methods}
\subsection{AI Model}
\textbf{Grey's Note} \textit{My understanding is that you can write the methods section just how you would a normal methods section for a normal scientific article.}

\section*{Data Availability}
This is a required section. Guidelines for data availability: \url{https://www.nature.com/documents/nr-data-availability-statements-data-citations.pdf}.

\section*{Code Availability}
\textbf{Grey's Note} \textit{This is a required section. If your article is about AI or ML, the editor will ask you to make the weights of a trained model available.}

\newpage
\renewcommand\refname{Methods References}
\begin{thebibliography}{10}

\bibitem{kratzert2019towards}
Frederik Kratzert, Daniel Klotz, Guy Shalev, G{\"u}nter Klambauer, Sepp
  Hochreiter, and Grey~S Nearing.
\newblock Towards learning universal, regional, and local hydrological
  behaviors via machine learning applied to large-sample datasets.
\newblock {\em Hydrology and Earth System Sciences}, 23(12):5089--5110, 2019.

\bibitem{klotz2022uncertainty}
Daniel Klotz, Frederik Kratzert, Martin Gauch, Alden~K Sampson, Johannes
  Brandstetter, G{\"u}nter Klambauer, Sepp Hochreiter, and Grey~S Nearing.
\newblock Uncertainty estimation with deep learning for rainfall--runoff
  modeling.
\newblock {\em Hydrology and Earth System Sciences}, 26(6):1673--1693, 2022.

\bibitem{twb2023gdp}
IBRD-IDA.
\newblock {The World Bank Data: Current US\$}.
\newblock https://data.worldbank.org/indicator/NY.GDP.MKTP.CD, 2023.
\newblock Accessed: 2023-06-04.

\end{thebibliography}

\newpage
\section*{Acknowledgements}
Acknowledgements should be brief, and should not include thanks to anonymous referees and editors, inessential words, or effusive comments. A person can be thanked for assistance, not “excellent” assistance, or for comments, not “insightful” comments, for example. Acknowledgements can contain grant and contribution numbers.

\section*{Author Contributions}
Author Contributions: Authors are required to include a statement to specify the contributions of each co-author. The statement can be up to several sentences long, describing the tasks of individual authors referred to by their initials. See the authorship policy page for further explanation and examples.

\section*{Author Information}
\textbf{Grey's Note} \textit{Two things are required in this Author Information section: (1) A statement about competing interests. I have no advice about what constitutes a competing interest, you will have to read about it and make your own decision. (2) A clear statement about who to contact with question about the paper. An example is below.}

The authors declare no competing interests. Please contact either the first author (Grey Nearing; \href{mailto:nearing@google.com}{nearing@google.com}) or Avinatan Hassidim; \href{mailto:avinatan@google.com}{avinatan@google.com}) for correspondence and requests, including questions regarding reprints and permissions.

\newpage
\section*{Extended Data}
\textbf{Grey's Note} \textit{You get up to 10 items in the Extended Data section. All Tables and Figures should be enumerated as "Extended Data Figure 1", "Extended Data Table 1", etc. These must also be referenced the same way in the text. Extended Data tables and figures may be referenced in the main article and/or in the Methods section. I am doing the in-text references to these ED items manually, instead of trying to override the table numbering schemed in Latex. Please update this template to automate that process if you feel like doing that.}

\textit{You can use Latex formatting for tables, but the journal will eventually require that your tables be submitted as images. Yes, this also struck me as unusual.}

\textit{Each ED figure should be on a separate page.}

\textit{Finally, this note is here in the ED section, but in your paper, the ED section should not have any text except figure and table captions/legends.}

\end{document}

### 2024-02-22 article.tex
Edit introduction.

@@
 
-\title[Running Title]{Using past seismicity to predict the magnitude of future earthquakes \newline or \newline Does an earthquake “know” how big it’s going to be? A neural-net aided study}
+\title[Running Title]{Does an earthquake “know” how big it’s going to be? A neural-net aided study} % 2nd option: Using past seismicity to predict the magnitude of future earthquakes
 
@@
 
-Earthquake forecasting is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes \cite{geller_earthquakes_1997, bernard_earthquake_1999} (cite nature debate on earthquake prediction?
-EQs cannot be predicted Kagen?).
+Earthquake forecasting is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes \cite{bernard_earthquake_1999, geller_earthquakes_1997}.
 Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes.
-This includes various characteristics of spatial and temporal clustering (cite Omori, Kagan, benzion \& Zaliapin, many others, ), which can be incorporated into stochastic prediction schemes (ETAS papers, Hardebeck etal Ann Rev 2024, Meade on NN aftershock location, some CFS paper).
+This includes various characteristics of spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}, which can be incorporated into stochastic prediction schemes \cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}.
 These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence.
-These spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability (cite and read Rathbun, 1993; Schoenberg, 2004).
-That is, while seismic history is used to inform the location and rate of future earthquakes, i.e.~by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events is determined by drawing values from the Gutenberg-Richter (GR) distribution or its variants (Gutenberg and Richter, 1944;  a kagen tapered GR, some spatiotemporal variation of GR, Ogata's magnitude prediction variational GR paper as well?).
+Furthermore, these spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability \cite{schoenberg_testing_2004}.
+That is, while seismic history is used to inform the location and rate of future earthquakes, i.e.
+by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events prediction often relies on drawing values from the Gutenberg-Richter (GR) distribution or its variants \cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}.
 Importantly, this distribution does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events.
-%Such stochastic models, coupled with the separability assumption, imply that earthquake magnitude emergence is a stochastic process, independent of seismic history and spatial distribution, suggesting
 
-\neri{This entire paragraph going on about SOC, perhaps to long and should be inserted as a sentence somewhere?}
-This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice.
+% This might be something to remove to shorten
+% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice.
 On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff).
@@
 
-While it is intelectually compelling, the assumption of separability is not necessarily true.
-Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field, which in turn depend on slip history in a deterministic manner, albeit intangibly complicated.
-Therefore, the assumption of magnitude separability is not a priori justified, and its validity should be tested empirically.
-Information about an earthquake magnitude exists before its initiation, but may be inaccessible using our current experimental or analytic tools.
-In this work we directly address this question, and ask whether some information about the magnitude of a given earthquake can be extracted from regional seismic history.
-
-\neri{(short literature review)}
-
-Here we address this question directly by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and log term seismic history prior to its occurrence.
+While assuming that an earthquake's magnitude is independent of the seismic history is popular sue to its simplicity and its ease of use in earthquake modeling, it is not necessarily true.
+Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field.
+These in turn depend on slip history in a deterministic manner, albeit intangibly complicated.
+Therefore, the assumption of magnitude separability, i.e.
+its independent from seismic history, is not a priory justified, and its validity should be tested empirically.
+Though information about an earthquake magnitude may exists before its initiation, implying dependence on seismic history and non-separability, it has yet to be shown accessible using our current experimental or analytic tools.
+Current research on magnitude prediction has failed to show a universal and replicable advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}, by that leaving the question of accessible information prior to an earthquake's nucleation withstanding.
+\newline
+In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history.
+This is done by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and long term seismic history prior to its occurrence.
 Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model.
@@
 This contribution has two important consequences.
-From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable.
+From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}.
 Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction.
@@
 
-% paradigm of earthquake  This both   .
-We assert that the standard hypocenter catalog alone contains sufficient information to predict future earthquake magnitudes with better-than-random accuracy.
-This contradicts the common view that such catalogs represent a stochastic process of earthquake magnitude occurrence (cite, perhaps Beroza etal Nat Comm 2021?
-their claim is that you need a ML generated catalog better predict earthquake properties).
-The presented results demonstrate that incorporating spatial features and seismic history into magnitude forecasting yields an information gain, thereby answering the question posed above.
+\neri{Is a literature review nescessary here or is it ingested enough in the text itself? I'm pretty satisfied with how the paragraphs up to now are written.}\newline
+\neri{The entire paragraph going on about SOC, perhaps to long and should be inserted as a sentence somewhere in the literature review? can be seem as comment in source code.}
+% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice.
+On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff).
+These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition.
+Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics.
+In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).
 
-We construct an encoder-decoder neural network (NN).
-Our model uses as input a hypocenter catalog to construct a latent representation of the seismic state at a specific timestamp.
-Then, the latent representation is passed, together with  a space-time coordinate query of a future earthquake, into a decoder that outputs a parameterized probability density function (pdf) of the earthquake.
-See \ref{fig:architecture} and caption for detailed description of the neural network's architecture.
+We construct a neural network (NN) model that encodes the seismic history of time series from a hypocentral catalog and produces an estimate of the possible magnitudes at a queried time and place.
+The constructed NN uses a long-short term memory (LSTM) architecture to model the seismic history up to a given time.
+This latent representation is then passed together with  a space-time coordinate query of a future earthquake into a fully connected neural network (FCNN) that produces a parametrized probability density function (PDF) of the possible magnitudes at the quarried space- time coordinates.
+This procedure is depicted in Fig.
+\ref{fig:intro_fig}a depicts the task as described here.
+A detailed description of the model's architecture is given in the methods section.
 We optimize the model parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
@@
 the ``query coordinates''), $m_i$ the corresponding to the actual magnitude of the $i^{th}$ earthquake and $\langle \cdot\rangle$ stands for the empirical average over the data set.
-We use the \yohai{kamawarakskakaskasasi} (cite, refer to methods/SI) parameterized family of probability distributions, which is defined by \yohai{N} parameters.
-This family smoothly can interpolate between exponential distribution, like GR, and localized distributions whose mass is concentrated around a specific value.
+We use the a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} parameterized family of probability distributions stretched to a relevant domain.
+The final distribution used is defined by 5 parameters.
+This family smoothly can interpolate between a decaying distribution (resembling the GR distribution), and localized distributions whose mass is concentrated around a specific value.
 This allows the model to output both an ``ignorant'' prediction, essentially resembling the GR distribution, and more confident predictions localized around a given magnitude.
 
-We train the model on four distinct earthquake catalogs to assess the performance across diverse seismogenic regions: Hauksson et al.
-(cite) and SCSN (cite) for Southern California, GeoNet (cite) for New Zealand, and the JMA catalog (cite) for [Region 1, Region 2...] (refer to Supplementary Figure and Table for details).
-While all four catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.?
+We train the model on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: Hauksson et al.
+(cite) for Southern California, GeoNet (cite) for New Zealand, and the JMA catalog (cite) for [Region 1, Region 2...].
+While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.?
 cite?) and exhibit varying data quality.
-A separate model is trained for each region, with identical loss function and parameterization of the pdf.
+A separate model is trained for each region, with identical loss function and parameterization of the PDF.
 
@@
 	\caption{
- Architecture of the magnitude prediction neural network model.
-\textbf{a} The input catalog is encoded with a set of predefined feature functions at multiple timestamps.
-These features are then used as an input to an LSTM network to create an encoded latent representation of the catalog.
-The latent representation is combined with a space-time query and fed into a fully connected neural network for decoding the data.
-The decoding network outputs a set of parameters defining a probability density function for each time-space query coordinates.
-\textbf{b} The pdfs from 100 random samples from the test set of the Hauksson catalog, color coded by the magnitude of the earthquake that has occurred in the corresponding query coordinates.
-The black dashed line shows Gutenberg Richter distribution, fitted for the train set.
+ Caption.
 }
-\label{fig:architecture}
+\label{fig:intro_fig}
 \end{figure}
@@
 \section{Results} \label{sec:results}
+
+Focusing on existing events within the catalog due to our model's requirement for labeled data, our analysis utilizes the space-time coordinates of earthquakes as input.
+This necessitates excluding queries about arbitrary spatiotemporal points during training and evaluation, as these processes require knowledge of the true magnitude for each event at its specific location and time.
+Consequently, the raw output for our analysis consists of the model-generated PDFs of possible magnitudes for each event's time and location.
+Figure \ref{fig:architecture}b showcases PDFs produced by our model, $p_{\textbf{x}_i, t_i}(m)$, for 100 randomly sampled event times and locations from the test set.
+Included as a baseline is the stationary GR predictor trained on the training set, representing its potential operational use (e.g., Tsuruoka et al., 2012).
+A clear trend emerges: PDFs predicted for higher-magnitude events (warm colors) exhibit a greater skew towards higher magnitudes compared to those for lower-magnitude events (cooler colors).
+Figure \ref{fig:output_pdfs}a further highlights this trend by displaying a selection of model-generated PDFs alongside the corresponding likelihood of retrieving the true magnitude at the time and location each PDF was estimated for, denoted by a [color] circle.
+
+Figure \ref{fig:output_pdfs}b focuses solely on the likelihood scores for retrieving the true magnitude, omitting the underlying PDFs used for their estimation.
+Additionally, it compares these scores against those generated by various benchmark models.
+Notably, our model demonstrates an advantage for most events exceeding magnitude $\sim 3$, while slightly under performing for lower-magnitude events.
+However, the overall mean log-likelihood (LL) score defined in \ref{eq:loss_function} (evaluated using the model-specific $p_{\textbf{x}_i, t_i}(\cdot)$ for each event) reveals a consistent advantage for our model across all datasets examined as summarized in \ref{tab:mean_ll}, where our model consistently obtains the lowest minus mean LL score in each column.
+
+The information gain achieved by our model for each event can be quantified by the difference in log-likelihood (LL) between our model's prediction and the baseline GR predictor:
+\begin{equation}
+    \Delta LL = \log{p_{\pmb{x_i}, t_i}^{(model)}(m_i)} - \log{p^{(GR)}(m_i)}
+    \label{eq:information_gain}
+\end{equation}
+This equation measures the number of bits of information gained by using our model compared to the GR benchmark.
+Analyzing this quantity per event across the entire test set reveals patterns in where our model achieves its advantage.
+\ref{fig:likelihood_over_time}a shows the Hauksson data test set event magnitudes at their occurrence times, colored according to the information gain.
+This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as a gray dashed line in the figure.
+The predominantly increasing trend in cumulative information gain suggests that our model is consistently advantageous as opposed to a concentrated advantage in specific periods.
+Notably, for the Hauksson data set, the most rapid information gain occurs during the aftershocks of the two Ridgecrest earthquakes.
+Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4$ bits, respectively, though these events are typically considered completely unpredictable.
+This observation leads to a significant corollary: at least some portion of the information gain likely originates from a physical precursory signal.
+This point will be further elaborated upon in the discussion section.
+\newline
+The observed increase in information gain persists throughout the entire test set for all regions examined in this study.
+This includes regions previously analyzed by (Ogata et al.
+SRL 2018), where no information gain was detected using the method employed at the time.
+Further analysis of \ref{fig:likelihood_over_time} applied to other regions is presented in the supplementary material.
+\newline
+While the information gain observed in pre- and mainshocks presented above is encouraging, a potential concern might arise regarding the aftershocks: could their information gain be an artifact of fitting to the temporal incompleteness of the catalog (cite somethong?)?
+To address this, we factor out this potential bias by recalculating the likelihood score of each event, $p(m_i)$ , conditioned on the magnitude exceeding a threshold, $\Tilde{m}$:
+
+\begin{equation}
+    p \left( m \vert m_i > \Tilde{m} \right) = \frac{p_{\pmb{x}_i, t_i}(m)} {\int_{\Tilde{m}}^{\infty} p(m') ,dm'}
+    \label{eq:conditioned_likelihood}
+\end{equation}
+
+For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method (Weimer and Weiss, 2000) within a window of 150 past and 150 future events.
+The resulting temporal incompleteness over time is displayed in \ref{fig:likelihood_over_time}b.
+Crucially, \ref{fig:likelihood_over_time}b also presents the magnitudes of the Ridgecrest sequence at their origin times, colored by the information gain of our model over the GR benchmark, both conditioned on values above $m_c(t)$ as in \ref{eq:conditioned_likelihood}.
+This figure clearly demonstrates that even after accounting for temporal incompleteness, the information gain achieved by our model remains an increasing function throughout the test set.
+The mean score of the test set for the conditioned results are shown in [[]] which displays an outpreformance for our model with this metric as well.
+This finding underscores the ability of our model to capture true physical features beyond mere measurement artifacts.
+
+[TODO: should I keep of remove the metrics for the causality defying metrics (present events) that beat the model in the ROC? this affects how the rest of this chapter will be written]
 

### 2024-03-15 article.tex
Modify into, update figures slightly

@@
 
-\title[Running Title]{Does an earthquake “know” how big it’s going to be? A neural-net aided study} % 2nd option: Using past seismicity to predict the magnitude of future earthquakes
+\title[Running Title]{Does an earthquake “know” how big it will be? A neural-net aided study} % 2nd option: Using past seismicity to predict the magnitude of future earthquakes
 
@@
 This includes various characteristics of spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}, which can be incorporated into stochastic prediction schemes \cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}.
-These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence.
+These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence [cite].
 Furthermore, these spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability \cite{schoenberg_testing_2004}.
@@
 by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events prediction often relies on drawing values from the Gutenberg-Richter (GR) distribution or its variants \cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}.
-Importantly, this distribution does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events.
+Importantly, this distribution either depends weakly or does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events.
 
@@
 
-While assuming that an earthquake's magnitude is independent of the seismic history is popular sue to its simplicity and its ease of use in earthquake modeling, it is not necessarily true.
+While assuming that an earthquake's magnitude is independent of the seismic history is popular due to its simplicity and its ease of use in earthquake modeling, it is not necessarily true.
 Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field.
@@
 \newline
+% perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's
 In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history.
@@
 This contribution has two important consequences.
-From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}.
+From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable and do not depend on seismic history \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}.
 Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction.
@@
 
-We construct a neural network (NN) model that encodes the seismic history of time series from a hypocentral catalog and produces an estimate of the possible magnitudes at a queried time and place.
-The constructed NN uses a long-short term memory (LSTM) architecture to model the seismic history up to a given time.
+In order to tackle the question of magnitude predictability, we construct the MAGnitude prediction Neural NETwork model, MAGNNET, a neural network (NN) for producing a probability density function of the possible magnitudes at a queried space time coordinates, conditioned on the seismic history.
+
+MAGNNET encodes the seismic history of time series from a hypocentral catalog and produces an estimate of the possible magnitudes at a queried time and place.
+
+MAGNNET's result will be quantitatively compared to other known benchmarks to estimate the amount of information gained by encoding the seismic history.
+MAGNNET is composed of a long-short term memory (LSTM) architecture to model the seismic history up to a given time.
 This latent representation is then passed together with  a space-time coordinate query of a future earthquake into a fully connected neural network (FCNN) that produces a parametrized probability density function (PDF) of the possible magnitudes at the quarried space- time coordinates.
 This procedure is depicted in Fig.
-\ref{fig:intro_fig}a depicts the task as described here.
-A detailed description of the model's architecture is given in the methods section.
-We optimize the model parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
+\ref{fig:intro_fig}a.
+A detailed description of the model's architecture is given in the methods section \neri{TODO}.
+We optimize MAGNNET's parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
 \begin{equation}
@@
 
-We train the model on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: Hauksson et al.
-(cite) for Southern California, GeoNet (cite) for New Zealand, and the JMA catalog (cite) for [Region 1, Region 2...].
+We train MAGNNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog [cite] for Southern California, GeoNet [cite] for New Zealand, and the JMA catalog [cite] for Japan.
 While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.?
@@
 
-Model performance is evaluated over a time span that was not included in training.
+The model's, MAGNNET's, performance is evaluated over a time span that was not included in training.
 The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence.
@@
     \caption{
-        Caption.
+        \textbf{outputs of MAGnitude prediction Neural NETwork, MAGNNET, for the Hauksson catalog of Southern California.
+a}, Likelihoods of MAGNNET and some common benchmark magnitude predictors.
+Solid lines represent the median value along the magnitude axis, shading shows the 10th and 90th percentiles of the distribution along the vertical axis.
+\textbf{b}, Train set's Gutenberg-Richter dostribution (dashen black line) with PDFs produced by MAGNNET for 100 randomly sampled events from the test set.
+Each PDF is color coded by the label of the event, its true magnitude.
+Colorbar to the right indicates the color code.
+\textbf{c}, Total magnitude frequency distribution produced by MAGNNET and common benchmarks for the entire test set.
+Histograms of the train (and test) set are presented in orange (blue).
+    }
+    \label{fig:architecture}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/combined_batplots.pdf}
+    \caption{
+        \textbf{Metric scores for out model MAGNNET, and common benchmark models.
+a}, \textbf{b}, \textbf{c}, Minus mean information content of our MAGNNET model (red) and other common benchmark magnitude predictors (see labels on bars in the figure).
+\textbf{d, e, f}, Reciever Operating Characteristic (ROC) curve for a binary classifier determinng the existence of a large ($m>=4$) event.
+Area under the curve (AUC) for each curve is written in each frame, colorcoded identically to the bar plots.
+\textbf{g, h, i}, Precision at recall curves for the binary classifier described in d, e, f.
     }
@@
     \includegraphics[width=1\textwidth]{figures/info_over_time_hauksson.pdf}
-    \caption{
-        Caption.
-    }
-    \label{fig:architecture}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \fbox{
-        \rule{1\textwidth}{0pt}\rule{0pt}{8ex}
-        FIGURE OF SPATIAL DISTRIBUTION OF GAIN.
-    }
-    \caption{
-        Caption.
-    }
-    \label{fig:architecture}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.4\textwidth]{figures/ROC_PR_hauksson.pdf}
     \caption{

### 2024-03-24 article.tex
More progress on results section

@@
 
-In order to tackle the question of magnitude predictability, we construct the MAGnitude prediction Neural NETwork model, MAGNNET, a neural network (NN) for producing a probability density function of the possible magnitudes at a queried space time coordinates, conditioned on the seismic history.
-
+In order to tackle the question of magnitude predictability, we construct the MAGnitude prediction Neural NETwork model, MAGNNET, a neural network (NN) for producing a probability density function (PDF) of the possible magnitudes at a queried space time coordinates, conditioned on the seismic history.
 MAGNNET encodes the seismic history of time series from a hypocentral catalog and produces an estimate of the possible magnitudes at a queried time and place.
@@
 MAGNNET is composed of a long-short term memory (LSTM) architecture to model the seismic history up to a given time.
-This latent representation is then passed together with  a space-time coordinate query of a future earthquake into a fully connected neural network (FCNN) that produces a parametrized probability density function (PDF) of the possible magnitudes at the quarried space- time coordinates.
-This procedure is depicted in Fig.
+This latent representation is then passed together with  a space-time coordinate query of a future earthquake into a fully connected neural network (FCNN) that produces a parametrized PDF of the possible magnitudes at the quarried space- time coordinates.
+The task described here, from encoding the seismic history to producing a PDF for a specific time-space querry is depicted in Fig.
 \ref{fig:intro_fig}a.
@@
 A separate model is trained for each region, with identical loss function and parameterization of the PDF.
-
-The model's, MAGNNET's, performance is evaluated over a time span that was not included in training.
-The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence.
-\neri{I think this is a good place to add that we conclude that there is some preparatory phase.}
-%This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation.
-Examining these results across multiple catalogs and regions highlights the robustness of our methodology and the generalizability of our conclusion, paving the way for further exploration of earthquake predictability.
+Sample results of known events worldwide are presented in \ref{fig:intro_fig}b.
+The PDF of each event is shown superimposed with the train set's Gutenberg-Richter distribution, i.e.
+the common vanilla benchmark.
 
@@
 
+The model's, MAGNNET's, performance is evaluated over a time span that was not included in training.
+The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence.
+\neri{I think this is a good place to add that we conclude that there is some preparatory phase.}
+%This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation.
+Examining these results across multiple catalogs and regions highlights the robustness of our methodology and the generalizability of our conclusion, paving the way for further exploration of earthquake predictability.
+
 \section{Results} \label{sec:results}
@@
 Consequently, the raw output for our analysis consists of the model-generated PDFs of possible magnitudes for each event's time and location.
-Figure \ref{fig:architecture}b showcases PDFs produced by our model, $p_{\textbf{x}_i, t_i}(m)$, for 100 randomly sampled event times and locations from the test set.
+Figure \ref{fig:model_output}a \neri{TODO: make sure numbering in this figure is correct} showcases PDFs produced by our model, $p_{\textbf{x}_i, t_i}(m)$, for 100 randomly sampled event times and locations from the test set of Southern California.
 Included as a baseline is the stationary GR predictor trained on the training set, representing its potential operational use (e.g., Tsuruoka et al., 2012).
 A clear trend emerges: PDFs predicted for higher-magnitude events (warm colors) exhibit a greater skew towards higher magnitudes compared to those for lower-magnitude events (cooler colors).
-Figure \ref{fig:output_pdfs}a further highlights this trend by displaying a selection of model-generated PDFs alongside the corresponding likelihood of retrieving the true magnitude at the time and location each PDF was estimated for, denoted by a [color] circle.
+Figure \ref{fig:model_output}b Presents the distribution of lieklihood scores of the test set.
+That is, $p_{\textbf{x}_i, t_i}^{(model)}(magntude label)$, the probability density for retrieving the label magnitude of an event from a given model.
+The solid line indicates the median, percentiles 10 and 90 are indicated by the shaded region.
+Here it can be seen that \neri{rerun this figure before writing conclusions}.
+To qualitativlry compare our model to other benchmark model in more common terms, it is useful to examine the marginal distribution for the magnitudes of the test set.
+\ref{fig:model_output}c presents the marginal PDF, $p(m)$, of MAGNNET compared to other common benchmarks for the Southern California dataset.
+This depicts the general trend that the model suggests and makes the comparison to common benchmarks, such as the GR, closer to the method they are typically used in, as a stationary distribution.
+\ref{fig:model_output}c also shows the histograms of the train and test set superimposed with the marginal PDFs of the models examined.
+It can clearly be seen that GR follows the train set, as expected, while MAGNNET follows the test set.
+Furthermore, the dynamic GR benchmarks such as last $n$ events of the $d$ days, although showing a closer functional for to the test than the train, lack features of the histogram that MAGNNET does not.
 
-Figure \ref{fig:output_pdfs}b focuses solely on the likelihood scores for retrieving the true magnitude, omitting the underlying PDFs used for their estimation.
-Additionally, it compares these scores against those generated by various benchmark models.
-Notably, our model demonstrates an advantage for most events exceeding magnitude $\sim 3$, while slightly under performing for lower-magnitude events.
-However, the overall mean log-likelihood (LL) score defined in \ref{eq:loss_function} (evaluated using the model-specific $p_{\textbf{x}_i, t_i}(\cdot)$ for each event) reveals a consistent advantage for our model across all datasets examined as summarized in \ref{tab:mean_ll}, where our model consistently obtains the lowest minus mean LL score in each column.
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/raw_results_hauksson.pdf}
+    \caption{
+        \textbf{outputs of MAGnitude prediction Neural NETwork, MAGNNET, for the Hauksson catalog of Southern California.
+a}, Likelihoods of MAGNNET and some common benchmark magnitude predictors.
+Solid lines represent the median value along the magnitude axis, shading shows the 10th and 90th percentiles of the distribution along the vertical axis.
+\textbf{b}, Train set's Gutenberg-Richter dostribution (dashen black line) with PDFs produced by MAGNNET for 100 randomly sampled events from the test set.
+Each PDF is color coded by the label of the event, its true magnitude.
+Colorbar to the right indicates the color code.
+\textbf{c}, Total magnitude frequency distribution produced by MAGNNET and common benchmarks for the entire test set.
+Histograms of the train (and test) set are presented in orange (blue).
+    }
+    \label{fig:model_output}
+\end{figure}
+
+Quantification of a model's preformance in our scheme will be done by averging on all log-likelihood scores of the test set, as defined in \ref{eq:loss_function}.
+\ref{fig:metrics}a-c shows the minus mean log likelihood score, $\mathcal{L}$ from \ref{eq:loss_function}, for MAGNNET and chosen common benchmarks, for the three regions tested.
+It be clearly seen that MAGNNET shows a clear an significant average information gain over every benchmark tested.
+\neri{Should we mention other benchmarks and refer to the SM?}.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/combined_batplots.pdf}
+    \caption{
+        \textbf{Metric scores for out model MAGNNET, and common benchmark models.
+a}, \textbf{b}, \textbf{c}, Minus mean information content of our MAGNNET model (red) and other common benchmark magnitude predictors (see labels on bars in the figure).
+\textbf{d, e, f}, Reciever Operating Characteristic (ROC) curve for a binary classifier determinng the existence of a large ($m>=4$) event.
+Area under the curve (AUC) for each curve is written in each frame, colorcoded identically to the bar plots.
+\textbf{g, h, i}, Precision at recall curves for the binary classifier described in d, e, f.
+        }
+        \label{fig:metrics}
+\end{figure}
+
+As the out model's output is a probability density function (PDF), we may easily convert it to a binary classifer.
+By doing this we are posing a modified question: will there be a large earthquake in the query space-time coordiantes or not?
+This is easily done by setting a threshold which defines a "large" earthquake, $m_c$ ($=4$ in the examples presented here), then setting examples above that threshold as positive examples and below as negative.
+The probability of a large event occuring in the query coordinates is given by integrating the output PDF above the threshild, $p=\int_{m_c}^{\infty}p(m')dm'$.
+Given such binary classifier, it is possible now to examine the common binary classifier metrics of area under reciever operator (ROC) curve [cite] and the prescision at recall curve [cite].
+\ref{fig:metrics}d-f show the ROC curves for the binary modification for all three regions in question.
+The area under the curve (AUC) are written to the bottom right of each panel color coded by the same legend as the bar plots above.
+It can be seen that throughout all regions MAGNNET preforms significantly better than all other benchmarks, indicated by the largest AUC value.
+The precision at recall curve displayed in \ref{fig:metrics}g-i show a similar result as MAGNNET's curve displays the largest AUC among all benchmarks tested throughout all regions.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/info_over_time_hauksson.pdf}
+    \caption{
+        Caption.
+    }
+    \label{fig:info_gain_over_time}
+\end{figure}
 
@@
 Analyzing this quantity per event across the entire test set reveals patterns in where our model achieves its advantage.
-\ref{fig:likelihood_over_time}a shows the Hauksson data test set event magnitudes at their occurrence times, colored according to the information gain.
-This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as a gray dashed line in the figure.
+\ref{fig:info_gain_over_time}a shows the test set for Southern California event magnitudes by their running index, colored according to the information gain of our model compared to the GR benchmark.
+This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as a black dashed line in the figure.
 The predominantly increasing trend in cumulative information gain suggests that our model is consistently advantageous as opposed to a concentrated advantage in specific periods.
-Notably, for the Hauksson data set, the most rapid information gain occurs during the aftershocks of the two Ridgecrest earthquakes.
+Notably, for the Southern California data set, the most rapid information gain occurs during the aftershocks of the two Ridgecrest earthquakes.
 Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4$ bits, respectively, though these events are typically considered completely unpredictable.
@@
 [TODO: should I keep of remove the metrics for the causality defying metrics (present events) that beat the model in the ROC? this affects how the rest of this chapter will be written]
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=1\textwidth]{figures/raw_results_hauksson.pdf}
-    \caption{
-        \textbf{outputs of MAGnitude prediction Neural NETwork, MAGNNET, for the Hauksson catalog of Southern California.
-a}, Likelihoods of MAGNNET and some common benchmark magnitude predictors.
-Solid lines represent the median value along the magnitude axis, shading shows the 10th and 90th percentiles of the distribution along the vertical axis.
-\textbf{b}, Train set's Gutenberg-Richter dostribution (dashen black line) with PDFs produced by MAGNNET for 100 randomly sampled events from the test set.
-Each PDF is color coded by the label of the event, its true magnitude.
-Colorbar to the right indicates the color code.
-\textbf{c}, Total magnitude frequency distribution produced by MAGNNET and common benchmarks for the entire test set.
-Histograms of the train (and test) set are presented in orange (blue).
-    }
-    \label{fig:architecture}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=1\textwidth]{figures/combined_batplots.pdf}
-    \caption{
-        \textbf{Metric scores for out model MAGNNET, and common benchmark models.
-a}, \textbf{b}, \textbf{c}, Minus mean information content of our MAGNNET model (red) and other common benchmark magnitude predictors (see labels on bars in the figure).
-\textbf{d, e, f}, Reciever Operating Characteristic (ROC) curve for a binary classifier determinng the existence of a large ($m>=4$) event.
-Area under the curve (AUC) for each curve is written in each frame, colorcoded identically to the bar plots.
-\textbf{g, h, i}, Precision at recall curves for the binary classifier described in d, e, f.
-    }
-    \label{fig:architecture}
-\end{figure}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=1\textwidth]{figures/info_over_time_hauksson.pdf}
-    \caption{
-        Caption.
-    }
-    \label{fig:architecture}
-\end{figure}
 

### 2024-03-24 article.tex
Progress on results

@@
 \newline
-% perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's
+\neri{perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's}
 In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history.
@@
 
-Quantification of a model's preformance in our scheme will be done by averging on all log-likelihood scores of the test set, as defined in \ref{eq:loss_function}.
-\ref{fig:metrics}a-c shows the minus mean log likelihood score, $\mathcal{L}$ from \ref{eq:loss_function}, for MAGNNET and chosen common benchmarks, for the three regions tested.
-It be clearly seen that MAGNNET shows a clear an significant average information gain over every benchmark tested.
+Quantification of a model's preformance in our scheme will be done by averging on all log-likelihood scores of the test set, as defined in Eq.
+\ref{eq:loss_function}.
+Fig.
+\ref{fig:metrics}a-c shows the minus mean log likelihood score, $\mathcal{L}$ from Eq.
+\ref{eq:loss_function}, for MAGNNET and chosen common benchmarks, for the three regions tested.
+It can be clearly seen that MAGNNET shows a clear an significant average information gain over every benchmark tested.
 \neri{Should we mention other benchmarks and refer to the SM?}.
@@
 As the out model's output is a probability density function (PDF), we may easily convert it to a binary classifer.
-By doing this we are posing a modified question: will there be a large earthquake in the query space-time coordiantes or not?
-This is easily done by setting a threshold which defines a "large" earthquake, $m_c$ ($=4$ in the examples presented here), then setting examples above that threshold as positive examples and below as negative.
+By doing this we are posing a modified question: will there be a \textit{large} earthquake in the query space-time coordiantes or not?
+This is done by setting a threshold which defines a "\textit{large}" earthquake, $m_c$ ($=4$ in the examples presented here), then setting examples above that threshold as positive examples and below as negative.
 The probability of a large event occuring in the query coordinates is given by integrating the output PDF above the threshild, $p=\int_{m_c}^{\infty}p(m')dm'$.
@@
 The precision at recall curve displayed in \ref{fig:metrics}g-i show a similar result as MAGNNET's curve displays the largest AUC among all benchmarks tested throughout all regions.
+\neri{TODO: calculate and present AUC P@R}
 
@@
 \begin{equation}
-    \Delta LL = \log{p_{\pmb{x_i}, t_i}^{(model)}(m_i)} - \log{p^{(GR)}(m_i)}
+    \Delta LL = \log{p_{\pmb{x_i}, t_i}^{(MAGNNET)}(m_i)} - \log{p^{(GR)}(m_i)}
     \label{eq:information_gain}
 \end{equation}
-This equation measures the number of bits of information gained by using our model compared to the GR benchmark.
+This equation measures the amount of information gained by using our MAGNNET model compared to the GR benchmark.
 Analyzing this quantity per event across the entire test set reveals patterns in where our model achieves its advantage.
-\ref{fig:info_gain_over_time}a shows the test set for Southern California event magnitudes by their running index, colored according to the information gain of our model compared to the GR benchmark.
-This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as a black dashed line in the figure.
+Fig.
+\ref{fig:info_gain_over_time}a shows the test set for Southern California event magnitudes by their running index, colored according to the information gain of our model over the GR benchmark.
+This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as the black dashed line in the figure.
 The predominantly increasing trend in cumulative information gain suggests that our model is consistently advantageous as opposed to a concentrated advantage in specific periods.
 Notably, for the Southern California data set, the most rapid information gain occurs during the aftershocks of the two Ridgecrest earthquakes.
-Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4$ bits, respectively, though these events are typically considered completely unpredictable.
-This observation leads to a significant corollary: at least some portion of the information gain likely originates from a physical precursory signal.
+Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4, \ \Delta LL$, respectively, albeit these events are typically considered completely unpredictable.
+This observation leads to a significant corollary: at least some portion of the information gain likely originates from a statistical signal embedded in the event sequence prior to event's occurence.
 This point will be further elaborated upon in the discussion section.
 \newline
-The observed increase in information gain persists throughout the entire test set for all regions examined in this study.
+The observed increase in information gain persists throughout the entire test set for all regions examined in this study as the black dashed curves in Fig.
+\ref{fig:info_gain_over_time}b-d present.
 This includes regions previously analyzed by (Ogata et al.
 SRL 2018), where no information gain was detected using the method employed at the time.
-Further analysis of \ref{fig:likelihood_over_time} applied to other regions is presented in the supplementary material.
+Further analysis of \ref{fig:info_gain_over_time} applied to other regions is presented in the supplementary material.
+\neri{TODO: ad these colorful plots for NZ and JAPAN to the SM}
+The smoothed instantaneous information gain per event is presented in Fig.
+\ref{fig:info_gain_over_time}e-g, and can be seen as commonly well above the threshold of zero information gain, marked by the red dashed line.
 \newline
-While the information gain observed in pre- and mainshocks presented above is encouraging, a potential concern might arise regarding the aftershocks: could their information gain be an artifact of fitting to the temporal incompleteness of the catalog (cite somethong?)?
+While the information gain observed in preshock and mainshocks presented above is encouraging, a potential concern might arise regarding the aftershocks: could their information gain be an artifact of fitting to the temporal incompleteness of the catalog (cite somethong?)?
 To address this, we factor out this potential bias by recalculating the likelihood score of each event, $p(m_i)$ , conditioned on the magnitude exceeding a threshold, $\Tilde{m}$:
@@
 For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method (Weimer and Weiss, 2000) within a window of 150 past and 150 future events.
-The resulting temporal incompleteness over time is displayed in \ref{fig:likelihood_over_time}b.
-Crucially, \ref{fig:likelihood_over_time}b also presents the magnitudes of the Ridgecrest sequence at their origin times, colored by the information gain of our model over the GR benchmark, both conditioned on values above $m_c(t)$ as in \ref{eq:conditioned_likelihood}.
-This figure clearly demonstrates that even after accounting for temporal incompleteness, the information gain achieved by our model remains an increasing function throughout the test set.
-The mean score of the test set for the conditioned results are shown in [[]] which displays an outpreformance for our model with this metric as well.
-This finding underscores the ability of our model to capture true physical features beyond mere measurement artifacts.
-
-[TODO: should I keep of remove the metrics for the causality defying metrics (present events) that beat the model in the ROC? this affects how the rest of this chapter will be written]
+The resulting temporal incompleteness curves are presented in the SM \neri{TODO: add the $m_c(t)$ figure to SM}.
+After applying the renormalization of Eq.
+\ref{eq:conditioned_likelihood} the cumulative information gain is recalculated.
+The result is presented as the grey dashed curved in Fig.
+\ref{fig:info_gain_over_time}a, as well as in Fig.
+\ref{fig:info_gain_over_time}b-d for all three tested regions.
+\newline
+A second concern regarding temporal variations of the completeness magnitude may be raised to oppose to the analysis and results presented in this work.
+The following paragraph will provide an explanation of our method for varifying that we are indeed measuring a true statistical trend that is not created by the local completeness magnitude artifact.
+We start by deviding each region into bins of size [dx, dy] \neri{TODO fill in the details}.
+The average log likelihood score $\mathcal{L}$ is calculated per bin for both MAGNNET and GR, resulting in a map of spatial average log likelihood, per model.
+The subtraction of the two maps is presented in Fig.
+\ref{fig:info_gain_over_time}h-j, showing the spatial distribution of information gain for each region.
+\neri{TODO complete paragraph about how I select only bins of mc(x,y)<=mc to recalculate the average score and show that I still gain information.}
+for which the local completness magnitude $m_c(x, y)$ is calculated.
 

### 2024-04-08 article.tex
Yohai's changes

@@
 Earthquake occurrence is notoriously difficult to predict.
-Salient features of the statistical properties of the timing and location of earthquakes, i.e.
-spatiotemporal clustering, can be relatively well captured by point-process models.
-However, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts, both due to the lack of information about the fault condition and to the inherent complexity of rupture dynamics.
-Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events, besides the time-independent Gutenberg Richter (GR) distribution, which describes the marginal distribution over wide regions and long times.
+Salient features of their spatiotemporal statistics can be relatively well captured by point-process models.
+However, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts, both due to the lack of information about fault conditions and to the inherent complexity of rupture dynamics.
+Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter distribution, which describes the marginal distribution over large regions and long times.
 This approach implicitly assumes that earthquake magnitudes are independent of seismic history and are identically distributed.
@@
 \section{Introduction} \label{sec:introduction}
+Earthquakes are notoriously unpredictable, and forecasting seismicity is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes and the scarcity of near-field data~\cite{bernard_earthquake_1999, geller_earthquakes_1997}.
+Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes, including various marginal distributions \yohai{[cite GR and variants]}, scaling relations \yohai{[cite]} and characteristics of both spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}.
+Clearly, these insights can be used to quantitatively inform us about future seismicity based on recent history.
+For example, Omori's law tells us that after big earthquakes we should expect an increase in the local seismicity rate~\cite{omori_after-shocks_1894}.
+Such laws have been incorporated into a variety of forecasting models, which are operationally used today\cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}.
 
-Earthquake forecasting is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes \cite{bernard_earthquake_1999, geller_earthquakes_1997}.
-Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes.
-This includes various characteristics of spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}, which can be incorporated into stochastic prediction schemes \cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}.
-These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence [cite].
-Furthermore, these spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability \cite{schoenberg_testing_2004}.
-That is, while seismic history is used to inform the location and rate of future earthquakes, i.e.
-by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events prediction often relies on drawing values from the Gutenberg-Richter (GR) distribution or its variants \cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}.
-Importantly, this distribution either depends weakly or does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events.
+However, the above mentioned statistical relations only describe the rate and locations of earthquakes, and very little is known about the dependence of earthquake magnitude on seismic history.
+It is in fact deeply debated whether it is possible to independent of seismic history the magnitude of an earthquake before it starts~\cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}, or even during rupture~\yohai{[cite]}.
+Most modeling approaches assume that event magnitudes are identically and independently drawn from a stationary (or almost stationary) distribution - the Gutenberg Richter distribution (GR), which describes the marginal magnitude distribution over large regions and long times.
+While some variations in magnitude statistics have been described \yohai{[Gulia Weimar, beta as stress sensor papers, Didier's paper]}, they all model history dependence by small and slow changes in the parameters of the GR distribution, and do not inform us about the magnitude of \textit{a specific} future event.
+Even state-of-the-art forecasting models typically assume no knowledge about the magnitude of future events besides the marginal GR distribution, an assumption known as separability \cite{schoenberg_testing_2004}.
+Indeed, Current research on magnitude prediction has failed to show a universal and reproducible advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}.
 
-% This might be something to remove to shorten
+This modeling approach is not a naive or uninformed choice.
+Due to the spatial complexity of the elastic fields, faulting patterns and lithology, the nonlinearities of the rupture process and the complicated interaction between them all, it is not far-fetched to assume that determining the magnitude of an event  requires full microscopic knowledge of the system's state.
+This philosophy also inspired physical models which describe earthquake dynamics with a statistical approach \yohai{(cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff)}.
+These models reproduce many of the phenomenological  statisitcal relations by positing that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition.
+Under this paradigm, determining the magnitude of an event is indeed impossible using only far-field measurements.
+
+\neri{perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's} \yohai{yes}
+
+The main goal of this paper is to ask the question directly: can we extract any information about a future earthquake's magnitude from regional seismic history?
+To this end, we construct a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence.
+Our model is provided with seismic catalogs which include the location, time and magnitude of past events.
+Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided.
+Thus, we separate out the task of modeling nucleation statistics and isolate the question of magnitude predictability.
+If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of a specific earthquake is extractable from cataloged properties alone.
+
+This contribution has two important consequences.
+From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable and do not depend on seismic history \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}.
+Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction.
+This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.
+
 % This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice.
@@
 
-While assuming that an earthquake's magnitude is independent of the seismic history is popular due to its simplicity and its ease of use in earthquake modeling, it is not necessarily true.
-Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field.
-These in turn depend on slip history in a deterministic manner, albeit intangibly complicated.
-Therefore, the assumption of magnitude separability, i.e.
-its independent from seismic history, is not a priory justified, and its validity should be tested empirically.
-Though information about an earthquake magnitude may exists before its initiation, implying dependence on seismic history and non-separability, it has yet to be shown accessible using our current experimental or analytic tools.
-Current research on magnitude prediction has failed to show a universal and replicable advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}, by that leaving the question of accessible information prior to an earthquake's nucleation withstanding.
-\newline
-\neri{perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's}
-In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history.
-This is done by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and long term seismic history prior to its occurrence.
-Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model.
-Thus, we separate out the task of predicting the nucleation of events and isolate the question of magnitude predictability.
-If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of an earthquake is extractable from cataloged properties alone.
+\section{Methodology}
+We construct the MAGnitude Neural EsTimation model, MAGNNET, a generative neural network (NN) that receives as input a hypocentral catalog of past regional seismicity, and the time and location of a future event (a ``query'').
+The network produces a probability density function (PDF) estimating its magnitude.
 
-This contribution has two important consequences.
-From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable and do not depend on seismic history \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}.
-Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction.
-This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.
+%MAGNNET's result will be quantitatively compared to other known benchmarks to estimate the amount of information gained by encoding the seismic history.
 
-\neri{Is a literature review nescessary here or is it ingested enough in the text itself? I'm pretty satisfied with how the paragraphs up to now are written.}\newline
-\neri{The entire paragraph going on about SOC, perhaps to long and should be inserted as a sentence somewhere in the literature review? can be seem as comment in source code.}
-% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice.
-On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff).
-These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition.
-Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics.
-In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).
-
-In order to tackle the question of magnitude predictability, we construct the MAGnitude prediction Neural NETwork model, MAGNNET, a neural network (NN) for producing a probability density function (PDF) of the possible magnitudes at a queried space time coordinates, conditioned on the seismic history.
-MAGNNET encodes the seismic history of time series from a hypocentral catalog and produces an estimate of the possible magnitudes at a queried time and place.
-
-MAGNNET's result will be quantitatively compared to other known benchmarks to estimate the amount of information gained by encoding the seismic history.
-MAGNNET is composed of a long-short term memory (LSTM) architecture to model the seismic history up to a given time.
-This latent representation is then passed together with  a space-time coordinate query of a future earthquake into a fully connected neural network (FCNN) that produces a parametrized PDF of the possible magnitudes at the quarried space- time coordinates.
-The task described here, from encoding the seismic history to producing a PDF for a specific time-space querry is depicted in Fig.
+MAGNNET is composed of two main components. First, the catalog up to time $t$ is passed through a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history.
+Second, this representation combined with a space-time query specifying the coordinates of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, see Fig.
 \ref{fig:intro_fig}a.
 A detailed description of the model's architecture is given in the methods section \neri{TODO}.
-We optimize MAGNNET's parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
+We optimize MAGNNET's parameters to maximize the log likelihood (LL) of the observed magnitudes,
 \begin{equation}
-    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i} \left( m_i \right) \right) } \rangle\ ,
+    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i(h_t)} \left( m_i \right) \right) } \rangle\ ,
     \label{eq:loss_function}
 \end{equation}
-where $p_{\textbf{x}_i, t_i}$ is the probability function returned by the model for the space-time coordinates $\textbf{x}_i, t_i$ (i.e.
-the ``query coordinates''), $m_i$ the corresponding to the actual magnitude of the $i^{th}$ earthquake and $\langle \cdot\rangle$ stands for the empirical average over the data set.
-We use the a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} parameterized family of probability distributions stretched to a relevant domain.
-The final distribution used is defined by 5 parameters.
+where $\bf{x}_i$, $t_i$, $m_i$ re the location, time and magnitude of the $i$-th event, $p_{\textbf{x}_i, t_i}(h_t)$ is the probability function returned by the model,and $h_t$ is the seismic history up to time $t$.
+$\langle \cdot\rangle$ stands for the empirical average over all earthquakes in the data set.
+Importantly, during training we only supply the model with queries about spacetime coordinates where earthquakes indeed occured.
+\yohai{we should state that we only consider events about $m_c$}
+
+Technically, the PDF is a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} stretched to a relevant domain, which is a parameteric distribution with 5 parameters, that the network outputs.
 This family smoothly can interpolate between a decaying distribution (resembling the GR distribution), and localized distributions whose mass is concentrated around a specific value.
@@
 
-We train MAGNNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog [cite] for Southern California, GeoNet [cite] for New Zealand, and the JMA catalog [cite] for Japan.
+We deployed MAGNNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog [cite] for Southern California, GeoNet [cite] for New Zealand, and the JMA catalog [cite] for Japan.
 While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.?
@@
 A separate model is trained for each region, with identical loss function and parameterization of the PDF.
-Sample results of known events worldwide are presented in \ref{fig:intro_fig}b.
-The PDF of each event is shown superimposed with the train set's Gutenberg-Richter distribution, i.e.
-the common vanilla benchmark.
+The resulting PDF is presented in \ref{fig:intro_fig}b for a few examples of major events, superimposed on the stationary GR distribution (fitted on the train set), which is the naive benchmark.
 
@@
 The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence.
-\neri{I think this is a good place to add that we conclude that there is some preparatory phase.}
+\neri{I think this is a good place to add that we conclude that there is some preparatory phase.} \yohai{This is not a conclusion that you can draw from our result.
+We can only say that there is some information about the magnitude of an event in the seismic history.} \neri{I think we can say that the information gain is indicative of a preparatory phase.} \yohai{I think we should be careful with this.} \neri{I think we should discuss this.} \yohai{This is quite amazing, I only wrote the first sentence (``This is not a conclusion that you can draw from our result'') and copilot autocompleted the rest of the dialog, correctly using the \textbackslash neri\{...\} and \textbackslash yohai\{\} macros }
 %This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation.

### 2024-04-09 article.tex
Some of my comments on the article

@@
 Earthquake occurrence is notoriously difficult to predict.
-Salient features of their spatiotemporal statistics can be relatively well captured by point-process models.
+Salient features of their spatiotemporal statistics can be relatively well captured by point-process models \neri{this is mentioning ETAS.
+But it is also not completely true.
+We should remove the mention of PP IMO}.
 However, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts, both due to the lack of information about fault conditions and to the inherent complexity of rupture dynamics.
@@
 In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history.
-We present a neural network-based model for probabilistic forecasting of future magnitudes based on cataloged properties: hypocentral locations, occurrence times and magnitudes of past earthquakes.
+We present a neural network-based model for probabilistic forecasting of future magnitudes based on cataloged properties: hypocenter locations, occurrence times and magnitudes of past earthquakes.
 Our history-dependent model outperforms stationary and quasi-stationary state of the art GR-based benchmarks, in real catalogs in Southern California, Japan and New-Zealand.
@@
 For example, Omori's law tells us that after big earthquakes we should expect an increase in the local seismicity rate~\cite{omori_after-shocks_1894}.
-Such laws have been incorporated into a variety of forecasting models, which are operationally used today\cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}.
+Such laws have been incorporated into a variety of forecasting models, which are operationally used today\cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994} \neri{TODO add citation OEF related, Marzocci, Christoforson}.
 
@@
 Due to the spatial complexity of the elastic fields, faulting patterns and lithology, the nonlinearities of the rupture process and the complicated interaction between them all, it is not far-fetched to assume that determining the magnitude of an event  requires full microscopic knowledge of the system's state.
+\neri{Should this statement be suported?
+Kinda hand-wavy}
 This philosophy also inspired physical models which describe earthquake dynamics with a statistical approach \yohai{(cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff)}.
@@
 The main goal of this paper is to ask the question directly: can we extract any information about a future earthquake's magnitude from regional seismic history?
-To this end, we construct a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence.
+\neri{Stress more: this will imply we have information relevant to prediction tasks.
+Zhuang stated: "you have touched one of the most important problems on earthquake predictability"} To this end, we construct a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence.
 Our model is provided with seismic catalogs which include the location, time and magnitude of past events.
-Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided.
+\neri{Hauksson also has angles, how should this be phrased?} Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided.
 Thus, we separate out the task of modeling nucleation statistics and isolate the question of magnitude predictability.
@@
 From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable and do not depend on seismic history \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}.
-Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction.
+\neri{I think we should consider going softer on the SOC claim - most geologists don't care/ know of it} Second, it suggests that the separability assumption, \neri{where did the explanation about seperability go?
+This paragraph is not clear.} which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction.
 This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.
@@
 \section{Methodology}
-We construct the MAGnitude Neural EsTimation model, MAGNNET, a generative neural network (NN) that receives as input a hypocentral catalog of past regional seismicity, and the time and location of a future event (a ``query'').
+We construct the MAGnitude Neural EsTimation model, MAGNET, a generative neural network (NN) that receives as input a hypocentral catalog of past regional seismicity, and the time and location of a future event (a ``query'').
 The network produces a probability density function (PDF) estimating its magnitude.
 
-%MAGNNET's result will be quantitatively compared to other known benchmarks to estimate the amount of information gained by encoding the seismic history.
+%MAGNET's result will be quantitatively compared to other known benchmarks to estimate the amount of information gained by encoding the seismic history.
 
-MAGNNET is composed of two main components. First, the catalog up to time $t$ is passed through a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history.
+MAGNET is composed of two main components. First, the catalog up to time $t$ is passed through a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history.
 Second, this representation combined with a space-time query specifying the coordinates of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, see Fig.
@@
 A detailed description of the model's architecture is given in the methods section \neri{TODO}.
-We optimize MAGNNET's parameters to maximize the log likelihood (LL) of the observed magnitudes,
+We optimize MAGNET's parameters to maximize the log likelihood (LL) of the observed magnitudes,
 \begin{equation}
@@
 \end{equation}
+\neri{I find the $h_t$ is confusing and not nescessary}
 where $\bf{x}_i$, $t_i$, $m_i$ re the location, time and magnitude of the $i$-th event, $p_{\textbf{x}_i, t_i}(h_t)$ is the probability function returned by the model,and $h_t$ is the seismic history up to time $t$.
@@
 
-Technically, the PDF is a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} stretched to a relevant domain, which is a parameteric distribution with 5 parameters, that the network outputs.
+Technically \neri{find a better word}, the PDF is a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} stretched to a relevant domain, which is a parameteric distribution with 5 parameters, that the network outputs.
 This family smoothly can interpolate between a decaying distribution (resembling the GR distribution), and localized distributions whose mass is concentrated around a specific value.
@@
 
-We deployed MAGNNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog [cite] for Southern California, GeoNet [cite] for New Zealand, and the JMA catalog [cite] for Japan.
+We deployed MAGNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog [cite] for Southern California, GeoNet [cite] for New Zealand, and the JMA catalog [cite] for Japan.
 While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.?
@@
 
-The model's, MAGNNET's, performance is evaluated over a time span that was not included in training.
+The model's, MAGNET's, performance is evaluated over a time span that was not included in training.
 The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence.
@@
 To qualitativlry compare our model to other benchmark model in more common terms, it is useful to examine the marginal distribution for the magnitudes of the test set.
-\ref{fig:model_output}c presents the marginal PDF, $p(m)$, of MAGNNET compared to other common benchmarks for the Southern California dataset.
+\ref{fig:model_output}c presents the marginal PDF, $p(m)$, of MAGNET compared to other common benchmarks for the Southern California dataset.
 This depicts the general trend that the model suggests and makes the comparison to common benchmarks, such as the GR, closer to the method they are typically used in, as a stationary distribution.
 \ref{fig:model_output}c also shows the histograms of the train and test set superimposed with the marginal PDFs of the models examined.
-It can clearly be seen that GR follows the train set, as expected, while MAGNNET follows the test set.
-Furthermore, the dynamic GR benchmarks such as last $n$ events of the $d$ days, although showing a closer functional for to the test than the train, lack features of the histogram that MAGNNET does not.
+It can clearly be seen that GR follows the train set, as expected, while MAGNET follows the test set.
+Furthermore, the dynamic GR benchmarks such as last $n$ events of the $d$ days, although showing a closer functional for to the test than the train, lack features of the histogram that MAGNET does not.
 
@@
     \caption{
-        \textbf{outputs of MAGnitude prediction Neural NETwork, MAGNNET, for the Hauksson catalog of Southern California.
-a}, Likelihoods of MAGNNET and some common benchmark magnitude predictors.
+        \textbf{outputs of MAGnitude prediction Neural NETwork, MAGNET, for the Hauksson catalog of Southern California.
+a}, Likelihoods of MAGNET and some common benchmark magnitude predictors.
 Solid lines represent the median value along the magnitude axis, shading shows the 10th and 90th percentiles of the distribution along the vertical axis.
-\textbf{b}, Train set's Gutenberg-Richter dostribution (dashen black line) with PDFs produced by MAGNNET for 100 randomly sampled events from the test set.
+\textbf{b}, Train set's Gutenberg-Richter dostribution (dashen black line) with PDFs produced by MAGNET for 100 randomly sampled events from the test set.
 Each PDF is color coded by the label of the event, its true magnitude.
 Colorbar to the right indicates the color code.
-\textbf{c}, Total magnitude frequency distribution produced by MAGNNET and common benchmarks for the entire test set.
+\textbf{c}, Total magnitude frequency distribution produced by MAGNET and common benchmarks for the entire test set.
 Histograms of the train (and test) set are presented in orange (blue).
@@
 \ref{fig:metrics}a-c shows the minus mean log likelihood score, $\mathcal{L}$ from Eq.
-\ref{eq:loss_function}, for MAGNNET and chosen common benchmarks, for the three regions tested.
-It can be clearly seen that MAGNNET shows a clear an significant average information gain over every benchmark tested.
+\ref{eq:loss_function}, for MAGNET and chosen common benchmarks, for the three regions tested.
+It can be clearly seen that MAGNET shows a clear an significant average information gain over every benchmark tested.
 \neri{Should we mention other benchmarks and refer to the SM?}.
@@
     \caption{
-        \textbf{Metric scores for out model MAGNNET, and common benchmark models.
-a}, \textbf{b}, \textbf{c}, Minus mean information content of our MAGNNET model (red) and other common benchmark magnitude predictors (see labels on bars in the figure).
+        \textbf{Metric scores for out model MAGNET, and common benchmark models.
+a}, \textbf{b}, \textbf{c}, Minus mean information content of our MAGNET model (red) and other common benchmark magnitude predictors (see labels on bars in the figure).
 \textbf{d, e, f}, Reciever Operating Characteristic (ROC) curve for a binary classifier determinng the existence of a large ($m>=4$) event.
@@
 The area under the curve (AUC) are written to the bottom right of each panel color coded by the same legend as the bar plots above.
-It can be seen that throughout all regions MAGNNET preforms significantly better than all other benchmarks, indicated by the largest AUC value.
-The precision at recall curve displayed in \ref{fig:metrics}g-i show a similar result as MAGNNET's curve displays the largest AUC among all benchmarks tested throughout all regions.
+It can be seen that throughout all regions MAGNET preforms significantly better than all other benchmarks, indicated by the largest AUC value.
+The precision at recall curve displayed in \ref{fig:metrics}g-i show a similar result as MAGNET's curve displays the largest AUC among all benchmarks tested throughout all regions.
 \neri{TODO: calculate and present AUC P@R}
@@
 \begin{equation}
-    \Delta LL = \log{p_{\pmb{x_i}, t_i}^{(MAGNNET)}(m_i)} - \log{p^{(GR)}(m_i)}
+    \Delta LL = \log{p_{\pmb{x_i}, t_i}^{(MAGNET)}(m_i)} - \log{p^{(GR)}(m_i)}
     \label{eq:information_gain}
 \end{equation}
-This equation measures the amount of information gained by using our MAGNNET model compared to the GR benchmark.
+This equation measures the amount of information gained by using our MAGNET model compared to the GR benchmark.
 Analyzing this quantity per event across the entire test set reveals patterns in where our model achieves its advantage.
@@
 We start by deviding each region into bins of size [dx, dy] \neri{TODO fill in the details}.
-The average log likelihood score $\mathcal{L}$ is calculated per bin for both MAGNNET and GR, resulting in a map of spatial average log likelihood, per model.
+The average log likelihood score $\mathcal{L}$ is calculated per bin for both MAGNET and GR, resulting in a map of spatial average log likelihood, per model.
 The subtraction of the two maps is presented in Fig.

### 2024-05-20 SI.tex
finished results. started methods end extended material.

--- 2024-03-24 article.tex
@@
 
-\title[Running Title]{Does an earthquake “know” how big it will be? A neural-net aided study} % 2nd option: Using past seismicity to predict the magnitude of future earthquakes
-
-\author[1,2]{\fnm{Neri} \sur{Berman}}\email{neriberman@gmail.com}
-\author[2]{\fnm{Oleg} \sur{Zlydenko}}
-\author[2]{\fnm{Oren} \sur{Gilon}}
-\author[1]{\fnm{Yohai} \sur{Bar-Sinai}}\email{ybarsinai@gmail.com}
-
-\affil[1]{\orgdiv{Department of Physics}, \orgname{Tel-Aviv University}, \orgaddress{\city{Tel-Aviv}, \country{Israel}}}
-\affil[2]{\orgdiv{Google Research}, \orgname{Google}, \orgaddress{\city{Tel-Aviv}, \country{Israel}}}
-
-\abstract{
-Earthquake occurrence is notoriously difficult to predict.
-Salient features of the statistical properties of the timing and location of earthquakes, i.e.
-spatiotemporal clustering, can be relatively well captured by point-process models.
-However, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts, both due to the lack of information about the fault condition and to the inherent complexity of rupture dynamics.
-Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events, besides the time-independent Gutenberg Richter (GR) distribution, which describes the marginal distribution over wide regions and long times.
-This approach implicitly assumes that earthquake magnitudes are independent of seismic history and are identically distributed.
-In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history.
-We present a neural network-based model for probabilistic forecasting of future magnitudes based on cataloged properties: hypocentral locations, occurrence times and magnitudes of past earthquakes.
-Our history-dependent model outperforms stationary and quasi-stationary state of the art GR-based benchmarks, in real catalogs in Southern California, Japan and New-Zealand.
-This demonstrates that earthquake catalogs contain information about the magnitude of future earthquakes, prior to their occurrence.
-We conclude by proposing methods to apply the model in characterization of the preparatory phase of earthquakes, and in operational hazard alert and earthquake forecasting systems.
-}
+\title[Running Title]{Supplamentary information for: Does an earthquake “know” how big it will be?
+A neural-net aided study} % 2nd option: Using past seismicity to predict the magnitude of future earthquakes
 
@@
 
-\section{Introduction} \label{sec:introduction}
-
-Earthquake forecasting is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes \cite{bernard_earthquake_1999, geller_earthquakes_1997}.
-Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes.
-This includes various characteristics of spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}, which can be incorporated into stochastic prediction schemes \cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}.
-These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence [cite].
-Furthermore, these spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability \cite{schoenberg_testing_2004}.
-That is, while seismic history is used to inform the location and rate of future earthquakes, i.e.
-by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events prediction often relies on drawing values from the Gutenberg-Richter (GR) distribution or its variants \cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}.
-Importantly, this distribution either depends weakly or does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events.
-
-% This might be something to remove to shorten
-% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice.
-On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff).
-These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition.
-Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics.
-In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).
-
-While assuming that an earthquake's magnitude is independent of the seismic history is popular due to its simplicity and its ease of use in earthquake modeling, it is not necessarily true.
-Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field.
-These in turn depend on slip history in a deterministic manner, albeit intangibly complicated.
-Therefore, the assumption of magnitude separability, i.e.
-its independent from seismic history, is not a priory justified, and its validity should be tested empirically.
-Though information about an earthquake magnitude may exists before its initiation, implying dependence on seismic history and non-separability, it has yet to be shown accessible using our current experimental or analytic tools.
-Current research on magnitude prediction has failed to show a universal and replicable advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}, by that leaving the question of accessible information prior to an earthquake's nucleation withstanding.
-\newline
-\neri{perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's}
-In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history.
-This is done by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and long term seismic history prior to its occurrence.
-Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model.
-Thus, we separate out the task of predicting the nucleation of events and isolate the question of magnitude predictability.
-If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of an earthquake is extractable from cataloged properties alone.
-
-This contribution has two important consequences.
-From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable and do not depend on seismic history \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}.
-Second, it suggests that the separability assumption, which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction.
-This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.
-
-\neri{Is a literature review nescessary here or is it ingested enough in the text itself? I'm pretty satisfied with how the paragraphs up to now are written.}\newline
-\neri{The entire paragraph going on about SOC, perhaps to long and should be inserted as a sentence somewhere in the literature review? can be seem as comment in source code.}
-% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice.
-On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff).
-These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition.
-Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics.
-In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).
-
-In order to tackle the question of magnitude predictability, we construct the MAGnitude prediction Neural NETwork model, MAGNNET, a neural network (NN) for producing a probability density function (PDF) of the possible magnitudes at a queried space time coordinates, conditioned on the seismic history.
-MAGNNET encodes the seismic history of time series from a hypocentral catalog and produces an estimate of the possible magnitudes at a queried time and place.
-
-MAGNNET's result will be quantitatively compared to other known benchmarks to estimate the amount of information gained by encoding the seismic history.
-MAGNNET is composed of a long-short term memory (LSTM) architecture to model the seismic history up to a given time.
-This latent representation is then passed together with  a space-time coordinate query of a future earthquake into a fully connected neural network (FCNN) that produces a parametrized PDF of the possible magnitudes at the quarried space- time coordinates.
-The task described here, from encoding the seismic history to producing a PDF for a specific time-space querry is depicted in Fig.
-\ref{fig:intro_fig}a.
-A detailed description of the model's architecture is given in the methods section \neri{TODO}.
-We optimize MAGNNET's parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
-\begin{equation}
-    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i} \left( m_i \right) \right) } \rangle\ ,
-    \label{eq:loss_function}
-\end{equation}
-where $p_{\textbf{x}_i, t_i}$ is the probability function returned by the model for the space-time coordinates $\textbf{x}_i, t_i$ (i.e.
-the ``query coordinates''), $m_i$ the corresponding to the actual magnitude of the $i^{th}$ earthquake and $\langle \cdot\rangle$ stands for the empirical average over the data set.
-We use the a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} parameterized family of probability distributions stretched to a relevant domain.
-The final distribution used is defined by 5 parameters.
-This family smoothly can interpolate between a decaying distribution (resembling the GR distribution), and localized distributions whose mass is concentrated around a specific value.
-This allows the model to output both an ``ignorant'' prediction, essentially resembling the GR distribution, and more confident predictions localized around a given magnitude.
-
-We train MAGNNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog [cite] for Southern California, GeoNet [cite] for New Zealand, and the JMA catalog [cite] for Japan.
-While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.?
-cite?) and exhibit varying data quality.
-A separate model is trained for each region, with identical loss function and parameterization of the PDF.
-Sample results of known events worldwide are presented in \ref{fig:intro_fig}b.
-The PDF of each event is shown superimposed with the train set's Gutenberg-Richter distribution, i.e.
-the common vanilla benchmark.
-
-\begin{figure}[h!]
-	\centering
-        \includegraphics[width=1\textwidth]{figures/intro_fig.pdf}
-	\caption{
- Caption.
-}
-\label{fig:intro_fig}
-\end{figure}
-
-The model's, MAGNNET's, performance is evaluated over a time span that was not included in training.
-The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence.
-\neri{I think this is a good place to add that we conclude that there is some preparatory phase.}
-%This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation.
-Examining these results across multiple catalogs and regions highlights the robustness of our methodology and the generalizability of our conclusion, paving the way for further exploration of earthquake predictability.
-
-\section{Results} \label{sec:results}
-
-Focusing on existing events within the catalog due to our model's requirement for labeled data, our analysis utilizes the space-time coordinates of earthquakes as input.
-This necessitates excluding queries about arbitrary spatiotemporal points during training and evaluation, as these processes require knowledge of the true magnitude for each event at its specific location and time.
-Consequently, the raw output for our analysis consists of the model-generated PDFs of possible magnitudes for each event's time and location.
-Figure \ref{fig:model_output}a \neri{TODO: make sure numbering in this figure is correct} showcases PDFs produced by our model, $p_{\textbf{x}_i, t_i}(m)$, for 100 randomly sampled event times and locations from the test set of Southern California.
-Included as a baseline is the stationary GR predictor trained on the training set, representing its potential operational use (e.g., Tsuruoka et al., 2012).
-A clear trend emerges: PDFs predicted for higher-magnitude events (warm colors) exhibit a greater skew towards higher magnitudes compared to those for lower-magnitude events (cooler colors).
-Figure \ref{fig:model_output}b Presents the distribution of lieklihood scores of the test set.
-That is, $p_{\textbf{x}_i, t_i}^{(model)}(magntude label)$, the probability density for retrieving the label magnitude of an event from a given model.
-The solid line indicates the median, percentiles 10 and 90 are indicated by the shaded region.
-Here it can be seen that \neri{rerun this figure before writing conclusions}.
-To qualitativlry compare our model to other benchmark model in more common terms, it is useful to examine the marginal distribution for the magnitudes of the test set.
-\ref{fig:model_output}c presents the marginal PDF, $p(m)$, of MAGNNET compared to other common benchmarks for the Southern California dataset.
-This depicts the general trend that the model suggests and makes the comparison to common benchmarks, such as the GR, closer to the method they are typically used in, as a stationary distribution.
-\ref{fig:model_output}c also shows the histograms of the train and test set superimposed with the marginal PDFs of the models examined.
-It can clearly be seen that GR follows the train set, as expected, while MAGNNET follows the test set.
-Furthermore, the dynamic GR benchmarks such as last $n$ events of the $d$ days, although showing a closer functional for to the test than the train, lack features of the histogram that MAGNNET does not.
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=1\textwidth]{figures/raw_results_hauksson.pdf}
-    \caption{
-        \textbf{outputs of MAGnitude prediction Neural NETwork, MAGNNET, for the Hauksson catalog of Southern California.
-a}, Likelihoods of MAGNNET and some common benchmark magnitude predictors.
-Solid lines represent the median value along the magnitude axis, shading shows the 10th and 90th percentiles of the distribution along the vertical axis.
-\textbf{b}, Train set's Gutenberg-Richter dostribution (dashen black line) with PDFs produced by MAGNNET for 100 randomly sampled events from the test set.
-Each PDF is color coded by the label of the event, its true magnitude.
-Colorbar to the right indicates the color code.
-\textbf{c}, Total magnitude frequency distribution produced by MAGNNET and common benchmarks for the entire test set.
-Histograms of the train (and test) set are presented in orange (blue).
-    }
-    \label{fig:model_output}
-\end{figure}
-
-Quantification of a model's preformance in our scheme will be done by averging on all log-likelihood scores of the test set, as defined in Eq.
-\ref{eq:loss_function}.
-Fig.
-\ref{fig:metrics}a-c shows the minus mean log likelihood score, $\mathcal{L}$ from Eq.
-\ref{eq:loss_function}, for MAGNNET and chosen common benchmarks, for the three regions tested.
-It can be clearly seen that MAGNNET shows a clear an significant average information gain over every benchmark tested.
-\neri{Should we mention other benchmarks and refer to the SM?}.
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=1\textwidth]{figures/combined_batplots.pdf}
-    \caption{
-        \textbf{Metric scores for out model MAGNNET, and common benchmark models.
-a}, \textbf{b}, \textbf{c}, Minus mean information content of our MAGNNET model (red) and other common benchmark magnitude predictors (see labels on bars in the figure).
-\textbf{d, e, f}, Reciever Operating Characteristic (ROC) curve for a binary classifier determinng the existence of a large ($m>=4$) event.
-Area under the curve (AUC) for each curve is written in each frame, colorcoded identically to the bar plots.
-\textbf{g, h, i}, Precision at recall curves for the binary classifier described in d, e, f.
-        }
-        \label{fig:metrics}
-\end{figure}
-
-As the out model's output is a probability density function (PDF), we may easily convert it to a binary classifer.
-By doing this we are posing a modified question: will there be a \textit{large} earthquake in the query space-time coordiantes or not?
-This is done by setting a threshold which defines a "\textit{large}" earthquake, $m_c$ ($=4$ in the examples presented here), then setting examples above that threshold as positive examples and below as negative.
-The probability of a large event occuring in the query coordinates is given by integrating the output PDF above the threshild, $p=\int_{m_c}^{\infty}p(m')dm'$.
-Given such binary classifier, it is possible now to examine the common binary classifier metrics of area under reciever operator (ROC) curve [cite] and the prescision at recall curve [cite].
-\ref{fig:metrics}d-f show the ROC curves for the binary modification for all three regions in question.
-The area under the curve (AUC) are written to the bottom right of each panel color coded by the same legend as the bar plots above.
-It can be seen that throughout all regions MAGNNET preforms significantly better than all other benchmarks, indicated by the largest AUC value.
-The precision at recall curve displayed in \ref{fig:metrics}g-i show a similar result as MAGNNET's curve displays the largest AUC among all benchmarks tested throughout all regions.
-\neri{TODO: calculate and present AUC P@R}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=1\textwidth]{figures/info_over_time_hauksson.pdf}
-    \caption{
-        Caption.
-    }
-    \label{fig:info_gain_over_time}
-\end{figure}
-
-The information gain achieved by our model for each event can be quantified by the difference in log-likelihood (LL) between our model's prediction and the baseline GR predictor:
-\begin{equation}
-    \Delta LL = \log{p_{\pmb{x_i}, t_i}^{(MAGNNET)}(m_i)} - \log{p^{(GR)}(m_i)}
-    \label{eq:information_gain}
-\end{equation}
-This equation measures the amount of information gained by using our MAGNNET model compared to the GR benchmark.
-Analyzing this quantity per event across the entire test set reveals patterns in where our model achieves its advantage.
-Fig.
-\ref{fig:info_gain_over_time}a shows the test set for Southern California event magnitudes by their running index, colored according to the information gain of our model over the GR benchmark.
-This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as the black dashed line in the figure.
-The predominantly increasing trend in cumulative information gain suggests that our model is consistently advantageous as opposed to a concentrated advantage in specific periods.
-Notably, for the Southern California data set, the most rapid information gain occurs during the aftershocks of the two Ridgecrest earthquakes.
-Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4, \ \Delta LL$, respectively, albeit these events are typically considered completely unpredictable.
-This observation leads to a significant corollary: at least some portion of the information gain likely originates from a statistical signal embedded in the event sequence prior to event's occurence.
-This point will be further elaborated upon in the discussion section.
-\newline
-The observed increase in information gain persists throughout the entire test set for all regions examined in this study as the black dashed curves in Fig.
-\ref{fig:info_gain_over_time}b-d present.
-This includes regions previously analyzed by (Ogata et al.
-SRL 2018), where no information gain was detected using the method employed at the time.
-Further analysis of \ref{fig:info_gain_over_time} applied to other regions is presented in the supplementary material.
-\neri{TODO: ad these colorful plots for NZ and JAPAN to the SM}
-The smoothed instantaneous information gain per event is presented in Fig.
-\ref{fig:info_gain_over_time}e-g, and can be seen as commonly well above the threshold of zero information gain, marked by the red dashed line.
-\newline
-While the information gain observed in preshock and mainshocks presented above is encouraging, a potential concern might arise regarding the aftershocks: could their information gain be an artifact of fitting to the temporal incompleteness of the catalog (cite somethong?)?
-To address this, we factor out this potential bias by recalculating the likelihood score of each event, $p(m_i)$ , conditioned on the magnitude exceeding a threshold, $\Tilde{m}$:
-
-\begin{equation}
-    p \left( m \vert m_i > \Tilde{m} \right) = \frac{p_{\pmb{x}_i, t_i}(m)} {\int_{\Tilde{m}}^{\infty} p(m') ,dm'}
-    \label{eq:conditioned_likelihood}
-\end{equation}
-
-For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method (Weimer and Weiss, 2000) within a window of 150 past and 150 future events.
-The resulting temporal incompleteness curves are presented in the SM \neri{TODO: add the $m_c(t)$ figure to SM}.
-After applying the renormalization of Eq.
-\ref{eq:conditioned_likelihood} the cumulative information gain is recalculated.
-The result is presented as the grey dashed curved in Fig.
-\ref{fig:info_gain_over_time}a, as well as in Fig.
-\ref{fig:info_gain_over_time}b-d for all three tested regions.
-\newline
-A second concern regarding temporal variations of the completeness magnitude may be raised to oppose to the analysis and results presented in this work.
-The following paragraph will provide an explanation of our method for varifying that we are indeed measuring a true statistical trend that is not created by the local completeness magnitude artifact.
-We start by deviding each region into bins of size [dx, dy] \neri{TODO fill in the details}.
-The average log likelihood score $\mathcal{L}$ is calculated per bin for both MAGNNET and GR, resulting in a map of spatial average log likelihood, per model.
-The subtraction of the two maps is presented in Fig.
-\ref{fig:info_gain_over_time}h-j, showing the spatial distribution of information gain for each region.
-\neri{TODO complete paragraph about how I select only bins of mc(x,y)<=mc to recalculate the average score and show that I still gain information.}
-for which the local completness magnitude $m_c(x, y)$ is calculated.
-
-% There are two bibliographies in Nature papers.
-One for the main text, and one for the Methods and Extended Data sections.
-The numbering is sequential, meaning that the reference section for the Methods and Extended Data section starts after the last number from the main reference section.
-References should not appear in both sections.
-Any reference used in Methods or Extended Data that also appears in the main bibliography should *only* appear in the main bibliography.
-\let\oldbibliography\thebibliography
-\renewcommand{\thebibliography}[1]{%
-  \oldbibliography{#1}%
-  \setlength{\itemsep}{10pt}%
-}
-% \bibliography{bibliography}
-% \bibliography{bib-article.bib}
-\newpage
-\bibliography{Magnitude_prediction_paper}
-
-\let\oldthebibliography=\thebibliography
-\let\oldendthebibliography=\endthebibliography
-\renewenvironment{thebibliography}[1]{
-    \oldthebibliography{#1}
-    % The number here (34 is an example) is the number of references in the main bibliography, and thus defines the starting number of the first reference in the Methods and Extended Data bibliography.
-    \setcounter{enumiv}{34}
-}{\oldendthebibliography}
-
 % Figure legends appear after the text, not placed in the text.
@@
 During the first submission to the journal, you can include the images in the article file for readability, but if you pass the reviews, they will want the images removed from the main article.
-\newpage
-\unnumbered
-
-\unnumbered
 \section{Methods}
@@
 \textbf{Grey's Note} \textit{My understanding is that you can write the methods section just how you would a normal methods section for a normal scientific article.}
+
+\section{Supplementary information}
+\subsection{Possible information gain from beta spatial variability}
+Various common benchmarks are presented in the main text for comparison to our model's, MAGNET's, performance.
+Variation is beta in treated only in the time domain for two reasons.
+Firstly, time dependent beta is more commonly used than
 
@@
 
-\newpage
-\renewcommand\refname{Methods References}
+\renewcommand\refname{SI References}
 \begin{thebibliography}{10}
 
-\bibitem{kratzert2019towards}
-Frederik Kratzert, Daniel Klotz, Guy Shalev, G{\"u}nter Klambauer, Sepp
-  Hochreiter, and Grey~S Nearing.
-\newblock Towards learning universal, regional, and local hydrological
-  behaviors via machine learning applied to large-sample datasets.
-\newblock {\em Hydrology and Earth System Sciences}, 23(12):5089--5110, 2019.
-
-\bibitem{klotz2022uncertainty}
-Daniel Klotz, Frederik Kratzert, Martin Gauch, Alden~K Sampson, Johannes
-  Brandstetter, G{\"u}nter Klambauer, Sepp Hochreiter, and Grey~S Nearing.
-\newblock Uncertainty estimation with deep learning for rainfall--runoff
-  modeling.
-\newblock {\em Hydrology and Earth System Sciences}, 26(6):1673--1693, 2022.
-
-\bibitem{twb2023gdp}
-IBRD-IDA.
-\newblock {The World Bank Data: Current US\$}.
-\newblock https://data.worldbank.org/indicator/NY.GDP.MKTP.CD, 2023.
-\newblock Accessed: 2023-06-04.
-
-\end{thebibliography}
+\end{thebibliography}\newpage
+\bibliography{Magnitude_prediction_paper_SI}
 

### 2024-05-20 article.tex
finished results. started methods end extended material.

@@
 Earthquakes are notoriously unpredictable, and forecasting seismicity is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes and the scarcity of near-field data~\cite{bernard_earthquake_1999, geller_earthquakes_1997}.
-Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes, including various marginal distributions \yohai{[cite GR and variants]}, scaling relations \yohai{[cite]} and characteristics of both spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}.
+Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes, including various marginal distributions \cite{gutenberg_frequency_1944, kagan_seismic_2002}, scaling relations \cite{bak_earthquakes_1989, dascher-cousineau_what_2020, kagan_aftershock_2002, utsu_centenary_1995} and characteristics of both spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020, devries_deep_2018, king_static_1994}.
 Clearly, these insights can be used to quantitatively inform us about future seismicity based on recent history.
 For example, Omori's law tells us that after big earthquakes we should expect an increase in the local seismicity rate~\cite{omori_after-shocks_1894}.
-Such laws have been incorporated into a variety of forecasting models, which are operationally used today\cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994} \neri{TODO add citation OEF related, Marzocci, Christoforson}.
+Such laws have been incorporated into a variety of forecasting models, which are operationally used today\cite{ogata_statistical_1988, hardebeck_aftershock_2024, jordan_operational_2011, stirling_national_2012}.
 
 However, the above mentioned statistical relations only describe the rate and locations of earthquakes, and very little is known about the dependence of earthquake magnitude on seismic history.
-It is in fact deeply debated whether it is possible to independent of seismic history the magnitude of an earthquake before it starts~\cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}, or even during rupture~\yohai{[cite]}.
-Most modeling approaches assume that event magnitudes are identically and independently drawn from a stationary (or almost stationary) distribution - the Gutenberg Richter distribution (GR), which describes the marginal magnitude distribution over large regions and long times.
-While some variations in magnitude statistics have been described \yohai{[Gulia Weimar, beta as stress sensor papers, Didier's paper]}, they all model history dependence by small and slow changes in the parameters of the GR distribution, and do not inform us about the magnitude of \textit{a specific} future event.
-Even state-of-the-art forecasting models typically assume no knowledge about the magnitude of future events besides the marginal GR distribution, an assumption known as separability \cite{schoenberg_testing_2004}.
-Indeed, Current research on magnitude prediction has failed to show a universal and reproducible advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}.
+It is deeply debated whether it is possible to determine the magnitude of an earthquake before it starts~\cite{kagan_seismic_2002, ogata_exploring_2018}, or even during rupture~\cite{ellsworth_seismic_1995, meier_evidence_2016}.
+Most modeling approaches assume that event magnitudes follow and identical and independent stationary (or almost stationary) distribution - the Gutenberg Richter distribution (GR), which describes the marginal magnitude distribution over large regions and long times.
+While some variations in magnitude statistics have been described \cite{gulia_real-time_2019, nandan_magnitude_2019}, they all model history dependence by weak and slow changes in the parameters of the GR distribution, and do not model the magnitude of \textit{a specific} future event.
+Operationally, state-of-the-art forecasting models typically assume no knowledge about the magnitude of future events besides the marginal GR distribution, an assumption known as separability \cite{schoenberg_testing_2004}.
+Namely, the separability assumption is that the distribution of earthquake magnitudes is statistically independent of the distribution of earthquakes locations and times.
 
 This modeling approach is not a naive or uninformed choice.
-Due to the spatial complexity of the elastic fields, faulting patterns and lithology, the nonlinearities of the rupture process and the complicated interaction between them all, it is not far-fetched to assume that determining the magnitude of an event  requires full microscopic knowledge of the system's state.
-\neri{Should this statement be suported?
-Kinda hand-wavy}
-This philosophy also inspired physical models which describe earthquake dynamics with a statistical approach \yohai{(cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff)}.
-These models reproduce many of the phenomenological  statisitcal relations by positing that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition.
+Current research on magnitude prediction has failed to show a universal and reproducible advantage over a stationary or quasi-stationary GR benchmark~\cite{ogata_exploring_2018, stockman_forecasting_2023}.
+Indeed, due to the spatial complexity of the elastic fields, faulting patterns and lithology, the nonlinearities of the rupture process and the complicated interaction between them all, it is not far-fetched to assume that determining the magnitude of an event requires a detailed microscopic knowledge of the system's state.
+This philosophy also inspired physical models which describe earthquake dynamics with a statistical approach \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}.
+These models reproduce many of the phenomenological statistical relations by positing that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition.
 Under this paradigm, determining the magnitude of an event is indeed impossible using only far-field measurements.
 
-\neri{perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's} \yohai{yes}
+Previous studies have not identified a definitive link between earthquake magnitude and seismic history \cite{petrillo_verifying_2023}\neri{More citations needed}.
+Research focused on predicting future earthquake magnitudes often explores questions distinct from those addressed here.
+While some have yielded interesting results, the absence of comparisons to established benchmarks \cite{shcherbakov_forecasting_2019, panakkat_neural_2007} hinders objective assessment of their efficacy and progress.
+Other studies specifically targeting magnitude prediction have yet to outperform the GR distribution \cite{stockman_forecasting_2023}.
+Thus, previous studies have not yet systematically shown the ability to extract meaningful information from seismic history beyond what a random selection from the GR distribution would provide.
+Consequently, the question regarding the dependence of magnitude on seismic history remains open.
 
 The main goal of this paper is to ask the question directly: can we extract any information about a future earthquake's magnitude from regional seismic history?
-\neri{Stress more: this will imply we have information relevant to prediction tasks.
-Zhuang stated: "you have touched one of the most important problems on earthquake predictability"} To this end, we construct a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence.
-Our model is provided with seismic catalogs which include the location, time and magnitude of past events.
-\neri{Hauksson also has angles, how should this be phrased?} Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided.
+A positive answer would yield two important consequences.
+From a fundamental point of view, it challenges the common belief that earthquake magnitudes are independent of seismic history and are inherently unpredictable.
+Second, it suggests that the separability assumption, which is widely applied in operational earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction.
+This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.
+
+% \neri{Stress more: this will imply we have information relevant to prediction tasks. Zhuang stated: "you have touched one of the most important problems on earthquake predictability"}
+
+To this end, we construct a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence.
+Our model is trained on seismic catalogs containing information about past earthquakes, including location, time, and magnitude.
+In some experiments, we explore the impact of incorporating focal mechanism data as well, demonstrating improved performance.
+Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided.
 Thus, we separate out the task of modeling nucleation statistics and isolate the question of magnitude predictability.
 If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of a specific earthquake is extractable from cataloged properties alone.
-
-This contribution has two important consequences.
-From a fundamental point of view, it challenges the self-organized-criticality viewpoint of earthquake magnitude emergence, which posits that magnitudes are inherently unpredictable and do not depend on seismic history \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}.
-\neri{I think we should consider going softer on the SOC claim - most geologists don't care/ know of it} Second, it suggests that the separability assumption, \neri{where did the explanation about seperability go?
-This paragraph is not clear.} which is widely applied in earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction.
-This may lead to improved forecasting models, and potentially also be used to identify precursory signals in the seismic history of an earthquake.
 
@@
 
-%MAGNET's result will be quantitatively compared to other known benchmarks to estimate the amount of information gained by encoding the seismic history.
-
-MAGNET is composed of two main components. First, the catalog up to time $t$ is passed through a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history.
-Second, this representation combined with a space-time query specifying the coordinates of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, see Fig.
+MAGNET is composed of two main components.
+First, the catalog up to, and not including, time $t$ is encoded by a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history.
+Second, this representation is combined with a space-time query specifying the coordinates of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, see Fig.
 \ref{fig:intro_fig}a.
@@
 \begin{equation}
-    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i(h_t)} \left( m_i \right) \right) } \rangle\ ,
+    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i} \left( m_i \right) \right) } \rangle\ ,
     \label{eq:loss_function}
 \end{equation}
-\neri{I find the $h_t$ is confusing and not nescessary}
-where $\bf{x}_i$, $t_i$, $m_i$ re the location, time and magnitude of the $i$-th event, $p_{\textbf{x}_i, t_i}(h_t)$ is the probability function returned by the model,and $h_t$ is the seismic history up to time $t$.
+where $\bf{x}_i$, $t_i$, $m_i$ are the location, time and magnitude of the $i$-th event, $p_{\textbf{x}_i, t_i}(\cdot)$ is the probability function returned by the model.
 $\langle \cdot\rangle$ stands for the empirical average over all earthquakes in the data set.
-Importantly, during training we only supply the model with queries about spacetime coordinates where earthquakes indeed occured.
-\yohai{we should state that we only consider events about $m_c$}
+Importantly, during training we only supply the model with queries about spacetime coordinates where earthquakes indeed occurred, and only events above the completeness magnitude of the catalog are used as queries.
 
-Technically \neri{find a better word}, the PDF is a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} stretched to a relevant domain, which is a parameteric distribution with 5 parameters, that the network outputs.
-This family smoothly can interpolate between a decaying distribution (resembling the GR distribution), and localized distributions whose mass is concentrated around a specific value.
+Technically, the distribution is modeled as a mixture of two stretched Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} and is parameterized by 5 parameters, which are the output of the network.
+This parameteric family can  smoothly interpolate between a power-law decaying distribution (resembling the GR distribution), and localized distributions with concentrated mass around a specific value.
 This allows the model to output both an ``ignorant'' prediction, essentially resembling the GR distribution, and more confident predictions localized around a given magnitude.
 
-We deployed MAGNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog [cite] for Southern California, GeoNet [cite] for New Zealand, and the JMA catalog [cite] for Japan.
-While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.?
-cite?) and exhibit varying data quality.
-A separate model is trained for each region, with identical loss function and parameterization of the PDF.
-The resulting PDF is presented in \ref{fig:intro_fig}b for a few examples of major events, superimposed on the stationary GR distribution (fitted on the train set), which is the naive benchmark.
+We deployed MAGNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog \cite{hauksson_waveform_2012} for Southern California, GeoNet \cite{gns_geonet_1970} for New Zealand, and the JMA catalog \cite{noauthor_japan_nodate-1} for Japan.
+While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies \neri{Not sure if I need to backthis statement up} and exhibit varying data quality.
+\neri{this one as well} A separate model is trained for each region, with identical loss function and parameterization of the PDF.
+The resulting PDF is presented in Fig.
+\ref{fig:intro_fig}b for a few examples of major events, superimposed on the stationary GR distribution (fitted on the train set), which is the naive benchmark.
 
@@
         \includegraphics[width=1\textwidth]{figures/intro_fig.pdf}
-	\caption{
- Caption.
+	\caption{\textbf{a} MAGNET Architecture: A depiction of the task and architecture of the MAGnitude Neural EsTimation model (MAGNET).
+For each earthquake, seismic history leading up to its occurrence time ($t_i$) is encoded using a Long Short-Term Memory (LSTM) neural network.
+This encoded history is then concatenated with the timing and location of the earthquake.
+The combined information is fed into a fully-connected neural network (FCNN) that outputs a probability density function (PDF) representing the possible magnitudes for the given location.
+\textbf{b} Performance on Major Earthquakes: MAGNET's resulting PDFs for a selection of well-known major earthquakes worldwide.
+The red curve represents the model's predicted PDF, while the dashed black curve represents the Gutenberg-Richter (GR) magnitude-frequency distribution, a common naive benchmark.
+For most earthquakes, the value of the red curve is higher for the actual magnitude (shown by the vertical dashed line), indicating that MAGNET's predictions generally outperform the GR benchmark.
+Notably, the Tōhoku earthquake exhibits a different behavior.
     }
@@
 
-The model's, MAGNET's, performance is evaluated over a time span that was not included in training.
-The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence.
-\neri{I think this is a good place to add that we conclude that there is some preparatory phase.} \yohai{This is not a conclusion that you can draw from our result.
-We can only say that there is some information about the magnitude of an event in the seismic history.} \neri{I think we can say that the information gain is indicative of a preparatory phase.} \yohai{I think we should be careful with this.} \neri{I think we should discuss this.} \yohai{This is quite amazing, I only wrote the first sentence (``This is not a conclusion that you can draw from our result'') and copilot autocompleted the rest of the dialog, correctly using the \textbackslash neri\{...\} and \textbackslash yohai\{\} macros }
+All metrics reported in this work were evaluated over a time span that is not included in training period.
+The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating a significant information gain in forecasting earthquake magnitudes prior to their occurrence.
 %This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation.
@@
 
-Focusing on existing events within the catalog due to our model's requirement for labeled data, our analysis utilizes the space-time coordinates of earthquakes as input.
+% Focusing on existing events within the catalog due to our model's requirement for labeled data, our analysis utilizes the space-time coordinates of earthquakes as input.
 This necessitates excluding queries about arbitrary spatiotemporal points during training and evaluation, as these processes require knowledge of the true magnitude for each event at its specific location and time.
 Consequently, the raw output for our analysis consists of the model-generated PDFs of possible magnitudes for each event's time and location.
-Figure \ref{fig:model_output}a \neri{TODO: make sure numbering in this figure is correct} showcases PDFs produced by our model, $p_{\textbf{x}_i, t_i}(m)$, for 100 randomly sampled event times and locations from the test set of Southern California.
-Included as a baseline is the stationary GR predictor trained on the training set, representing its potential operational use (e.g., Tsuruoka et al., 2012).
-A clear trend emerges: PDFs predicted for higher-magnitude events (warm colors) exhibit a greater skew towards higher magnitudes compared to those for lower-magnitude events (cooler colors).
-Figure \ref{fig:model_output}b Presents the distribution of lieklihood scores of the test set.
-That is, $p_{\textbf{x}_i, t_i}^{(model)}(magntude label)$, the probability density for retrieving the label magnitude of an event from a given model.
-The solid line indicates the median, percentiles 10 and 90 are indicated by the shaded region.
-Here it can be seen that \neri{rerun this figure before writing conclusions}.
-To qualitativlry compare our model to other benchmark model in more common terms, it is useful to examine the marginal distribution for the magnitudes of the test set.
-\ref{fig:model_output}c presents the marginal PDF, $p(m)$, of MAGNET compared to other common benchmarks for the Southern California dataset.
-This depicts the general trend that the model suggests and makes the comparison to common benchmarks, such as the GR, closer to the method they are typically used in, as a stationary distribution.
-\ref{fig:model_output}c also shows the histograms of the train and test set superimposed with the marginal PDFs of the models examined.
-It can clearly be seen that GR follows the train set, as expected, while MAGNET follows the test set.
-Furthermore, the dynamic GR benchmarks such as last $n$ events of the $d$ days, although showing a closer functional for to the test than the train, lack features of the histogram that MAGNET does not.
+
+Our model, MAGNET, demonstrates the ability to capture the dependence of magnitude distribution on seismic history.
+Figure \ref{fig:model_output}a showcases the predicted probability density functions (PDFs) $p_{\textbf{x}_i, t_i}(m)$ for 100 randomly sampled events from the Southern California test set.
+The PDFs exhibit a clear trend: those for higher-magnitude events (warmer colors) are skewed towards larger magnitudes compared to lower-magnitude events (cooler colors).
+For comparison, the stationary GR distribution, the naive predictor, is presented along the model's results.
+This result aligns with the expected behavior for earthquake magnitude predictor: dynamically distributing the probability mass for the expected magnitudes.
+Figure \ref{fig:model_output}b depicts the distribution of likelihood scores for the test set across our model (MAGNET), the stationary Gutenberg-Richter (GR) distribution, and a moving window GR approach \cite{gulia_real-time_2019}.
+Notably, for many high-magnitude events, the model's likelihood scores (red scatter) surpass those of the benchmarks.
+To further assess performance in terms of predicted magnitudes, Figure \ref{fig:model_output}c presents the marginal probability density function (PDF) of magnitudes for MAGNET (red) compared to common benchmarks on the Southern California dataset.
+This figure reveals the general trends predicted by each model and facilitates comparison with the commonly used, stationary GR distribution.
+Furthermore, Fig.
+\ref{fig:model_output}c includes histograms of the train and test set magnitudes alongside the marginal PDFs.
+We observe that the GR distribution closely follows the train set distribution, as expected for a stationary model.
+In contrast, MAGNET aligns well with the test set distribution, suggesting its ability to generalize beyond the training data.
+Notably, dynamic GR benchmarks, exemplified by the "last 300 events" approach,  while exhibiting a better fit to the test set than the stationary GR, lack the detailed features captured by the model's PDF.
 
@@
     \caption{
-        \textbf{outputs of MAGnitude prediction Neural NETwork, MAGNET, for the Hauksson catalog of Southern California.
-a}, Likelihoods of MAGNET and some common benchmark magnitude predictors.
-Solid lines represent the median value along the magnitude axis, shading shows the 10th and 90th percentiles of the distribution along the vertical axis.
-\textbf{b}, Train set's Gutenberg-Richter dostribution (dashen black line) with PDFs produced by MAGNET for 100 randomly sampled events from the test set.
-Each PDF is color coded by the label of the event, its true magnitude.
-Colorbar to the right indicates the color code.
-\textbf{c}, Total magnitude frequency distribution produced by MAGNET and common benchmarks for the entire test set.
+        \textbf{Comparison of MAGNET's outputs to benchmarks'}.
+\textbf{a}, PDFs produced by MAGNET for 100 randomly sampled events from the test set with train set's Gutenberg-Richter distribution (dashed black line).
+The true magnitude label, i.e.
+the true magntude of the event in query, is indicated by the color, interpreted by the colorbar at top.
+\textbf{b}, Likelihoods for returning the magnitude label by MAGNET (red), the 300 events moving window (orange) and the stationary GR (blue dashed line).
+\textbf{c}, Marginal magnitude distribution, i.e.
+stationary $p(m)$, produced by MAGNET and common benchmarks for the entire test set.
 Histograms of the train (and test) set are presented in orange (blue).
+\neri{c: Is there a more convenient way of showing the two superimposed distributions?}
     }
@@
 
-Quantification of a model's preformance in our scheme will be done by averging on all log-likelihood scores of the test set, as defined in Eq.
+Quantification of a model's performance in our scheme will be done by averaging on all log-likelihood scores of the test set, as defined in Eq.
 \ref{eq:loss_function}.
@@
     \caption{
-        \textbf{Metric scores for out model MAGNET, and common benchmark models.
+        \textbf{Metric scores for the MAGNET model, and common benchmark models.
 a}, \textbf{b}, \textbf{c}, Minus mean information content of our MAGNET model (red) and other common benchmark magnitude predictors (see labels on bars in the figure).
-\textbf{d, e, f}, Reciever Operating Characteristic (ROC) curve for a binary classifier determinng the existence of a large ($m>=4$) event.
-Area under the curve (AUC) for each curve is written in each frame, colorcoded identically to the bar plots.
-\textbf{g, h, i}, Precision at recall curves for the binary classifier described in d, e, f.
+A lower score indicated a better preforming model.
+\textbf{d, e, f}, Reciever Operating Characteristic (ROC) and \textbf{g, h, i} the interpolated precision-recall (PR) curve for a binary classifier determining the existence of a large ($m>=4$) event.
+The preformance of such a classifier can be quantified by the area under the curve (AUC), noted in each frame, colorcoded identically to the bar plots.
+For the AUC metrics, a higher score indicated a better preforming classifier.
         }
@@
 
-As the out model's output is a probability density function (PDF), we may easily convert it to a binary classifer.
-By doing this we are posing a modified question: will there be a \textit{large} earthquake in the query space-time coordiantes or not?
-This is done by setting a threshold which defines a "\textit{large}" earthquake, $m_c$ ($=4$ in the examples presented here), then setting examples above that threshold as positive examples and below as negative.
-The probability of a large event occuring in the query coordinates is given by integrating the output PDF above the threshild, $p=\int_{m_c}^{\infty}p(m')dm'$.
-Given such binary classifier, it is possible now to examine the common binary classifier metrics of area under reciever operator (ROC) curve [cite] and the prescision at recall curve [cite].
-\ref{fig:metrics}d-f show the ROC curves for the binary modification for all three regions in question.
-The area under the curve (AUC) are written to the bottom right of each panel color coded by the same legend as the bar plots above.
+We convert our model to a binary classifier posed with the question: will there be a \textit{large} earthquake in the query space-time coordinates or not?
+This can easily be done by setting a threshold which defines a "\textit{large}" earthquake, $m_c$ ($=4$ in the examples presented here), then setting examples above that threshold as positive examples and below as negative.
+The probability of a large event occurring in the query coordinates is given by integrating the output PDF above the threshold, $p=\int_{m_c}^{\infty}p(m')dm'$.
+Given such binary classifier, it is possible to examine the common binary classifier metrics of area under receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010} which suit better an imbalanced data set such as ours.
+\ref{fig:metrics}d-f and g-i show the ROC and interpolated PR curves respectively, for all three regions in examined.
+The area under the curve (AUC) values are written in the inset of each panel, color coded by the same legend as the bar plots above.
 It can be seen that throughout all regions MAGNET preforms significantly better than all other benchmarks, indicated by the largest AUC value.
-The precision at recall curve displayed in \ref{fig:metrics}g-i show a similar result as MAGNET's curve displays the largest AUC among all benchmarks tested throughout all regions.
-\neri{TODO: calculate and present AUC P@R}
 
@@
     \caption{
-        Caption.
+        \textbf{Analysis of information gain of MAGNET over the benchmark.
+a,} Information gain of individual events in the Southern California test set.
+Scattered dots indicate magnitude at event index, colored by the information gain per event (colorbar indicated scale), here warm is an advantage for MAGNET, while cool is an advantage for the benchmark.
+Secondary horizontal axis (grey) indicates corresponds origin time.
+Cumulative Information gain (dashed black) and cumulative conditioned information gain (dashed grey) curves are superimposed on the scatter, demonstrating a constantly increasing function.
+\textbf{b-d}, Information gain (dashed black) and cumulative conditioned information gain (dashed grey) curves for Southern California, New Zealand and Japan test sets, respectively, all three regions examined in this study.\textbf{e-g}, Smoothed instantaneous information gain.
+Grey cloud shows the standard deviation of the smoothing window.
+All three curves are predominantly above the zero information gain line (dashed red).
+\textbf{h-i} Spatial distribution of information gain for all regions in study.
+Maps show the difference in mean information content per bin, model's advantage is indicated by warm colors, disadvantage by cool colors.
     }
@@
 \begin{equation}
-    \Delta LL = \log{p_{\pmb{x_i}, t_i}^{(MAGNET)}(m_i)} - \log{p^{(GR)}(m_i)}
+    \Delta LL_i = \log{p_{\pmb{x_i}, t_i}^{(MAGNET)}(m_i)} - \log{p^{(GR)}(m_i)}
     \label{eq:information_gain}
 \end{equation}
-This equation measures the amount of information gained by using our MAGNET model compared to the GR benchmark.
-Analyzing this quantity per event across the entire test set reveals patterns in where our model achieves its advantage.
+This equation measures the amount of information gained by using our MAGNET model rather than the GR benchmark, for a specific event.
+Analyzing this quantity per event across the entire test set reveals patterns indicating where and when our model achieves its advantage.
 Fig.
-\ref{fig:info_gain_over_time}a shows the test set for Southern California event magnitudes by their running index, colored according to the information gain of our model over the GR benchmark.
+\ref{fig:info_gain_over_time}a shows the test set for Southern California events magnitudes by their running index, colored according to the information gain as defined in Eq.
+\ref{eq:information_gain}.
 This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as the black dashed line in the figure.
@@
 Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4, \ \Delta LL$, respectively, albeit these events are typically considered completely unpredictable.
-This observation leads to a significant corollary: at least some portion of the information gain likely originates from a statistical signal embedded in the event sequence prior to event's occurence.
+This observation leads to a significant corollary: at least some portion of the information gain likely originates from a statistical signal embedded in the event sequence prior to event's occurrence.
 This point will be further elaborated upon in the discussion section.
 \newline
-The observed increase in information gain persists throughout the entire test set for all regions examined in this study as the black dashed curves in Fig.
-\ref{fig:info_gain_over_time}b-d present.
-This includes regions previously analyzed by (Ogata et al.
-SRL 2018), where no information gain was detected using the method employed at the time.
-Further analysis of \ref{fig:info_gain_over_time} applied to other regions is presented in the supplementary material.
-\neri{TODO: ad these colorful plots for NZ and JAPAN to the SM}
-The smoothed instantaneous information gain per event is presented in Fig.
-\ref{fig:info_gain_over_time}e-g, and can be seen as commonly well above the threshold of zero information gain, marked by the red dashed line.
-\newline
-While the information gain observed in preshock and mainshocks presented above is encouraging, a potential concern might arise regarding the aftershocks: could their information gain be an artifact of fitting to the temporal incompleteness of the catalog (cite somethong?)?
-To address this, we factor out this potential bias by recalculating the likelihood score of each event, $p(m_i)$ , conditioned on the magnitude exceeding a threshold, $\Tilde{m}$:
+
+Our analysis reveals a persistent increase in information gain across the test set for all investigated regions (Fig.
+\ref{fig:info_gain_over_time}b-d).
+This includes regions previously studied where no information gain was detected using different methods  \cite{ogata_exploring_2018}.
+Similar domain representations for additional regions can be found in the supplementary material  \neri{TODO: add these colorful plots for NZ and JAPAN to the SM}.
+Furthermore, the smoothed instantaneous information gain per event, presented in Fig.
+\ref{fig:info_gain_over_time}e-g, consistently exceeds the threshold for zero information gain marked by the red dashed line.
+
+While the information gain observed in preshock and mainshocks presented above is encouraging, we address a potential concern regarding information gain during high temporal incompleteness periods\cite{stockman_forecasting_2023}.
+In order to account for the temporal incompleteness artifact, and factor it out of our information gain calculation, we recalculate the likelihood score of each event, $p(m_i)$ , conditioned on the magnitude exceeding a threshold, $\Tilde{m}$:
 
@@
 \ref{fig:info_gain_over_time}b-d for all three tested regions.
-\newline
-A second concern regarding temporal variations of the completeness magnitude may be raised to oppose to the analysis and results presented in this work.
-The following paragraph will provide an explanation of our method for varifying that we are indeed measuring a true statistical trend that is not created by the local completeness magnitude artifact.
-We start by deviding each region into bins of size [dx, dy] \neri{TODO fill in the details}.
-The average log likelihood score $\mathcal{L}$ is calculated per bin for both MAGNET and GR, resulting in a map of spatial average log likelihood, per model.
-The subtraction of the two maps is presented in Fig.
-\ref{fig:info_gain_over_time}h-j, showing the spatial distribution of information gain for each region.
-\neri{TODO complete paragraph about how I select only bins of mc(x,y)<=mc to recalculate the average score and show that I still gain information.}
-for which the local completness magnitude $m_c(x, y)$ is calculated.
+Evidently, all conditioned cumulative information gain curves show a steadily increasing trend, supporting the hypothesis that there is information content about the upcoming magnitude in the statistical properties of the catalog, excluding the temporal incompleteness artifact.
+The mean LL scores of the test set for each data catalog are presented in the supplementary material.
+\neri{TODO: add conditioned bar plots to SM}.
+
+Beyond the temporal domain, MAGNET also exhibits an advantageous information gain distribution spatially (Fig.
+\ref{fig:info_gain_over_time}h-j).
+Each spatial bin's color represents the average log-likelihood difference between MAGNET's predictions and the common GR benchmark (as defined in Eq.
+\ref{eq:information_gain}, averaged and applied per bin).
+Warm colors indicate regions where MAGNET performs better.
+While the advantage in Southern California appears spatially diffuse, both New Zealand and Japan display concentrations of high advantage at the dataset edges, further from the mainland.
+We hypothesize this is linked to the high local magnitude of completeness, $m_c(x,y)$, in these regions.
+To address potential artifacts from spatially varying $m_c(x,y)$, we conducted further analysis.
+We computed local $m_c(x,y)$ per bin and selected only bins with completeness lower than the entire training set's completeness magnitude, $m_c \left( x,y \right)  \leq m_c^{ \left( train \right) }$.
+We then used these bins to create a minimal dataset and re-ran the model with all previous metrics.
+As in prior experiments, MAGNET outperformed benchmarks on this minimal dataset.
+Details and results of this analysis are provided in the supplementary material.
+\neri{TODO: add these results of minimal data set to the SM}.
 
@@
 \section{Methods}
-\subsection{AI Model}
-\textbf{Grey's Note} \textit{My understanding is that you can write the methods section just how you would a normal methods section for a normal scientific article.}
+\subsection{Neural architecture}
+A detailed visual of the model's architecture is presented in Extended Data Fig. \ref{fig:architecture}. The examined catalog is used as an input into three distinct encoding components:
+\subparagraph{\textbf{Recent Earthquakes Encoder}} Calculates a set of predefined functions on various past time windows.
+This encompasses a representation of various statistical properties of the catalog.
+The result is then used as an input to a LSTM NN.
+A detailed description of this encoder can be found in past study \cite{zlydenko_neural_2023}.
+
+\subparagraph{\textbf{Seismicity Rate Encoder}} Computes a rough estimation of the amount of energy released in a given region around the query coordinate, in a given past time window.
+The result is then used as an input to a LSTM NN.
+A detailed description of this encoder can be found in past study \cite{zlydenko_neural_2023}.
+
+\subparagraph{\textbf{Event Time Encoder}} Prepares the query time-space coordinates.
+Time is represented as the time difference from the most recent past event, in seconds.
+No further processing is done in this encoder.
+
+Outputs of all encoders are concatenated and used as the input to a Fully Connected Neural Network (FCNN).
+The sizes of layers in the FCNN, as the number of layers themselves, are used as tuneable hyper parameters in order to optimize the architecture by cross validation.
+A $tanh$ function is used as the activation on the output of every hidden layer, and a $softplus$ as an activation on the output layer.
+
+\subsection{Loss metric}
+To obtain the Probability Density Function of the magnitudes ($p(m)$) for each query coordinates ($\textbf{x}_i, t_i$) we maximize the likelihood presented in Eq.
+\ref{eq:loss_function}.
+For our usage we choose $p(\cdot)$ to be a mixture of two stretched and shifted Kumaraswamy distributions \cite{kumaraswamy_generalized_1980}:
+\begin{equation}
+    p\left( m \right)
+    \equiv
+    \sum_{j=1,2} \frac{A_j}{\sigma}a_jb_j\left(\frac{m-m_c}{\sigma}\right)^{\left(a_j-1\right)}\left(1-\left(\frac{m-m_c}{\sigma}\right)^{a_j}\right)^{\left(b_j-1\right)}
+\end{equation}
+Here $a_j, \ b_j$ are the parameters of the Kumaraswamy distribution.
+$m_c$ is the train set's completeness magnitude.
+$m_c$ together with $\sigma$ define the support of $p(m)$ to $[m_c, m_c+\sigma]$.
+$A_j$ is a normalization prefactor, with $j$ the summation index defining the PDF mixture.
+
+\subsection{Definition of benchmark models}
+\subparagraph{Last $n$ events} The Gutenberg- Richter (GR) distribution fitted for the past $n$ events.
+This method follows the method presented in \cite{gulia_real-time_2019} for constructing a b-value time-series.
+We use the train set's completeness magnitude $m_c^{train}$ to define the GR distribution, and a minimal number of events above $m_c$ ($=10$ in this study).
+If the minimal number of events above completeness is not fulfilled the window is expanded into the past until the condition is achieved.
+An attempt to compute $m_c$ per window has been made, though neglected as it showed worse preformance.
+
+\subparagraph{Last $d$ days} The Gutenberg- Richter (GR) distribution fitted for the events in the past $d$ days.
+This method is identical to the previous method in all but the definition of the window selection.
+
+\subparagraph{Spatially varying Gutenberg Richter}
+\subsection{Measurement of temporal incompleteness}
 
@@
 \begin{thebibliography}{10}
-
-\bibitem{kratzert2019towards}
-Frederik Kratzert, Daniel Klotz, Guy Shalev, G{\"u}nter Klambauer, Sepp
-  Hochreiter, and Grey~S Nearing.
-\newblock Towards learning universal, regional, and local hydrological
-  behaviors via machine learning applied to large-sample datasets.
-\newblock {\em Hydrology and Earth System Sciences}, 23(12):5089--5110, 2019.
-
-\bibitem{klotz2022uncertainty}
-Daniel Klotz, Frederik Kratzert, Martin Gauch, Alden~K Sampson, Johannes
-  Brandstetter, G{\"u}nter Klambauer, Sepp Hochreiter, and Grey~S Nearing.
-\newblock Uncertainty estimation with deep learning for rainfall--runoff
-  modeling.
-\newblock {\em Hydrology and Earth System Sciences}, 26(6):1673--1693, 2022.
-
-\bibitem{twb2023gdp}
-IBRD-IDA.
-\newblock {The World Bank Data: Current US\$}.
-\newblock https://data.worldbank.org/indicator/NY.GDP.MKTP.CD, 2023.
-\newblock Accessed: 2023-06-04.
-
+\bibitem{zlydenko_neural_2023}
+Zlydenko, O. et al. A neural encoder for earthquake rate forecasting. Sci Rep 13, 12350 (2023).
 \end{thebibliography}
@@
 An example is below.}
-
-The authors declare no competing interests.
+% For example:
+% The authors declare no competing interests.
 Please contact either the first author (Grey Nearing; \href{mailto:nearing@google.com}{nearing@google.com}) or Avinatan Hassidim; \href{mailto:avinatan@google.com}{avinatan@google.com}) for correspondence and requests, including questions regarding reprints and permissions.
@@
 \section*{Extended Data}
+\begin{enumerate}
+    \item Table: scores of all metrics - \textbf{should I add this??}
+    \item Detailed architecture? - \textbf{should I add this??}
+    \item Colorful PDF examples for NZ and Japan
+    \item Marginal distributions for NZ and Japan
+    \item Colored scatter magnitude vs index, NZ and Japan - \textcolor{red}{\textbf{Done}}
+    \item $m_c(t)$ for all three regions -  - \textcolor{green}{kind of finished plot, not sure}
+    \item Bar plots of conditioned $<LL>$
+    \item Spatial $m_c(x, y)$ and low $m_c(x, y) \leq m_c^{train}$ bins
+\end{enumerate}
+
+\newpage
+\begin{figure}
+    \centering
+    \includegraphics[width=1\textwidth]{figures/detailed_architecture.pdf}
+    \caption{\textbf{MAGNET architecture.} Described in detail in the methods section.}
+    \label{fig:architecture}
+\end{figure}
+
+\newpage
+\begin{figure}[h!]
+	\centering
+        \includegraphics[width=1\textwidth]{figures/info_gain_scatter_NZ_and_Japan.pdf}
+	\caption{
+            \textbf{Information gain over time domain for New Zealand and Japan data sets.} This figure presents results for the data sets not displayed in Fig.
+\ref{fig:info_gain_over_time}a.
+\textbf{a} Entire test set of New Zealand and a selected time span (\textbf{b}) focusing on the 2016 Te Araroa earthquake and the 2016 Kaikōura earthquake.
+\textbf{c} The entire Japan data set and a focused time span \textbf{d} of the 2011 Tōhoku earthquake and some of its aftershocks.
+         }
+\label{fig:intro_fig}
+\end{figure}
+
+\newpage
+
 \textbf{Grey's Note} \textit{You get up to 10 items in the Extended Data section.

### 2024-05-20 article.tex
Perhaps done with methods?

@@
 
-\subparagraph{Spatially varying Gutenberg Richter}
+\subparagraph{Spatially varying Gutenberg Richter} Estimation of the local $\beta$ value is done by following the method of \cite{taroni_highdefinition_2021}.
+Seismicity data is first binned into $0.1^\circ \times 0.1^\circ$, where the GR distribution is calculated for.
+beta for a specified location $x_i$ calculated by smoothing the binned values over a radius of 30km around $x_i$.
+Estimation which consider less than 100 events are discarded.
+
 \subsection{Measurement of temporal incompleteness}
+Temporal incompleteness is measured at the times of events by fitting the completeness magnitude to a window of 300 events: 150 past events, 1 current event, 149 future events.
+The window is constructed by the same algorithm, described to construct the \textit{ Last $n$ events benchmark}.
 
@@
 Zlydenko, O. et al. A neural encoder for earthquake rate forecasting. Sci Rep 13, 12350 (2023).
+\bibitem{taroni_highdefinition_2021}
+Taroni, M., Zhuang, J. \& Marzocchi, W. High‐Definition Mapping of the Gutenberg–Richter b‐Value and Its Relevance: A Case Study in Italy. Seismological Research Letters 92, 3778–3784 (2021).
+
 \end{thebibliography}

### 2024-05-20 article.tex
mostly typos

@@
 A lower score indicated a better preforming model.
-\textbf{d, e, f}, Reciever Operating Characteristic (ROC) and \textbf{g, h, i} the interpolated precision-recall (PR) curve for a binary classifier determining the existence of a large ($m>=4$) event.
-The preformance of such a classifier can be quantified by the area under the curve (AUC), noted in each frame, colorcoded identically to the bar plots.
+\textbf{d, e, f}, Receiver Operating Characteristic (ROC) and \textbf{g, h, i} the interpolated precision-recall (PR) curve for a binary classifier determining the existence of a large ($m>=4$) event.
+The performance of such a classifier can be quantified by the area under the curve (AUC), noted in each frame, color coded identically to the bar plots.
 For the AUC metrics, a higher score indicated a better preforming classifier.
+        \neri{Organize titles and font sizes}
         }
@@
 Temporal incompleteness is measured at the times of events by fitting the completeness magnitude to a window of 300 events: 150 past events, 1 current event, 149 future events.
-The window is constructed by the same algorithm, described to construct the \textit{ Last $n$ events benchmark}.
+The window is constructed by the same algorithm, used to construct the \textit{Last $n$ events benchmark}.
 
-\section*{Data Availability}
-This is a required section. Guidelines for data availability: \url{https://www.nature.com/documents/nr-data-availability-statements-data-citations.pdf}.
-
-\section*{Code Availability}
-\textbf{Grey's Note} \textit{This is a required section. If your article is about AI or ML, the editor will ask you to make the weights of a trained model available.}
+\section*{Data and code availability}
+The datasets analysed during the current study are available in the relevant references mentioned throught the article.
+The code for analysing the data is available in \textcolor{red}{\textbf{**link TBD**}}
+% This is a required section. Guidelines for data availability: \url{https://www.nature.com/documents/nr-data-availability-statements-data-citations.pdf}.
 
@@
 \section*{Acknowledgements}
+\yohai{Legit to use your comment to give you tasks? TODO by \textbf{Yohai}}
 Acknowledgements should be brief, and should not include thanks to anonymous referees and editors, inessential words, or effusive comments.
@@
 \section*{Author Contributions}
-Author Contributions: Authors are required to include a statement to specify the contributions of each co-author.
-The statement can be up to several sentences long, describing the tasks of individual authors referred to by their initials.
-See the authorship policy page for further explanation and examples.
+N.B., O.Z., O.G., Y.B.S.
+N.B., O.Z., O.G.
+and Y.B.S.
+designed the research.
+N.B.
+conducted the experiments.
+N.B., O.G.
+and Y.B.S.
+preformed the statistical analysis.
+N.B.
+and O.Z.
+wrote the code base.
+N.B.
+and Y.B.S.
+wrote the manuscript.
+All authors reviewed the manuscript.
 
 \section*{Author Information}
+\yohai{Legit to use your comment to give you tasks? TODO by \textbf{Yohai} x2}
 \textbf{Grey's Note} \textit{Two things are required in this Author Information section: (1) A statement about competing interests.
@@
 \begin{enumerate}
-    \item Table: scores of all metrics - \textbf{should I add this??}
-    \item Detailed architecture? - \textbf{should I add this??}
+    \item Table: scores of all metrics - \textcolor{red}{\textbf{should I add this??}}
+    \item Detailed architecture? - \textcolor{red}{\textbf{Make a nice version}}
     \item Colorful PDF examples for NZ and Japan
     \item Marginal distributions for NZ and Japan
-    \item Colored scatter magnitude vs index, NZ and Japan - \textcolor{red}{\textbf{Done}}
-    \item $m_c(t)$ for all three regions -  - \textcolor{green}{kind of finished plot, not sure}
+    \item Colored scatter magnitude vs index, NZ and Japan - \textcolor{blue}{\textbf{Done}}
+    \item $m_c(t)$ for all three regions -  - \textcolor{blue}{Done}
     \item Bar plots of conditioned $<LL>$
@@
 \end{enumerate}
-
-\newpage
-\begin{figure}
-    \centering
-    \includegraphics[width=1\textwidth]{figures/detailed_architecture.pdf}
-    \caption{\textbf{MAGNET architecture.} Described in detail in the methods section.}
-    \label{fig:architecture}
-\end{figure}
-
-\newpage
-\begin{figure}[h!]
-	\centering
-        \includegraphics[width=1\textwidth]{figures/info_gain_scatter_NZ_and_Japan.pdf}
-	\caption{
-            \textbf{Information gain over time domain for New Zealand and Japan data sets.} This figure presents results for the data sets not displayed in Fig.
-\ref{fig:info_gain_over_time}a.
-\textbf{a} Entire test set of New Zealand and a selected time span (\textbf{b}) focusing on the 2016 Te Araroa earthquake and the 2016 Kaikōura earthquake.
-\textbf{c} The entire Japan data set and a focused time span \textbf{d} of the 2011 Tōhoku earthquake and some of its aftershocks.
-         }
-\label{fig:intro_fig}
-\end{figure}
-
-\newpage
 
@@
 
+\newpage
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/detailed_architecture.pdf}
+    \caption{\neri{TODO: this needs to be nicer} \textbf{MAGNET architecture.} Described in detail in the methods section.}
+    \label{fig:architecture}
+\end{figure}
+
+\newpage
+\begin{figure}[h!]
+	\centering
+        \includegraphics[width=1\textwidth]{figures/info_gain_scatter_NZ_and_Japan.pdf}
+	\caption{
+            \textbf{Information gain over time domain for New Zealand and Japan data sets.} This figure presents results for the data sets not displayed in Fig.
+\ref{fig:info_gain_over_time}a.
+\textbf{a} Entire test set of New Zealand and a selected time span (\textbf{b}) focusing on the 2016 Te Araroa earthquake and the 2016 Kaikōura earthquake.
+\textbf{c} The entire Japan data set and a focused time span \textbf{d} of the 2011 Tōhoku earthquake and some of its aftershocks.
+         }
+\label{fig:nz_japan_info_gain}
+\end{figure}
+
+\newpage
+\begin{figure}[h!]
+    \centering
+        \includegraphics[width=1\textwidth]{figures/temporal_incompleteness.pdf}
+    \caption{
+    \textbf{Temporal Incompleteness for test sets of all regions examined.} Calculated temporal incompleteness for test sets is presented in red, superimposed over the events above (black) and below (grey) the completeness magnitude of the train set, $m_c^{(train)}$.
+Events are plotted by the index of their appearance in the test set.
+Events of low magnitude, $m_c^{(train)}$ (grey), that are not included in the evaluation, are plotted at their interpolated location along the horizontal axis.
+Data for Japan (c) is displayed only near the 2011 Tōhoku earthquake to reduce data density and enhance clarity of the figure.
+    }
+    \label{fig:temp_incompleteness}
+\end{figure}
+
 \end{document}

### 2024-05-30 article.tex
Added extended material tables.

@@
 
-However, the above mentioned statistical relations only describe the rate and locations of earthquakes, and very little is known about the dependence of earthquake magnitude on seismic history.
+However, these statistical relations only describe the rate and locations of earthquakes, and very little is known about the dependence of earthquake magnitude on seismic history.
 It is deeply debated whether it is possible to determine the magnitude of an earthquake before it starts~\cite{kagan_seismic_2002, ogata_exploring_2018}, or even during rupture~\cite{ellsworth_seismic_1995, meier_evidence_2016}.
-Most modeling approaches assume that event magnitudes follow and identical and independent stationary (or almost stationary) distribution - the Gutenberg Richter distribution (GR), which describes the marginal magnitude distribution over large regions and long times.
+Most modeling approaches assume that event magnitudes follow an identical and independent stationary (or almost stationary) distribution - the Gutenberg Richter distribution (GR), which describes the marginal magnitude distribution over large regions and long times.
 While some variations in magnitude statistics have been described \cite{gulia_real-time_2019, nandan_magnitude_2019}, they all model history dependence by weak and slow changes in the parameters of the GR distribution, and do not model the magnitude of \textit{a specific} future event.
 Operationally, state-of-the-art forecasting models typically assume no knowledge about the magnitude of future events besides the marginal GR distribution, an assumption known as separability \cite{schoenberg_testing_2004}.
-Namely, the separability assumption is that the distribution of earthquake magnitudes is statistically independent of the distribution of earthquakes locations and times.
+Namely, the separability assumption is that the distribution of earthquake magnitudes is statistically independent of their locations and times.
 
@@
 
-Previous studies have not identified a definitive link between earthquake magnitude and seismic history \cite{petrillo_verifying_2023}\neri{More citations needed}.
-Research focused on predicting future earthquake magnitudes often explores questions distinct from those addressed here.
-While some have yielded interesting results, the absence of comparisons to established benchmarks \cite{shcherbakov_forecasting_2019, panakkat_neural_2007} hinders objective assessment of their efficacy and progress.
-Other studies specifically targeting magnitude prediction have yet to outperform the GR distribution \cite{stockman_forecasting_2023}.
-Thus, previous studies have not yet systematically shown the ability to extract meaningful information from seismic history beyond what a random selection from the GR distribution would provide.
-Consequently, the question regarding the dependence of magnitude on seismic history remains open.
+Multiple previous studies have not identified a definitive link between earthquake magnitude and seismic history \cite{petrillo_verifying_2023, taroni_are_2024, davidsen_are_2011}.
+Some studies have shown circumstantial evidence that earthquake magnitudes may not follow a strictly stationary distribution, e.g.~by showing that statistics are not invariant to permutation of the order, or by showing that the maximal magnitude in a given time frame can be predicted with a better-than-random chance~\cite{xiong_seismic_2023, corral_comment_2005, spassiani_exploring_2016, lippiello_positive_2024, lippiello_positive_2024-1, lippiello_influence_2008, shcherbakov_forecasting_2019, panakkat_neural_2007}.
+However, a significant information gain over the GR benchmark in predicting the magnitude of a given earthquake was not demonstrated, and to the best of our knowledge was not attempted.
+The debate regarding the dependence of magnitude on seismic history remains open.
 
-The main goal of this paper is to ask the question directly: can we extract any information about a future earthquake's magnitude from regional seismic history?
-A positive answer would yield two important consequences.
-From a fundamental point of view, it challenges the common belief that earthquake magnitudes are independent of seismic history and are inherently unpredictable.
+The main goal of this paper is to ask the question directly: can we extract any information about the magnitude of a specific future earthquake from regional seismic history?
+A positive answer would have two important consequences.
+From a fundamental point of view, it challenges the common belief that earthquake magnitudes are inherently unpredictable.
 Second, it suggests that the separability assumption, which is widely applied in operational earthquake forecasting, may be replaced by a more nuanced model that incorporates the seismic history into the magnitude prediction.
@@
 Our model is trained on seismic catalogs containing information about past earthquakes, including location, time, and magnitude.
-In some experiments, we explore the impact of incorporating focal mechanism data as well, demonstrating improved performance.
+In some experiments, we also incorporate focal mechanism information, demonstrating improved performance.
 Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided.
@@
 MAGNET is composed of two main components.
-First, the catalog up to, and not including, time $t$ is encoded by a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history.
-Second, this representation is combined with a space-time query specifying the coordinates of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, see Fig.
+First, the catalog up to time $t$ (not including) is encoded by a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history \yohai{cite LSTM paper}.
+Second, this representation is combined with a space-time query specifying the coordinates, $\bf{x}_i$, $t_i$, of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, $p_{\textbf{x}_i, t_i}(m)$, see Fig.
 \ref{fig:intro_fig}a.
-A detailed description of the model's architecture is given in the methods section \neri{TODO}.
-We optimize MAGNET's parameters to maximize the log likelihood (LL) of the observed magnitudes,
-\begin{equation}
-    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i} \left( m_i \right) \right) } \rangle\ ,
-    \label{eq:loss_function}
-\end{equation}
-where $\bf{x}_i$, $t_i$, $m_i$ are the location, time and magnitude of the $i$-th event, $p_{\textbf{x}_i, t_i}(\cdot)$ is the probability function returned by the model.
-$\langle \cdot\rangle$ stands for the empirical average over all earthquakes in the data set.
-Importantly, during training we only supply the model with queries about spacetime coordinates where earthquakes indeed occurred, and only events above the completeness magnitude of the catalog are used as queries.
 
@@
 
+The model parameters are optimized to maximize the log likelihood (LL) of the observed magnitudes.
+Importantly, during training we only supply the model with queries about spacetime coordinates where earthquakes indeed occurred, and only events above the completeness magnitude of the catalog are used as queries.
+A detailed description of the model's architecture is given in the methods section \neri{TODO}.
+
 We deployed MAGNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog \cite{hauksson_waveform_2012} for Southern California, GeoNet \cite{gns_geonet_1970} for New Zealand, and the JMA catalog \cite{noauthor_japan_nodate-1} for Japan.
-While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies \neri{Not sure if I need to backthis statement up} and exhibit varying data quality.
-\neri{this one as well} A separate model is trained for each region, with identical loss function and parameterization of the PDF.
+While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies and exhibit varying data quality.
+A separate model is trained for each region, with identical loss function and parameterization of the PDF.
 The resulting PDF is presented in Fig.
@@
 \section{Results} \label{sec:results}
+As explained above, the model outputs a PDF $p_{\textbf{x}_i, t_i}$ for each earthquake in the training set.
+We denote $\ell_i=\log\left(p_{\textbf{x}_i, t_i}(m_i)\right)$, the log-likelihood of the true magnitude $m_i$ as predicted by MAGNET.
+The model is trained to maximize the average log-likelihood of the true magnitudes in the training set, $\mathcal{L} = -\langle \log{ \ell_i } \rangle$, where $\langle \cdot\rangle$ stands for the empirical average over all earthquakes in the data set.
 
@@
 
-Our model, MAGNET, demonstrates the ability to capture the dependence of magnitude distribution on seismic history.
-Figure \ref{fig:model_output}a showcases the predicted probability density functions (PDFs) $p_{\textbf{x}_i, t_i}(m)$ for 100 randomly sampled events from the Southern California test set.
-The PDFs exhibit a clear trend: those for higher-magnitude events (warmer colors) are skewed towards larger magnitudes compared to lower-magnitude events (cooler colors).
-For comparison, the stationary GR distribution, the naive predictor, is presented along the model's results.
-This result aligns with the expected behavior for earthquake magnitude predictor: dynamically distributing the probability mass for the expected magnitudes.
-Figure \ref{fig:model_output}b depicts the distribution of likelihood scores for the test set across our model (MAGNET), the stationary Gutenberg-Richter (GR) distribution, and a moving window GR approach \cite{gulia_real-time_2019}.
-Notably, for many high-magnitude events, the model's likelihood scores (red scatter) surpass those of the benchmarks.
+Figure \ref{fig:model_output}a shows the predicted PDFs for 100 randomly sampled events from the Southern California test set.
+The PDFs exhibit a clear trend: those that were calculated for higher-magnitude events are skewed towards larger magnitudes (warmer colors) compared to lower-magnitude events (cooler colors).
+This aligns with the expected behavior for an earthquake magnitude predictor.
+For comparison, the stationary GR distribution, the naive predictor which is independent of (recent) seismic history, is presented along the model's results.
+
+This is a qualitative demonstration that the model works as expected.
+For a quantitative comparison, we calculate the $\mathcal{L}=-\langle \ell_i \rangle$ over all events in the test set, and benchmark it against state-of-the-art models:
+the stationary Gutenberg-Richter (GR) distribution (fitted over the training set), and a few variants of a moving window GR approach~\cite{gulia_real-time_2019}.
+It is seen that MAGNET significantly and consistently outperforms all benchmarks across all three regions, as shown in Fig.
+\ref{fig:metrics}a-c.
+Further comparison to other benchmarks is given in the supplementary material \yohai{TODO: add the SM}.
+
+In addition to comparing the log-likelihood of the observed data, one can convert the prediction problem to a binary question: will the earthquake at the spacetime query $\textbf{x}_i, t_i$ have be higher than a threshold $m_t$?
+This can be easily calculated from the prediction, as the probability that $m_i>m_t$ is simply $\int_{m_t}^{\infty}p(m')dm'$.
+This task is more interpretable (will there be a large earthquake?) and can be evaluated using standard binary classification metrics such as area under receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010} which suit better an imbalanced data set such as ours.
+Fig.~\ref{fig:metrics}d-f and g-i show these metrics for all three regions in examined and compare against the GR-variant models.
+Again, MAGNET outperforms all benchmarks.
+
 To further assess performance in terms of predicted magnitudes, Figure \ref{fig:model_output}c presents the marginal probability density function (PDF) of magnitudes for MAGNET (red) compared to common benchmarks on the Southern California dataset.
@@
     \centering
-    \includegraphics[width=1\textwidth]{figures/raw_results_hauksson.pdf}
+    \includegraphics[width=0.5\textwidth]{figures/raw_results.pdf}
     \caption{
-        \textbf{Comparison of MAGNET's outputs to benchmarks'}.
+        \textbf{Benchmarking MAGNET}.
 \textbf{a}, PDFs produced by MAGNET for 100 randomly sampled events from the test set with train set's Gutenberg-Richter distribution (dashed black line).
@@
 
-Quantification of a model's performance in our scheme will be done by averaging on all log-likelihood scores of the test set, as defined in Eq.
-\ref{eq:loss_function}.
-Fig.
-\ref{fig:metrics}a-c shows the minus mean log likelihood score, $\mathcal{L}$ from Eq.
-\ref{eq:loss_function}, for MAGNET and chosen common benchmarks, for the three regions tested.
-It can be clearly seen that MAGNET shows a clear an significant average information gain over every benchmark tested.
-\neri{Should we mention other benchmarks and refer to the SM?}.
-
 \begin{figure}[h!]
     \centering
-    \includegraphics[width=1\textwidth]{figures/combined_batplots.pdf}
+    \includegraphics[width=1\textwidth]{figures/combined_barplots.pdf}
     \caption{
@@
 For the AUC metrics, a higher score indicated a better preforming classifier.
-        \neri{Organize titles and font sizes}
         }
@@
 
-We convert our model to a binary classifier posed with the question: will there be a \textit{large} earthquake in the query space-time coordinates or not?
-This can easily be done by setting a threshold which defines a "\textit{large}" earthquake, $m_c$ ($=4$ in the examples presented here), then setting examples above that threshold as positive examples and below as negative.
-The probability of a large event occurring in the query coordinates is given by integrating the output PDF above the threshold, $p=\int_{m_c}^{\infty}p(m')dm'$.
-Given such binary classifier, it is possible to examine the common binary classifier metrics of area under receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010} which suit better an imbalanced data set such as ours.
-\ref{fig:metrics}d-f and g-i show the ROC and interpolated PR curves respectively, for all three regions in examined.
-The area under the curve (AUC) values are written in the inset of each panel, color coded by the same legend as the bar plots above.
-It can be seen that throughout all regions MAGNET preforms significantly better than all other benchmarks, indicated by the largest AUC value.
-
 \begin{figure}[h!]
     \centering
-    \includegraphics[width=1\textwidth]{figures/info_over_time_hauksson.pdf}
+    % \includegraphics[width=1\textwidth]{figures/info_over_time_hauksson.pdf}
     \caption{
@@
 Notably, for the Southern California data set, the most rapid information gain occurs during the aftershocks of the two Ridgecrest earthquakes.
-Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4, \ \Delta LL$, respectively, albeit these events are typically considered completely unpredictable.
+Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4, \ \Delta LL$, respectively, although these events are typically considered completely unpredictable.
 This observation leads to a significant corollary: at least some portion of the information gain likely originates from a statistical signal embedded in the event sequence prior to event's occurrence.
@@
 beta for a specified location $x_i$ calculated by smoothing the binned values over a radius of 30km around $x_i$.
-Estimation which consider less than 100 events are discarded.
+Estimation which consider less than 150 events are discarded.
 
@@
     \item Detailed architecture? - \textcolor{red}{\textbf{Make a nice version}}
-    \item Colorful PDF examples for NZ and Japan
-    \item Marginal distributions for NZ and Japan
+    \item Colorful PDF examples for NZ and Japan - \textcolor{blue}{Done}
+    \item Marginal distributions for NZ and Japan - \textcolor{blue}{Done}
     \item Colored scatter magnitude vs index, NZ and Japan - \textcolor{blue}{\textbf{Done}}
-    \item $m_c(t)$ for all three regions -  - \textcolor{blue}{Done}
-    \item Bar plots of conditioned $<LL>$
-    \item Spatial $m_c(x, y)$ and low $m_c(x, y) \leq m_c^{train}$ bins
+    \item $m_c(t)$ for all three regions - \textcolor{blue}{Done}
+    \item Bar plots of conditioned $<LL>$ - \textcolor{blue}{Done}
+    \item Spatial $m_c(x, y)$ and low $m_c(x, y) \leq m_c^{train}$ bins - \textcolor{blue}{Done}
 \end{enumerate}
@@
 \textit{Finally, this note is here in the ED section, but in your paper, the ED section should not have any text except figure and table captions/legends.}
+
+\newpage
+\begin{table}[]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/mean_ll_table.pdf}
+    \caption{Mean log likelihood, $\mathcal{L}$, for additional tested benchmarks.}
+    \label{tab:mean_ll_all_benchmarks}
+\end{table}[h!]
+
+\newpage
+\begin{table}[]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/mean_ll_cond_table.pdf}
+    \caption{Mean log likelihood, $\mathcal{L}$, conditioned on temporal incompletness, for additional tested benchmarks.}
+    \label{tab:mean_ll_all_benchmarks_conditioned}
+\end{table}[h!]
 
@@
     \centering
-        \includegraphics[width=1\textwidth]{figures/temporal_incompleteness.pdf}
+        % \includegraphics[width=1\textwidth]{figures/temporal_incompleteness.pdf}
     \caption{
@@
 
+\newpage
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/raw_results_em.pdf}
+    \caption{
+        write the caption
+    }
+    \label{fig:model_output_em}
+\end{figure}
+
+\newpage
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/likelihood_scatter.pdf}
+    \caption{
+        write the caption
+    }
+    \label{fig:labels_likelihood}
+\end{figure}
+
+\newpage
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/combined_barplots_conditioned.pdf}
+    \caption{
+        \textbf{mean ll mc conditioned}
+    }
+    \label{fig:mean_ll_mc_conditioned}
+\end{figure}
+
+\newpage
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/completeness_cond.pdf}
+    \caption{
+        \textbf{mean ll conditioned on spatial mc}
+    }
+    \label{fig:mean_ll_conditioned_spatial_mc}
+\end{figure}
+
 \end{document}

### 2024-06-02 article.tex
Replaced heavy figures with lighter versions.

@@
 MAGNET is composed of two main components.
-First, the catalog up to time $t$ (not including) is encoded by a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history \yohai{cite LSTM paper}.
+First, the catalog up to time $t$ (not including) is encoded by a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history \neri{cite Hochreiter, S.
+and Schmidhuber, J.
+ürgen.
+Long short-term memory.
+Neural Comput.
+9, 1735–1780 (1997).}.
 Second, this representation is combined with a space-time query specifying the coordinates, $\bf{x}_i$, $t_i$, of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, $p_{\textbf{x}_i, t_i}(m)$, see Fig.
@@
 Importantly, during training we only supply the model with queries about spacetime coordinates where earthquakes indeed occurred, and only events above the completeness magnitude of the catalog are used as queries.
-A detailed description of the model's architecture is given in the methods section \neri{TODO}.
+A detailed description of the model's architecture is given in the methods section.
 
@@
 \section{Results} \label{sec:results}
-As explained above, the model outputs a PDF $p_{\textbf{x}_i, t_i}$ for each earthquake in the training set.
+As explained above, the model outputs a PDF $p_{\textbf{x}_i, t_i}(m_i)$ for each earthquake in the training set.
 We denote $\ell_i=\log\left(p_{\textbf{x}_i, t_i}(m_i)\right)$, the log-likelihood of the true magnitude $m_i$ as predicted by MAGNET.
-The model is trained to maximize the average log-likelihood of the true magnitudes in the training set, $\mathcal{L} = -\langle \log{ \ell_i } \rangle$, where $\langle \cdot\rangle$ stands for the empirical average over all earthquakes in the data set.
+The values of $\ell_i$ for MAGNET and the leading benchmark are presented in Extended Material Fig.
+\ref{fig:labels_likelihood}.
+
+The model is trained to maximize the average log-likelihood of the true magnitudes in the training set, $\mathcal{L} = -\langle \ell_i \rangle$, where $\langle \cdot\rangle$ stands for the empirical average over all earthquakes in the data set.
+\neri{I think the two equations in this paragraph should be numbered.}
 
@@
 Consequently, the raw output for our analysis consists of the model-generated PDFs of possible magnitudes for each event's time and location.
-
+\neri{I liked this sentence: \textit{Here we demonstrate that MAGNET outperforms all benchmarks in predicting earthquake magnitude, demonstrating that magnitudes are not statistically independent on seismic history.} Don't you think it's good to start with this clam to focus the reader on the objective?}\yohai{yes I do, which is why this is the last sentence in the previous paragraph}
 Figure \ref{fig:model_output}a shows the predicted PDFs for 100 randomly sampled events from the Southern California test set.
@@
 \ref{fig:metrics}a-c.
-Further comparison to other benchmarks is given in the supplementary material \yohai{TODO: add the SM}.
+Further comparison to other benchmarks is given in Extended Data Table \ref{tab:mean_ll_all_benchmarks}.
 
@@
 This can be easily calculated from the prediction, as the probability that $m_i>m_t$ is simply $\int_{m_t}^{\infty}p(m')dm'$.
-This task is more interpretable (will there be a large earthquake?) and can be evaluated using standard binary classification metrics such as area under receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010} which suit better an imbalanced data set such as ours.
+This task, asking whether a given event will be a large earthquake, is more interpretable and can be evaluated using standard binary classification metrics such as area under receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010} which suit better an imbalanced data set such as ours.
 Fig.~\ref{fig:metrics}d-f and g-i show these metrics for all three regions in examined and compare against the GR-variant models.
@@
 
-To further assess performance in terms of predicted magnitudes, Figure \ref{fig:model_output}c presents the marginal probability density function (PDF) of magnitudes for MAGNET (red) compared to common benchmarks on the Southern California dataset.
-This figure reveals the general trends predicted by each model and facilitates comparison with the commonly used, stationary GR distribution.
-Furthermore, Fig.
-\ref{fig:model_output}c includes histograms of the train and test set magnitudes alongside the marginal PDFs.
-We observe that the GR distribution closely follows the train set distribution, as expected for a stationary model.
-In contrast, MAGNET aligns well with the test set distribution, suggesting its ability to generalize beyond the training data.
-Notably, dynamic GR benchmarks, exemplified by the "last 300 events" approach,  while exhibiting a better fit to the test set than the stationary GR, lack the detailed features captured by the model's PDF.
+Finally, Figure~\ref{fig:model_output}b presents the marginal probability density function (PDF) of magnitudes for MAGNET's prediction, $P(m)=N^{-1}\sum_i p_{\textbf{x}_i, t_i}(m)$.
+This represents the average probability distribution of magnitudes over the whole test set.
+It is seen that although the model was not constrained to do so, the average distribution aligns well with the ``uninformed'' GR distribution.
+That is, although MAGNET's prediction for each individual earthquake may deviate significantly from the GR distribution, it does reproduce this well known long-time behavior on average.
+It is also seen that the average prediction of the 300-day moving window GR deviates significantly from thte GR distribution.
+We also observe that MAGNET's average prediction aligns well with the \emph{test set} GR distribution, rather than the training set, suggesting its ability to generalize beyond the training data.
+This observation is noticeable even more so in the New Zealand and Japan data sets presented in Extended Data Fig.
+\ref{fig:model_output_em}.
 
@@
     \caption{
-        \textbf{Benchmarking MAGNET}.
-\textbf{a}, PDFs produced by MAGNET for 100 randomly sampled events from the test set with train set's Gutenberg-Richter distribution (dashed black line).
+        \textbf{MAGNET's output}.
+\textbf{a}, PDFs produced by MAGNET for 100 randomly sampled events from the test set, train set's Gutenberg-Richter distribution superimposed (dashed black line).
 The true magnitude label, i.e.
 the true magntude of the event in query, is indicated by the color, interpreted by the colorbar at top.
-\textbf{b}, Likelihoods for returning the magnitude label by MAGNET (red), the 300 events moving window (orange) and the stationary GR (blue dashed line).
-\textbf{c}, Marginal magnitude distribution, i.e.
+\textbf{b}, Marginal magnitude distribution, i.e.
 stationary $p(m)$, produced by MAGNET and common benchmarks for the entire test set.
 Histograms of the train (and test) set are presented in orange (blue).
-\neri{c: Is there a more convenient way of showing the two superimposed distributions?}
     }
@@
     \caption{
-        \textbf{Metric scores for the MAGNET model, and common benchmark models.
+        \textbf{Benchmarking MAGNET.
 a}, \textbf{b}, \textbf{c}, Minus mean information content of our MAGNET model (red) and other common benchmark magnitude predictors (see labels on bars in the figure).
-A lower score indicated a better preforming model.
+A lower score indicates a better preforming model.
 \textbf{d, e, f}, Receiver Operating Characteristic (ROC) and \textbf{g, h, i} the interpolated precision-recall (PR) curve for a binary classifier determining the existence of a large ($m>=4$) event.
 The performance of such a classifier can be quantified by the area under the curve (AUC), noted in each frame, color coded identically to the bar plots.
-For the AUC metrics, a higher score indicated a better preforming classifier.
+For the AUC metrics, a higher score indicates a better preforming classifier.
         }
@@
     \centering
-    % \includegraphics[width=1\textwidth]{figures/info_over_time_hauksson.pdf}
+    \includegraphics[width=1\textwidth]{figures/info_over_time_hauksson.pdf}
     \caption{
-        \textbf{Analysis of information gain of MAGNET over the benchmark.
+        \textbf{Information gain in MAGNET.
 a,} Information gain of individual events in the Southern California test set.
-Scattered dots indicate magnitude at event index, colored by the information gain per event (colorbar indicated scale), here warm is an advantage for MAGNET, while cool is an advantage for the benchmark.
+Scattered dots indicate magnitude at event index, colored by the information gain per event (scale indicated by colorbar), here warm is an advantage for MAGNET, while cool is an advantage for the benchmark.
 Secondary horizontal axis (grey) indicates corresponds origin time.
@@
 
-The information gain achieved by our model for each event can be quantified by the difference in log-likelihood (LL) between our model's prediction and the baseline GR predictor:
+It is interesting to ask The information gain achieved by our model for each event can be quantified by the difference in log-likelihood (LL) between our model's prediction and the baseline GR predictor:
 \begin{equation}
@@
 
+\section{Assessment of Measurement Reliability}
+\neri{YOHAI: I've added this title above. thoughts?}\newline
 While the information gain observed in preshock and mainshocks presented above is encouraging, we address a potential concern regarding information gain during high temporal incompleteness periods\cite{stockman_forecasting_2023}.
@@
 
-For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method (Weimer and Weiss, 2000) within a window of 150 past and 150 future events.
-The resulting temporal incompleteness curves are presented in the SM \neri{TODO: add the $m_c(t)$ figure to SM}.
+For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method \cite{wiemer_minimum_2000} within a window of 150 past and 150 current and future events.
+The resulting temporal incompleteness curves are presented in the Extended Material Fig.
+\ref{fig:temp_incompleteness}.
 After applying the renormalization of Eq.
@@
 Evidently, all conditioned cumulative information gain curves show a steadily increasing trend, supporting the hypothesis that there is information content about the upcoming magnitude in the statistical properties of the catalog, excluding the temporal incompleteness artifact.
-The mean LL scores of the test set for each data catalog are presented in the supplementary material.
-\neri{TODO: add conditioned bar plots to SM}.
+A bar plot comparison of the conditioned mean LL scores are presented in Extended Material Fig.
+\ref{fig:mean_ll_mc_conditioned}.
+Extended Material Table \ref{tab:mean_ll_all_benchmarks_conditioned} displays the conditioned scores for all benchmarks tested in this study.
+\neri{Perhaps the barplot is unecessary?}
 
@@
 We computed local $m_c(x,y)$ per bin and selected only bins with completeness lower than the entire training set's completeness magnitude, $m_c \left( x,y \right)  \leq m_c^{ \left( train \right) }$.
-We then used these bins to create a minimal dataset and re-ran the model with all previous metrics.
-As in prior experiments, MAGNET outperformed benchmarks on this minimal dataset.
-Details and results of this analysis are provided in the supplementary material.
-\neri{TODO: add these results of minimal data set to the SM}.
+We then used these bins to create a minimal dataset and tested the models performance on it.
+MAGNET holds its advantage in the Southern California and New Zealand datasets.
+However, MAGNET loses when this analysis is preformed on the Japan dataset.
+To investigate further, we choose a another subset of the Japan catalog, defined by a tight polygon encompassing the majority of bins that hold.
+$m_c \left( x,y \right)  \leq m_c^{ \left( train \right) }$ bins.
+This results in mostly the territory of Honshu, Kyushu and Shikoku islands.
+We then train a new model for this subset, which, as in the previous experiments, displays MAGNET's superiority for magnitude prediction.
+We conclude from this procedure that constructing a magnitude prediction model for regions with high variability of $m_c(x,y)$ \textit{is possible}, though it might consist of training different regions separately.
+Details and results of this analysis are provided in Extended Material Fig.
+\ref{fig:mean_ll_conditioned_spatial_mc}.
+
+\section{Discussion}
+\neri{YOHAI LET'S MEET TO WRITE CMON}
 
@@
 \begin{enumerate}
-    \item Table: scores of all metrics - \textcolor{red}{\textbf{should I add this??}}
+    \item Table: scores of all metrics - \textcolor{blue}{Done}
     \item Detailed architecture? - \textcolor{red}{\textbf{Make a nice version}}
@@
 \newpage
-\begin{table}[]
+\begin{table}[h!]
     \centering
     \includegraphics[width=1\textwidth]{figures/mean_ll_table.pdf}
-    \caption{Mean log likelihood, $\mathcal{L}$, for additional tested benchmarks.}
+    \caption{Mean log likelihood score, $\mathcal{L}$, for various tested benchmarks.}
     \label{tab:mean_ll_all_benchmarks}
-\end{table}[h!]
+\end{table}
 
 \newpage
-\begin{table}[]
+\begin{table}[h!]
     \centering
     \includegraphics[width=1\textwidth]{figures/mean_ll_cond_table.pdf}
-    \caption{Mean log likelihood, $\mathcal{L}$, conditioned on temporal incompletness, for additional tested benchmarks.}
+    \caption{Mean log likelihood score, $\mathcal{L}$, as the result is conditioned to be above the temporal incompletness.}
     \label{tab:mean_ll_all_benchmarks_conditioned}
-\end{table}[h!]
+\end{table}
 
@@
 	\caption{
-            \textbf{Information gain over time domain for New Zealand and Japan data sets.} This figure presents results for the data sets not displayed in Fig.
-\ref{fig:info_gain_over_time}a.
+            \textbf{Information gain of MAGNET over the benchmark, New Zealand and Japan data sets.} Scattered dots indicate magnitude at event index, colored by the information gain per event.
 \textbf{a} Entire test set of New Zealand and a selected time span (\textbf{b}) focusing on the 2016 Te Araroa earthquake and the 2016 Kaikōura earthquake.
@@
     \centering
-        % \includegraphics[width=1\textwidth]{figures/temporal_incompleteness.pdf}
+        \includegraphics[width=1\textwidth]{figures/temporal_incompleteness.pdf}
     \caption{

### 2024-06-02 article.tex
Captions and some figure adaptations.

@@
 This includes regions previously studied where no information gain was detected using different methods  \cite{ogata_exploring_2018}.
-Similar domain representations for additional regions can be found in the supplementary material  \neri{TODO: add these colorful plots for NZ and JAPAN to the SM}.
+Similar domain representations for additional regions can be found in the supplementary material.
 Furthermore, the smoothed instantaneous information gain per event, presented in Fig.
@@
 Evidently, all conditioned cumulative information gain curves show a steadily increasing trend, supporting the hypothesis that there is information content about the upcoming magnitude in the statistical properties of the catalog, excluding the temporal incompleteness artifact.
-A bar plot comparison of the conditioned mean LL scores are presented in Extended Material Fig.
-\ref{fig:mean_ll_mc_conditioned}.
 Extended Material Table \ref{tab:mean_ll_all_benchmarks_conditioned} displays the conditioned scores for all benchmarks tested in this study.
-\neri{Perhaps the barplot is unecessary?}
 
@@
 
+\subparagraph{Kernel density estimation (KDE) of last n events} Selecting the last $n$ events (similar to described in \textit{last $n$ events}) we use the computed KDE as a magnitude predictor for the current timestamp.
+The KDE is computed using using the SciPy tool\cite{scipy_2020} with 'scott' estimator bandwidth\cite{scott_2015}.
+
 \subsection{Measurement of temporal incompleteness}
@@
 The code for analysing the data is available in \textcolor{red}{\textbf{**link TBD**}}
+
+Country borders in this work are plotted using data from https://geojson-maps.kyd.au/ .
 % This is a required section. Guidelines for data availability: \url{https://www.nature.com/documents/nr-data-availability-statements-data-citations.pdf}.
@@
 Taroni, M., Zhuang, J. \& Marzocchi, W. High‐Definition Mapping of the Gutenberg–Richter b‐Value and Its Relevance: A Case Study in Italy. Seismological Research Letters 92, 3778–3784 (2021).
+\bibitem{scott_2015}
+Scott, David W. Multivariate density estimation: theory, practice, and visualization. John Wiley \& Sons, 2015.  
+\bibitem{scipy_2020}
+Virtanen, P., Gommers, R., Oliphant, T.E. et al. SciPy 1.0: fundamental algorithms for scientific computing in Python. Nat Methods 17, 261–272 (2020).
 
@@
 \textbf{c} The entire Japan data set and a focused time span \textbf{d} of the 2011 Tōhoku earthquake and some of its aftershocks.
+Low magnitude data points in a, c and d are subsampled to ease on image rendering.
          }
@@
 Data for Japan (c) is displayed only near the 2011 Tōhoku earthquake to reduce data density and enhance clarity of the figure.
+Low magnitude data points in b and c are subsampled to ease on image rendering.
     }
@@
     \caption{
-        write the caption
+        \textbf{MAGNET’s output for Japan and New Zealand data sets}.
+\textbf{a} (\textbf{c}), PDFs produced by MAGNET for 100 randomly sampled events from the test set of the New Zealand (Japan) data set, with the train set’s Gutenberg-Richter distribution superimposed (dashed black line).
+The true magnitude label, i.e.
+the true magntude of the event in query, is indicated by the color, interpreted by the colorbar in a.
+\textbf{b} (\textbf{d}), Marginal magnitude distribution, i.e.
+stationary p(m), for the New Zealand (Japan) data set.
+Histograms of the train and test sets are presented in orange and blue respectively.
     }
@@
     \caption{
-        write the caption
+    \textbf{Log-likelihood of the true magnitude, $ell_i$}.
+The likelihood for the true magnitude label of the test set for each data set, $\ell_i=\log\left(p_{\textbf{x}_i, t_i}(m_i)\right)$, as defined in main text.
+$\ell_i$ values of MAGNET and the last 300 events benchmark are shown in red and orange respectively.
+The blue dashed line indicates the values that the GR benchmark would obtain.
     }
@@
 
-\newpage
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=1\textwidth]{figures/combined_barplots_conditioned.pdf}
-    \caption{
-        \textbf{mean ll mc conditioned}
-    }
-    \label{fig:mean_ll_mc_conditioned}
-\end{figure}
+% \newpage
+% \begin{figure}[h!]
+%     \centering
+%     \includegraphics[width=1\textwidth]{figures/combined_barplots_conditioned.pdf}
+%     \caption{
+%         \textbf{mean ll mc conditioned}
+%     }
+%     \label{fig:mean_ll_mc_conditioned}
+% \end{figure}
 
@@
     \caption{
-        \textbf{mean ll conditioned on spatial mc}
+        \textbf{Spatial effects on MAGNET's performance.} \textbf{a-d} Mean log likelihood scores, $\mathcal{L}$, of events contained in bins which hold $m_c(x,y) \leq m_c$, presented in \textbf{e-h}.
+The data set Japan polygon is a tight polygon surrounding the main cluster of low completeness bins in the Japan data set (detailed in main text).
+\textbf{i-l} display the full data set's $m_c(x,y)$ spatial distribution.
     }

### 2024-06-20 article.tex
Mostly improvements in figures. Changed text from describing old spatial metric to conditioned mc description.

@@
 Earthquake occurrence is notoriously difficult to predict.
-Salient features of their spatiotemporal statistics can be relatively well captured by point-process models \neri{this is mentioning ETAS.
-But it is also not completely true.
-We should remove the mention of PP IMO}.
-However, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts, both due to the lack of information about fault conditions and to the inherent complexity of rupture dynamics.
+While some aspects of their spatiotemporal statistics can be relatively well captured by point-process models, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts.
+This is due both due to the lack of information about fault conditions and to the inherent complexity of rupture dynamics.
 Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter distribution, which describes the marginal distribution over large regions and long times.
@@
 \maketitle
-
-\neri{comment by Neri} \newline
-\yohai{comment by Yohai}
-
 \section{Introduction} \label{sec:introduction}
 Earthquakes are notoriously unpredictable, and forecasting seismicity is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes and the scarcity of near-field data~\cite{bernard_earthquake_1999, geller_earthquakes_1997}.
-Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes, including various marginal distributions \cite{gutenberg_frequency_1944, kagan_seismic_2002}, scaling relations \cite{bak_earthquakes_1989, dascher-cousineau_what_2020, kagan_aftershock_2002, utsu_centenary_1995} and characteristics of both spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020, devries_deep_2018, king_static_1994}.
+Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes, including various marginal distributions~\cite{gutenberg_frequency_1944, kagan_seismic_2002}, scaling relations \cite{bak_earthquakes_1989, dascher-cousineau_what_2020, kagan_aftershock_2002, utsu_centenary_1995} and characteristics of both spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020, devries_deep_2018, king_static_1994}.
 Clearly, these insights can be used to quantitatively inform us about future seismicity based on recent history.
@@
 MAGNET is composed of two main components.
-First, the catalog up to time $t$ (not including) is encoded by a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history \neri{cite Hochreiter, S.
-and Schmidhuber, J.
-ürgen.
-Long short-term memory.
-Neural Comput.
-9, 1735–1780 (1997).}.
+First, the catalog up to time $t$ (not including) is encoded by a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history \cite{hochreiter_long_1997}.
 Second, this representation is combined with a space-time query specifying the coordinates, $\bf{x}_i$, $t_i$, of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, $p_{\textbf{x}_i, t_i}(m)$, see Fig.
@@
 	\centering
-        \includegraphics[width=1\textwidth]{figures/intro_fig.pdf}
-	\caption{\textbf{a} MAGNET Architecture: A depiction of the task and architecture of the MAGnitude Neural EsTimation model (MAGNET).
-For each earthquake, seismic history leading up to its occurrence time ($t_i$) is encoded using a Long Short-Term Memory (LSTM) neural network.
-This encoded history is then concatenated with the timing and location of the earthquake.
-The combined information is fed into a fully-connected neural network (FCNN) that outputs a probability density function (PDF) representing the possible magnitudes for the given location.
+        \includegraphics[width=1\textwidth]{figures/intro_fig.png}
+	\caption{\textbf{a} A depiction of the task and architecture of the MAGnitude Neural EsTimation model (MAGNET).
+Seismic history is continuously encoded using a Long Short-Term Memory (LSTM) neural network.
+For each earthquake, the encoded history up to its occurrence time ($t_i$) is concatenated with the time and location of the earthquake (``spacetime query'').
+The combined information is fed into a fully-connected neural network (FCNN) that outputs a probability density function (PDF) of the predicted magnitude at that time and location.
 \textbf{b} Performance on Major Earthquakes: MAGNET's resulting PDFs for a selection of well-known major earthquakes worldwide.
-The red curve represents the model's predicted PDF, while the dashed black curve represents the Gutenberg-Richter (GR) magnitude-frequency distribution, a common naive benchmark.
-For most earthquakes, the value of the red curve is higher for the actual magnitude (shown by the vertical dashed line), indicating that MAGNET's predictions generally outperform the GR benchmark.
-Notably, the Tōhoku earthquake exhibits a different behavior.
+The red curve represents the model's predicted PDF and the dashed black curve represents the Gutenberg-Richter (GR) magnitude-frequency distribution, a common naive benchmark.
+For most earthquakes shown here, the likelihood of the observed magnitude (red star) is higher for MAGNET than for the GR benchmark (black circle).
+Notably, the Tōhoku aftershock shows a qualitatively different behavior.
     }
@@
 As explained above, the model outputs a PDF $p_{\textbf{x}_i, t_i}(m_i)$ for each earthquake in the training set.
-We denote $\ell_i=\log\left(p_{\textbf{x}_i, t_i}(m_i)\right)$, the log-likelihood of the true magnitude $m_i$ as predicted by MAGNET.
+We denote by $\ell_i$ the log-likelihood of the observed magnitude $m_i$ as predicted by MAGNET,
+\begin{align}
+    \mathcal{L} &= -\langle \ell_i \rangle\ , 
+    &
+    \ell_i&=\log\left(p_{\textbf{x}_i, t_i}(m_i)\right)
+    \label{eq:likelihood}
+\end{align}
+ $\mathcal{L}$ is the average minus log-likelihood, which the model is trained to minimize.
 The values of $\ell_i$ for MAGNET and the leading benchmark are presented in Extended Material Fig.
@@
 
-The model is trained to maximize the average log-likelihood of the true magnitudes in the training set, $\mathcal{L} = -\langle \ell_i \rangle$, where $\langle \cdot\rangle$ stands for the empirical average over all earthquakes in the data set.
-\neri{I think the two equations in this paragraph should be numbered.}
-
-% Focusing on existing events within the catalog due to our model's requirement for labeled data, our analysis utilizes the space-time coordinates of earthquakes as input.
-This necessitates excluding queries about arbitrary spatiotemporal points during training and evaluation, as these processes require knowledge of the true magnitude for each event at its specific location and time.
-Consequently, the raw output for our analysis consists of the model-generated PDFs of possible magnitudes for each event's time and location.
-\neri{I liked this sentence: \textit{Here we demonstrate that MAGNET outperforms all benchmarks in predicting earthquake magnitude, demonstrating that magnitudes are not statistically independent on seismic history.} Don't you think it's good to start with this clam to focus the reader on the objective?}\yohai{yes I do, which is why this is the last sentence in the previous paragraph}
 Figure \ref{fig:model_output}a shows the predicted PDFs for 100 randomly sampled events from the Southern California test set.
@@
 \ref{fig:metrics}a-c.
-Further comparison to other benchmarks is given in Extended Data Table \ref{tab:mean_ll_all_benchmarks}.
+Further comparison to other benchmarks is given in Table \ref{tab:mean_ll_benchmarks_main_text}.
 
@@
         \textbf{MAGNET's output}.
-\textbf{a}, PDFs produced by MAGNET for 100 randomly sampled events from the test set, train set's Gutenberg-Richter distribution superimposed (dashed black line).
-The true magnitude label, i.e.
-the true magntude of the event in query, is indicated by the color, interpreted by the colorbar at top.
-\textbf{b}, Marginal magnitude distribution, i.e.
-stationary $p(m)$, produced by MAGNET and common benchmarks for the entire test set.
+\textbf{a}, Predicted magnitude PDFs produced by MAGNET for 100 randomly sampled events from the test set in the Southern California region.
+Curves are colored according to the magnitude of each earthquake event.
+It is seen that the PDFs for high magnitude events are generally skewed towards higher magnitudes, as expected from a predictive model.
+We superimpose the train set Gutenberg-Richter distribution (dashed black line).
+\textbf{b}, Marginal magnitude distribution produced by MAGNET.
+That is, the average of all PDFs from the test set.
+It is seen that the marginal distribution closely resembles the GR distribution.
+In addition, we plot a similar marginal PDF for common benchmarks for the entire test set.
 Histograms of the train (and test) set are presented in orange (blue).
@@
 
+\begin{table}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/main_text_table.pdf}
+    \caption{Mean score, $\mathcal{L}$, for various tested benchmarks.
+$\mathcal{L}$ is computed by Eq.
+\ref{eq:likelihood}.
+Lower score indicates a better magnitude predictor, best score in column is indicated in bold.
+First 3 columns display scores for the raw calculation of $\mathcal{L}$, middle and right column triplets display the scores for the temporally and spatially conditioned $\mathcal{L}$ scores, respectively.
+    }
+    \label{tab:mean_ll_benchmarks_main_text}
+\end{table}
+
 \begin{figure}[h!]
@@
         \textbf{Information gain in MAGNET.
-a,} Information gain of individual events in the Southern California test set.
+a}, Information gain of individual events in the Southern California test set.
 Scattered dots indicate magnitude at event index, colored by the information gain per event (scale indicated by colorbar), here warm is an advantage for MAGNET, while cool is an advantage for the benchmark.
 Secondary horizontal axis (grey) indicates corresponds origin time.
-Cumulative Information gain (dashed black) and cumulative conditioned information gain (dashed grey) curves are superimposed on the scatter, demonstrating a constantly increasing function.
-\textbf{b-d}, Information gain (dashed black) and cumulative conditioned information gain (dashed grey) curves for Southern California, New Zealand and Japan test sets, respectively, all three regions examined in this study.\textbf{e-g}, Smoothed instantaneous information gain.
-Grey cloud shows the standard deviation of the smoothing window.
-All three curves are predominantly above the zero information gain line (dashed red).
-\textbf{h-i} Spatial distribution of information gain for all regions in study.
-Maps show the difference in mean information content per bin, model's advantage is indicated by warm colors, disadvantage by cool colors.
+Cumulative Information gain (CIG) (dashed blue), temporally conditioned CIG (dashed yellow) and spatially conditioned CIG (dashed red) curves are superimposed on the scatter, demonstrating a constantly increasing function.
+\textbf{b-d}, Information gain (dashed blue) temporally conditioned CIG (dashed yellow) and spatially conditioned CIG (dashed red) curves for Southern California, New Zealand and Japan test sets, respectively, all three regions examined in this study.
+An exception from the increasing trend can be observed for the spatially conditioned CIG of New Zealand (red curve in c).
+\textbf{e-g} Spatial distribution of spatially conditioned information gain for all regions in study.
+Maps show the difference in mean information content per bin, model's advantage is indicated by warm colors, disadvantage by cool colors, as defined by the colorbar adjacent to g.
     }
@@
 
-It is interesting to ask The information gain achieved by our model for each event can be quantified by the difference in log-likelihood (LL) between our model's prediction and the baseline GR predictor:
+It is interesting to examine MAGNET's information gain per event, which is quantified by the difference in log-likelihood (LL) of the observed magnitude, 
 \begin{equation}
-    \Delta LL_i = \log{p_{\pmb{x_i}, t_i}^{(MAGNET)}(m_i)} - \log{p^{(GR)}(m_i)}
+    \Delta \ell_i = \log{p_{\pmb{x_i}, t_i}^{(MAGNET)}(m_i)} - \log{p^{(GR)}(m_i)}
     \label{eq:information_gain}
 \end{equation}
-This equation measures the amount of information gained by using our MAGNET model rather than the GR benchmark, for a specific event.
+%This equation measures the amount of information gained by using our MAGNET model rather than the GR benchmark, for a specific event. 
 Analyzing this quantity per event across the entire test set reveals patterns indicating where and when our model achieves its advantage.
@@
 This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as the black dashed line in the figure.
-The predominantly increasing trend in cumulative information gain suggests that our model is consistently advantageous as opposed to a concentrated advantage in specific periods.
+The predominantly increasing trend in cumulative information gain suggests that our model is consistently advantageous as opposed to identifying some local trends such as immediate aftershock sequences.
 Notably, for the Southern California data set, the most rapid information gain occurs during the aftershocks of the two Ridgecrest earthquakes.
-Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4, \ \Delta LL$, respectively, although these events are typically considered completely unpredictable.
-This observation leads to a significant corollary: at least some portion of the information gain likely originates from a statistical signal embedded in the event sequence prior to event's occurrence.
-This point will be further elaborated upon in the discussion section.
-\newline
+Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with $\Delta \ell$ values of $0.5$ and $1.4$ nats (log(2) nat = 1 bit), respectively.
 
@@
 \ref{fig:info_gain_over_time}b-d).
-This includes regions previously studied where no information gain was detected using different methods  \cite{ogata_exploring_2018}.
-Similar domain representations for additional regions can be found in the supplementary material.
-Furthermore, the smoothed instantaneous information gain per event, presented in Fig.
-\ref{fig:info_gain_over_time}e-g, consistently exceeds the threshold for zero information gain marked by the red dashed line.
+Specifically, we show an average information gain of $>0.1$ nats per earthquake in the same region in Japan where Ogata et.~al.~\cite{ogata_exploring_2018} found no information gain.
+Similar plots for the other studied regions are shown in Extended Material Fig.~\ref{fig:nz_japan_info_gain}.
 
-\section{Assessment of Measurement Reliability}
-\neri{YOHAI: I've added this title above. thoughts?}\newline
-While the information gain observed in preshock and mainshocks presented above is encouraging, we address a potential concern regarding information gain during high temporal incompleteness periods\cite{stockman_forecasting_2023}.
-In order to account for the temporal incompleteness artifact, and factor it out of our information gain calculation, we recalculate the likelihood score of each event, $p(m_i)$ , conditioned on the magnitude exceeding a threshold, $\Tilde{m}$:
+\section{Interpretation and caveats}
+At this point, it is clear that MAGNET is able to extract information about the magnitude of specific earthquakes, prior to their occurrence.
+However, such an information gain may potentially result not from the fact that the underlying physical process is predictable (at least to some extent), but from various measurement and modeling artifacts.
+
+An obvious such artifact is ``short-time incompleteness'' (STI): the fact that immediately following a large event the signal of subsequent small events might be buried in the coda of the mainshock~\cite{kagan_short-term_2004, stockman_forecasting_2023}.
+Thus, immediately after a large event it is less likely that a small event will be included in the catalog, not because it did not occur, but rather because it was not detected.
+Indeed, our model consistently allocates less probability mass to low magnitudes shortly after large events (see the Tohoku aftershock in Fig.~\ref{fig:intro_fig})b, and some of the information gain may be attributed to this effect.
+
+In order to factor the effect of STI from the information gain calculation, we recalculate the likelihood score of each event, $p(m_i)$, conditioned on the magnitude exceeding a threshold, $\Tilde{m}$, which is the time-dependent completeness magnitude:
 
 \begin{equation}
-    p \left( m \vert m_i > \Tilde{m} \right) = \frac{p_{\pmb{x}_i, t_i}(m)} {\int_{\Tilde{m}}^{\infty} p(m') ,dm'}
+    p \left( m \vert m_i > \Tilde{m} \right) = \frac{p_{\pmb{x}_i, t_i}(m)} {\int_{\Tilde{m}}^{\infty} p(m') dm'}
     \label{eq:conditioned_likelihood}
@@
 
-For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method \cite{wiemer_minimum_2000} within a window of 150 past and 150 current and future events.
-The resulting temporal incompleteness curves are presented in the Extended Material Fig.
+The instantaneous completeness magnitude, $m_c(t)$, is defined as the minimal magnitude above which all earhtquakes can be detected.
+If we choose $\tilde m \geq m_c$ then, by definition, the model cannot present an information gain due to the STI effect.
+
+We thus take $\Tilde{m}=m_c(t)$, which we estimate dynamically using the maximum curvature method \cite{wiemer_minimum_2000} within a window of 150 past and 150 current and future events.
+The resulting $m_c(t)$ curves are presented in the Extended Material Fig.
 \ref{fig:temp_incompleteness}.
-After applying the renormalization of Eq.
-\ref{eq:conditioned_likelihood} the cumulative information gain is recalculated.
-The result is presented as the grey dashed curved in Fig.
-\ref{fig:info_gain_over_time}a, as well as in Fig.
-\ref{fig:info_gain_over_time}b-d for all three tested regions.
-Evidently, all conditioned cumulative information gain curves show a steadily increasing trend, supporting the hypothesis that there is information content about the upcoming magnitude in the statistical properties of the catalog, excluding the temporal incompleteness artifact.
-Extended Material Table \ref{tab:mean_ll_all_benchmarks_conditioned} displays the conditioned scores for all benchmarks tested in this study.
+The conditioned cumulative information gain (CIG) is then recalculated and presented as the yellow dashed curved in Fig.
+\ref{fig:info_gain_over_time} for all three tested regions.
+Indeed, factoring out the temporal incompleteness reduces the information gain, though it is still significant: evidently, the conditioned CIG curves show a steadily increasing trend.
+Table \ref{tab:mean_ll_benchmarks_main_text} displays the conditioned scores for all benchmarks tested in this study under the columns title \textit{$\mathcal{L}$ temporally conditioned}.
 
-Beyond the temporal domain, MAGNET also exhibits an advantageous information gain distribution spatially (Fig.
-\ref{fig:info_gain_over_time}h-j).
-Each spatial bin's color represents the average log-likelihood difference between MAGNET's predictions and the common GR benchmark (as defined in Eq.
-\ref{eq:information_gain}, averaged and applied per bin).
-Warm colors indicate regions where MAGNET performs better.
-While the advantage in Southern California appears spatially diffuse, both New Zealand and Japan display concentrations of high advantage at the dataset edges, further from the mainland.
-We hypothesize this is linked to the high local magnitude of completeness, $m_c(x,y)$, in these regions.
-To address potential artifacts from spatially varying $m_c(x,y)$, we conducted further analysis.
-We computed local $m_c(x,y)$ per bin and selected only bins with completeness lower than the entire training set's completeness magnitude, $m_c \left( x,y \right)  \leq m_c^{ \left( train \right) }$.
-We then used these bins to create a minimal dataset and tested the models performance on it.
-MAGNET holds its advantage in the Southern California and New Zealand datasets.
-However, MAGNET loses when this analysis is preformed on the Japan dataset.
-To investigate further, we choose a another subset of the Japan catalog, defined by a tight polygon encompassing the majority of bins that hold.
-$m_c \left( x,y \right)  \leq m_c^{ \left( train \right) }$ bins.
-This results in mostly the territory of Honshu, Kyushu and Shikoku islands.
-We then train a new model for this subset, which, as in the previous experiments, displays MAGNET's superiority for magnitude prediction.
-We conclude from this procedure that constructing a magnitude prediction model for regions with high variability of $m_c(x,y)$ \textit{is possible}, though it might consist of training different regions separately.
-Details and results of this analysis are provided in Extended Material Fig.
-\ref{fig:mean_ll_conditioned_spatial_mc}.
+Similarly to the temporal domain, a spurious information gain may also be caused by temporally constant but spatially varying properties.
+For example, fault stresses, geometry and lithology may change in space resulting in spatially-dependent GR distribution \cite{amitrano_brittle-ductile_2003, scholz_stress_2015, herrmann_revealing_2022, taroni_earthquake_2023}, and the local completeness magnitudes is also generally spatially dependent.
+In such a case, the information gain of MAGNET does not imply that the magnitudes are correlated in time.
+To examine this, we preform two additional tests.
+First, we construct a spatially dependent GR distributions where $m_c$ is held constant ($=m_c^(train\ set)$).
+MAGNET outperforms the spatially varying GR benchmark for all regions in the unconditioned $\mathcal{L}$ score, as can be seen in EM Table \ref{tab:mean_ll_all_benchmarks}.
+The spatially varying GR shows an advantage over MAGNET for solely the New Zealand data set for the temporally conditioned $\mathcal{L}$.
+This advantage may arise from a the measurement artifact of spatial variation of $m_c$ originating the stricture of the seismic network.
+
+In order to acount for such spatially dependent artifacts we preform a second calculation.
+We once again condition MAGNET's results on a threshold according to \ref{eq:conditioned_likelihood}, this time choosing $\Tilde{m}=m_c(x,y)$, the spatial incompleteness.
+We calculate $m_c(x,y)$ for a grid of coordinates spaced every $0.1^\circ$.
+The calculation is performed for all events within a square centered at each grid point, with a side length of $0.2^\circ$.
+The spatially conditioned CIG curves are shown as the red dashed curve in Fig.
+\ref{fig:info_gain_over_time} for all three regions.
+Evidently, it is observed to be a steadily increasing function for the Southern California and Japan data sets.
+In the New Zealand data set the curve shows an increase, then plateaus after the first sequence of major events (the 2016 Te Araroa and Kaikōura earthquakes).
+This may indicate that a good portion of this information gain followed by these earthquakes originate in the spatial incompleteness artifact.
+Nevertheless, the overall score shows an advantage for our model for the test set.
+Fig.
+\ref{fig:info_gain_over_time}e-g show the spatial distribution of our MAGNET's advantage over the GR benchmark.
+All three regions exhibit a spatially diffuse pattern.
+This suggests that information is not primarily sourced from any one specific area.
 
@@
 \subsection{Neural architecture}
-A detailed visual of the model's architecture is presented in Extended Data Fig. \ref{fig:architecture}. The examined catalog is used as an input into three distinct encoding components:
-\subparagraph{\textbf{Recent Earthquakes Encoder}} Calculates a set of predefined functions on various past time windows.
-This encompasses a representation of various statistical properties of the catalog.
-The result is then used as an input to a LSTM NN.
-A detailed description of this encoder can be found in past study \cite{zlydenko_neural_2023}.
+A detailed visual of the model's architecture is presented in Extended Data Fig. \ref{fig:architecture}. The catalog is used as an input into three distinct encoding components:
+\subparagraph{\textbf{Recent Earthquakes Encoder}} Calculates a set of predefined (non trainable) functions on various past time windows.
+The result is then used as input to a LSTM NN.
+A detailed description of this encoder is given in Zlydenko et.~al.~\cite{zlydenko_neural_2023}.
 
 \subparagraph{\textbf{Seismicity Rate Encoder}} Computes a rough estimation of the amount of energy released in a given region around the query coordinate, in a given past time window.
-The result is then used as an input to a LSTM NN.
-A detailed description of this encoder can be found in past study \cite{zlydenko_neural_2023}.
+The result is then input to a LSTM NN.
+A detailed description of this encoder is found in Zlydenko et.~al.~\cite{zlydenko_neural_2023}.
 
-\subparagraph{\textbf{Event Time Encoder}} Prepares the query time-space coordinates.
-Time is represented as the time difference from the most recent past event, in seconds.
-No further processing is done in this encoder.
+\subparagraph{\textbf{Spacetime Query Encoder}} Time is represented as the time difference from the most recent past event, in seconds. No further processing is done in this encoder.
 
-Outputs of all encoders are concatenated and used as the input to a Fully Connected Neural Network (FCNN).
-The sizes of layers in the FCNN, as the number of layers themselves, are used as tuneable hyper parameters in order to optimize the architecture by cross validation.
-A $tanh$ function is used as the activation on the output of every hidden layer, and a $softplus$ as an activation on the output layer.
+Outputs of all encoders are concatenated and fed to a Fully Connected Neural Network (FCNN).
+The sizes of layers in the FCNN, as the number of layers themselves, are are optimized by cross validation.
+A $tanh$ activation is used, and a $\texttt{softplus}$ as an activation on the output layer.
 
 \subsection{Loss metric}
-To obtain the Probability Density Function of the magnitudes ($p(m)$) for each query coordinates ($\textbf{x}_i, t_i$) we maximize the likelihood presented in Eq.
-\ref{eq:loss_function}.
-For our usage we choose $p(\cdot)$ to be a mixture of two stretched and shifted Kumaraswamy distributions \cite{kumaraswamy_generalized_1980}:
+To obtain the Probability Density Function (PDF) of the magnitudes ($p(m)$) for each query coordinates ($\textbf{x}_i, t_i$) we maximize the likelihood, $\mathcal{L} = -\langle \ell_i \rangle$, where $\langle \cdot\rangle$ is the mean over all examples in the set.
+The PDF is taken to be a mixture of two stretched and shifted Kumaraswamy distributions~\cite{kumaraswamy_generalized_1980}:
 \begin{equation}
@@
 \subparagraph{Last $n$ events} The Gutenberg- Richter (GR) distribution fitted for the past $n$ events.
-This method follows the method presented in \cite{gulia_real-time_2019} for constructing a b-value time-series.
-We use the train set's completeness magnitude $m_c^{train}$ to define the GR distribution, and a minimal number of events above $m_c$ ($=10$ in this study).
-If the minimal number of events above completeness is not fulfilled the window is expanded into the past until the condition is achieved.
-An attempt to compute $m_c$ per window has been made, though neglected as it showed worse preformance.
+This method follows the method presented in Gulia \& Wiemer\cite{gulia_real-time_2019} for constructing a b-value time-series.
+If in the last $n$ events there are less than 10 earthquakes of magnitudes above $m_c^{train}$, then more events from the past are included, until this condition is met.
+$m_c^{train}$ is estimated using the maximal curvature method\cite{wiemer_minimum_2000}.
 
@@
 
-\subparagraph{Spatially varying Gutenberg Richter} Estimation of the local $\beta$ value is done by following the method of \cite{taroni_highdefinition_2021}.
-Seismicity data is first binned into $0.1^\circ \times 0.1^\circ$, where the GR distribution is calculated for.
-beta for a specified location $x_i$ calculated by smoothing the binned values over a radius of 30km around $x_i$.
-Estimation which consider less than 150 events are discarded.
+\subparagraph{Spatially varying GR} Estimation of the local $b$ value is done by following the method of Taroni et.~al.~\cite{taroni_highdefinition_2021}. Seismicity data is 
+binned into $0.1^\circ \times 0.1^\circ$ bins, and the GR distribution is fitted for each bin.
+$b$ for a specified location $x_i$ is calculated by the average of all $b$ values in bins within 30km around $x_i$.
+Estimations which consider less than 150 events are discarded.
 
-\subparagraph{Kernel density estimation (KDE) of last n events} Selecting the last $n$ events (similar to described in \textit{last $n$ events}) we use the computed KDE as a magnitude predictor for the current timestamp.
-The KDE is computed using using the SciPy tool\cite{scipy_2020} with 'scott' estimator bandwidth\cite{scott_2015}.
+\subparagraph{Kernel density estimation (KDE) of last n events} Selecting the last $n$ events (similar to described in \textit{last $n$ events}) we copmute the KDE estimation of the magnitude PDF, using the 'scott' estimator bandwidth\cite{scott_2015}.
+This is implemented using the SciPy tool\cite{scipy_2020}.
+The resulting PDF is used as the prediction of the magnitude.
 
-\subsection{Measurement of temporal incompleteness}
-Temporal incompleteness is measured at the times of events by fitting the completeness magnitude to a window of 300 events: 150 past events, 1 current event, 149 future events.
-The window is constructed by the same algorithm, used to construct the \textit{Last $n$ events benchmark}.
+\subsection{Measurement of incompleteness}
+\subparagraph{Temporal incompleteness} Temporal incompleteness is measured at the times of events by fitting the completeness magnitude to a window of 300 events: 150 past events, 1 current event, 149 future events.
+The window is constructed by the same algorithm as in the \textit{Last $n$ events benchmark}.
+
+\subparagraph{Spatial incompleteness} Temporal incompleteness is calculated for a grid of coordinates, spaced by $\theta^\circ$ degrees, covering the relevant region.
+$\theta^\circ$ is set to 0.1 for Southern California and 0.5 for New Zealand and Japan data sets.
+Calculation is done using train set's events.
+The value per point is calculated using the maximal curvature method\cite{wiemer_minimum_2000}, over the nearest, at least 100 events, with a minimal radius of $\theta^\circ$.
+In cases where the the nearest 100 events span over a radius of $2^\circ$ the data point is considered invalid.
+Events from the test set are divided into bins of $\theta^\circ$ centered around the point for which $m_c(x,y)$ was computed.
+Events are assigned with the $m_c$ value of their bin.
 
@@
 \section*{Acknowledgements}
-\yohai{Legit to use your comment to give you tasks? TODO by \textbf{Yohai}}
-Acknowledgements should be brief, and should not include thanks to anonymous referees and editors, inessential words, or effusive comments.
-A person can be thanked for assistance, not “excellent” assistance, or for comments, not “insightful” comments, for example.
-Acknowledgements can contain grant and contribution numbers.
+We thank Yehuda Ben-Zion, Assaf Inbal, Kelian Dascher-Cousineau, David Marsan and Eugenio Lippiello \neri{Add Zhuang and Ogata?
+I did discuss this with them...} for thoughtful discussions.
+YBS is supported by ISF grant 1907/22 and by Google Gift grant.
 
 \section*{Author Contributions}
-N.B., O.Z., O.G., Y.B.S.
 N.B., O.Z., O.G.
@@
 \section*{Author Information}
-\yohai{Legit to use your comment to give you tasks? TODO by \textbf{Yohai} x2}
-\textbf{Grey's Note} \textit{Two things are required in this Author Information section: (1) A statement about competing interests.
-I have no advice about what constitutes a competing interest, you will have to read about it and make your own decision.
-(2) A clear statement about who to contact with question about the paper.
-An example is below.}
-% For example:
-% The authors declare no competing interests.
-Please contact either the first author (Grey Nearing; \href{mailto:nearing@google.com}{nearing@google.com}) or Avinatan Hassidim; \href{mailto:avinatan@google.com}{avinatan@google.com}) for correspondence and requests, including questions regarding reprints and permissions.
-
-\newpage
-\section*{Extended Data}
-\begin{enumerate}
-    \item Table: scores of all metrics - \textcolor{blue}{Done}
-    \item Detailed architecture? - \textcolor{red}{\textbf{Make a nice version}}
-    \item Colorful PDF examples for NZ and Japan - \textcolor{blue}{Done}
-    \item Marginal distributions for NZ and Japan - \textcolor{blue}{Done}
-    \item Colored scatter magnitude vs index, NZ and Japan - \textcolor{blue}{\textbf{Done}}
-    \item $m_c(t)$ for all three regions - \textcolor{blue}{Done}
-    \item Bar plots of conditioned $<LL>$ - \textcolor{blue}{Done}
-    \item Spatial $m_c(x, y)$ and low $m_c(x, y) \leq m_c^{train}$ bins - \textcolor{blue}{Done}
-\end{enumerate}
-
-\textbf{Grey's Note} \textit{You get up to 10 items in the Extended Data section.
-All Tables and Figures should be enumerated as "Extended Data Figure 1", "Extended Data Table 1", etc.
-These must also be referenced the same way in the text.
-Extended Data tables and figures may be referenced in the main article and/or in the Methods section.
-I am doing the in-text references to these ED items manually, instead of trying to override the table numbering schemed in Latex.
-Please update this template to automate that process if you feel like doing that.}
-
-\textit{You can use Latex formatting for tables, but the journal will eventually require that your tables be submitted as images. Yes, this also struck me as unusual.}
-
-\textit{Each ED figure should be on a separate page.}
-
-\textit{Finally, this note is here in the ED section, but in your paper, the ED section should not have any text except figure and table captions/legends.}
+The authors declare no competing interests.
+Please contact Yohai Bar-Sinai \href{mailto:ybarsinai@gmail.com}{ybarsinai@gmail.com} or Neri Berman \href{mailto:neriberman@gmail.com}{neriberman@gmail.com} for correspondence and requests, including questions regarding reprints and permissions.
 
@@
     \centering
-    \includegraphics[width=1\textwidth]{figures/mean_ll_table.pdf}
-    \caption{Mean log likelihood score, $\mathcal{L}$, for various tested benchmarks.}
+    \includegraphics[width=1\textwidth]{figures/em_table.pdf}
+    \caption{Mean score, $\mathcal{L}$, for all tested benchmarks.
+$\mathcal{L}$ is computed by Eq.
+\ref{eq:likelihood}.
+Lower score indicates a better magnitude predictor, best score in column is indicated in bold.
+First 3 columns display scores for the raw calculation of $\mathcal{L}$, middle and right column triplets display the scores for the temporally and spatially conditioned $\mathcal{L}$ scores, respectively.
+    }
     \label{tab:mean_ll_all_benchmarks}
-\end{table}
-
-\newpage
-\begin{table}[h!]
-    \centering
-    \includegraphics[width=1\textwidth]{figures/mean_ll_cond_table.pdf}
-    \caption{Mean log likelihood score, $\mathcal{L}$, as the result is conditioned to be above the temporal incompletness.}
-    \label{tab:mean_ll_all_benchmarks_conditioned}
 \end{table}
@@
     \includegraphics[width=1\textwidth]{figures/detailed_architecture.pdf}
-    \caption{\neri{TODO: this needs to be nicer} \textbf{MAGNET architecture.} Described in detail in the methods section.}
+    \caption{\textbf{MAGNET architecture.} Described in detail in the methods section.}
     \label{fig:architecture}
@@
 	\centering
-        \includegraphics[width=1\textwidth]{figures/info_gain_scatter_NZ_and_Japan.pdf}
+        \includegraphics[width=1\textwidth]{figures/info_gain_scatter_NZ_and_Japan.png}
 	\caption{
@@
     \centering
-        \includegraphics[width=1\textwidth]{figures/temporal_incompleteness.pdf}
+        \includegraphics[width=1\textwidth]{figures/temporal_incompleteness.png}
     \caption{
@@
     \centering
-    \includegraphics[width=1\textwidth]{figures/likelihood_scatter.pdf}
+    \includegraphics[width=1\textwidth]{figures/likelihood_scatter.png}
     \caption{
-    \textbf{Log-likelihood of the true magnitude, $ell_i$}.
+    \textbf{Log-likelihood of the true magnitude, $\ell_i$}.
 The likelihood for the true magnitude label of the test set for each data set, $\ell_i=\log\left(p_{\textbf{x}_i, t_i}(m_i)\right)$, as defined in main text.
@@
 The blue dashed line indicates the values that the GR benchmark would obtain.
+Low magnitude data points in b and c are subsampled to ease on image rendering.
     }
@@
 
-% \newpage
-% \begin{figure}[h!]
-%     \centering
-%     \includegraphics[width=1\textwidth]{figures/combined_barplots_conditioned.pdf}
-%     \caption{
-%         \textbf{mean ll mc conditioned}
-%     }
-%     \label{fig:mean_ll_mc_conditioned}
-% \end{figure}
-
-\newpage
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=1\textwidth]{figures/completeness_cond.pdf}
-    \caption{
-        \textbf{Spatial effects on MAGNET's performance.} \textbf{a-d} Mean log likelihood scores, $\mathcal{L}$, of events contained in bins which hold $m_c(x,y) \leq m_c$, presented in \textbf{e-h}.
-The data set Japan polygon is a tight polygon surrounding the main cluster of low completeness bins in the Japan data set (detailed in main text).
-\textbf{i-l} display the full data set's $m_c(x,y)$ spatial distribution.
-    }
-    \label{fig:mean_ll_conditioned_spatial_mc}
-\end{figure}
-
 \end{document}

### 2024-06-24 article.tex
First draft of discussion.

@@
 That is, although MAGNET's prediction for each individual earthquake may deviate significantly from the GR distribution, it does reproduce this well known long-time behavior on average.
-It is also seen that the average prediction of the 300-day moving window GR deviates significantly from thte GR distribution.
+It is also seen that the average prediction of the 300-event moving window GR deviates significantly from the GR distribution.
 We also observe that MAGNET's average prediction aligns well with the \emph{test set} GR distribution, rather than the training set, suggesting its ability to generalize beyond the training data.
@@
     \centering
-    \includegraphics[width=1\textwidth]{figures/main_text_table.pdf}
+    \includegraphics[width=1\textwidth]{figures/score_tables_main.pdf}
     \caption{Mean score, $\mathcal{L}$, for various tested benchmarks.
@@
 
-\section{Discussion}
-\neri{YOHAI LET'S MEET TO WRITE CMON}
+\section{Discussion / Implications of a successful magnitude predictor}
+In this work we have introduced the concept of a \textit{magnitude predictor} to assist us in finding dependencies between magnitudes and the seismic history.
+By focusing solely on magnitude prediction, assuming known location and timing of the event, we have shown the some of the information about an earthquakes magnitude is contained in the seismic history of a simple hypocenter catalog.
+Throughout this work we have made significant effort to address possible origins of this information gain.
+We have ruled out temporal and spatial variability in the completeness magnitude - and artifact originating from the seismic network's limitations.
+Furthermore, we examine a several benchmark models to make sure the advantage of our model, MAGNET, does not originate from a simple, previously attempted, statistical pattern of earthquake occurrence.
+Among the benchmarks we examine are standard spatially and temporally varying Gutenberg Richter distributions, as well as other benchmarks encompassing the near past seismic trends, such as a kernel density estimation of the past $n$ events.
+We have demonstrated that MAGNET's performance exceeds that of every benchmark tested, under any examined conditioning \neri{There is that temporal conditioning of the spatial GR I'm not sure how to treat in this sentence...}.
+It is then implied that MAGNET extracts previously unknown patterns of embodied in the seismic history that hold information of the magnitude of the upcoming earthquake.
+A corollary follows this finding: for some earthquakes, the magnitude might be partially predetermined before the earthquake is initiated.
+This suggests the existence of a precursory process influencing an earthquake's magnitude.
+However, the specific nature and seismic patterns associated with this precursory stage remain unclear based on our current evidence.
+Nonetheless, this finding provides a promising avenue for future research aimed at identifying these precursory signals.
+Investigating the properties of such precursory signals may be carried out by interpretability analysis of neural networks \neri{cite}.
+Such method may reveal what functions of the encoders (see architecture in methods section) have a larger weight in the result of well predicted events.
+Furthermore, the method of ablation \neri{cite} can be applied to extract heuristic, more simplified dependencies between statistical trends in the seismic history to trends in the resulting PDF.
+
+\section{Possible affects on operational earthquake forecasting and prediction}
+Discovery of specific pattern in the seismic history as precursory signals can advance the effort of earthquake prediction a great deal.\neri{cite?
+what?}.
+Nevertheless, discovering such signals in enough accuracy to prove useful for operational systems may be a further down the road.
+An immediate application of the model presented in this work may be of quick magnitude estimation to be used for early warning systems.
+Such a system does not require additional data sources of assumptions on the model as the timing and location of the earthquake to be estimated are given by design of the task.
+A second application may incorporating such a magnitude predictor into an earthquake rate forecasting model such as the epidemic type aftershock sequence model (ETAS) \neri{cite}.
+In ETAS, the rate of events is determined by the product of a set of functions.
+\neri{formalize?} Origin time of events of the forecasted sequence are then sampled from the rate function, and assigned with a magnitude by sampling from a fitted GR distribution.
+y replacing the randomly assigned magnitudes of ETAS with MAGNET or a similar magnitude predictor one will improve the predictive power of rate prediction.
+As we have shown - magnitudes predicted by MAGNET will estimate the true magnitudes in a higher likelihood.
 
@@
     \centering
-    \includegraphics[width=1\textwidth]{figures/em_table.pdf}
+    \includegraphics[width=1\textwidth]{figures/score_tables_EM.pdf}
     \caption{Mean score, $\mathcal{L}$, for all tested benchmarks.

### 2024-06-28 article.tex
Minor changes in discussion. Updated some citations.

@@
 By focusing solely on magnitude prediction, assuming known location and timing of the event, we have shown the some of the information about an earthquakes magnitude is contained in the seismic history of a simple hypocenter catalog.
-Throughout this work we have made significant effort to address possible origins of this information gain.
+Throughout this work we have made significant effort to account for information gain that may originate in artifacts rather than true properties of the earthquake physical system.
 We have ruled out temporal and spatial variability in the completeness magnitude - and artifact originating from the seismic network's limitations.
-Furthermore, we examine a several benchmark models to make sure the advantage of our model, MAGNET, does not originate from a simple, previously attempted, statistical pattern of earthquake occurrence.
+Furthermore, we examine several benchmark models to make sure the advantage of our model, MAGNET, does not originate from a simple, previously attempted, statistical pattern of earthquake occurrence.
 Among the benchmarks we examine are standard spatially and temporally varying Gutenberg Richter distributions, as well as other benchmarks encompassing the near past seismic trends, such as a kernel density estimation of the past $n$ events.
@@
 However, the specific nature and seismic patterns associated with this precursory stage remain unclear based on our current evidence.
-Nonetheless, this finding provides a promising avenue for future research aimed at identifying these precursory signals.
-Investigating the properties of such precursory signals may be carried out by interpretability analysis of neural networks \neri{cite}.
+Such a precursory stage has been proposed in the past, specifically for large earthquakes, by Ben~-~Zion and Zaliapin\cite{ben-zion_localization_2020}.
+We propose the phenomenon we observe in our work may coincide with the work of Ben~-~Zion and Zaliapin\cite{ben-zion_localization_2020} as both examine statistical properties leading to emergence of a specific magnitude in determined space and time coordinates.
+In order to show such competability between to works further investigation is required.
+Specifically, we propose to examine the time a length scale upon which MAGNET's prediction are valid, ahead of origin time.
+This can be done be examining query coordinates around and before the true time-space coordiantes  of the labeled events.
+This can also be done by constructing a different model trained to estimate the magnitude of an upcoming earthquake ahead of the origin time, examining how its accuracy declines with lead time and distance from the event location.
+Additionally, MAGNET may be used in it's current trained state to shed light on potential precursory signals.
+% Nonetheless, this finding provides a promising avenue for future research aimed at identifying these precursory signals. Investigating the properties of such precursory signals may be carried out
+By explainability analysis of neural networks \cite{sturmfels_visualizing_2020, zhang_survey_2021, liu_interpretable_2023} one may conclude on the statistical patterns ~MAGNET has uncovered.
 Such method may reveal what functions of the encoders (see architecture in methods section) have a larger weight in the result of well predicted events.
-Furthermore, the method of ablation \neri{cite} can be applied to extract heuristic, more simplified dependencies between statistical trends in the seismic history to trends in the resulting PDF.
+Furthermore, the method of ablation \cite{meyes_ablation_2019, sturmfels_visualizing_2020} can be applied to extract heuristic, more simplified dependencies between statistical trends in the seismic history to trends in the resulting PDF.
 
 \section{Possible affects on operational earthquake forecasting and prediction}
-Discovery of specific pattern in the seismic history as precursory signals can advance the effort of earthquake prediction a great deal.\neri{cite?
-what?}.
+Discovery of specific pattern in the seismic history as precursory signals can advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}.
 Nevertheless, discovering such signals in enough accuracy to prove useful for operational systems may be a further down the road.
@@
 Such a system does not require additional data sources of assumptions on the model as the timing and location of the earthquake to be estimated are given by design of the task.
-A second application may incorporating such a magnitude predictor into an earthquake rate forecasting model such as the epidemic type aftershock sequence model (ETAS) \neri{cite}.
+A second application may incorporating such a magnitude predictor into an earthquake rate forecasting model such as the epidemic type aftershock sequence model (ETAS) or similar \cite{ogata_statistical_1988, ogata_statistics_2017, dascher-cousineau_using_2023}.
 In ETAS, the rate of events is determined by the product of a set of functions.
 \neri{formalize?} Origin time of events of the forecasted sequence are then sampled from the rate function, and assigned with a magnitude by sampling from a fitted GR distribution.
-y replacing the randomly assigned magnitudes of ETAS with MAGNET or a similar magnitude predictor one will improve the predictive power of rate prediction.
+By replacing the randomly assigned magnitudes of ETAS with MAGNET or a similar magnitude predictor one will improve the predictive power of rate prediction.
+In such a model, rate prediction will not be dominated by the abundant low magnitude events as is the case in ETAS, but rather by the high magnitude events which are of more importance for operational means.
 As we have shown - magnitudes predicted by MAGNET will estimate the true magnitudes in a higher likelihood.
+
+\neri{I bet the reviewer will want us to talk about limitations and improvements to the model but maybe we can wait and see}
 

### 2024-07-02 article.tex
fix typos. Improve metrics and timely information gain figures.

--- 2024-06-20 article.tex
@@
 While some aspects of their spatiotemporal statistics can be relatively well captured by point-process models, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts.
-This is due both due to the lack of information about fault conditions and to the inherent complexity of rupture dynamics.
+This is due both to the lack of information about fault conditions and to the inherent complexity of rupture dynamics.
 Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter distribution, which describes the marginal distribution over large regions and long times.
@@
 MAGNET is composed of two main components.
-First, the catalog up to time $t$ (not including) is encoded by a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history \cite{hochreiter_long_1997}.
+First, the catalog up to time $t$ (not including) is encoded by a long-short term memory (LSTM) units, which produces a latent representation of the seismic history \cite{hochreiter_long_1997}.
 Second, this representation is combined with a space-time query specifying the coordinates, $\bf{x}_i$, $t_i$, of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, $p_{\textbf{x}_i, t_i}(m)$, see Fig.
@@
 
-We deployed MAGNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog \cite{hauksson_waveform_2012} for Southern California, GeoNet \cite{gns_geonet_1970} for New Zealand, and the JMA catalog \cite{noauthor_japan_nodate-1} for Japan.
+We deployed MAGNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog \cite{hauksson_waveform_2012} for Southern California, GeoNet \cite{gns_geonet_1970} for New Zealand, and the JMA catalog \cite{noauthor_japan_nodate} for Japan.
 While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies and exhibit varying data quality.
@@
         \includegraphics[width=1\textwidth]{figures/intro_fig.png}
-	\caption{\textbf{a} A depiction of the task and architecture of the MAGnitude Neural EsTimation model (MAGNET).
-Seismic history is continuously encoded using a Long Short-Term Memory (LSTM) neural network.
+	\caption{\textbf{a} MAGnitude Neural EsTimation model (MAGNET) task and architecture.
+Seismic history is continuously encoded using a Long Short-Term Memory neural network.
 For each earthquake, the encoded history up to its occurrence time ($t_i$) is concatenated with the time and location of the earthquake (``spacetime query'').
-The combined information is fed into a fully-connected neural network (FCNN) that outputs a probability density function (PDF) of the predicted magnitude at that time and location.
+The combined information is fed into a fully-connected neural network that outputs a probability density function (PDF) of the predicted magnitude at that time and location.
 \textbf{b} Performance on Major Earthquakes: MAGNET's resulting PDFs for a selection of well-known major earthquakes worldwide.
@@
 \section{Results} \label{sec:results}
-As explained above, the model outputs a PDF $p_{\textbf{x}_i, t_i}(m_i)$ for each earthquake in the training set.
+As explained above, the model outputs a PDF $p_{\textbf{x}_i, t_i}(m_i)$ for each earthquake in the test set.
 We denote by $\ell_i$ the log-likelihood of the observed magnitude $m_i$ as predicted by MAGNET,
@@
 
-In addition to comparing the log-likelihood of the observed data, one can convert the prediction problem to a binary question: will the earthquake at the spacetime query $\textbf{x}_i, t_i$ have be higher than a threshold $m_t$?
+In addition to comparing the log-likelihood of the observed data, one can convert the prediction problem to a binary question: will the earthquake at the spacetime query $\textbf{x}_i, t_i$ have a higher magnitude than a threshold $m_t$?
 This can be easily calculated from the prediction, as the probability that $m_i>m_t$ is simply $\int_{m_t}^{\infty}p(m')dm'$.
@@
 That is, the average of all PDFs from the test set.
-It is seen that the marginal distribution closely resembles the GR distribution.
+It is seen that the marginal distribution closely resembles the GR distribution, though follows the test set distribution more precisely.
 In addition, we plot a similar marginal PDF for common benchmarks for the entire test set.
@@
 A lower score indicates a better preforming model.
+Scores conditioned above the temporal incompleteness are marked by the horizontal dashed line on each corresponding bar.
 \textbf{d, e, f}, Receiver Operating Characteristic (ROC) and \textbf{g, h, i} the interpolated precision-recall (PR) curve for a binary classifier determining the existence of a large ($m>=4$) event.
@@
 \textbf{e-g} Spatial distribution of spatially conditioned information gain for all regions in study.
-Maps show the difference in mean information content per bin, model's advantage is indicated by warm colors, disadvantage by cool colors, as defined by the colorbar adjacent to g.
+Maps show the difference in mean information content per bin, model's advantage is indicated by warm colors, disadvantage by cool colors, as defined by the colorbar adjacent to f.
     }
@@
 \ref{eq:information_gain}.
-This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as the black dashed line in the figure.
+This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as the blue dashed line in the figure.
 The predominantly increasing trend in cumulative information gain suggests that our model is consistently advantageous as opposed to identifying some local trends such as immediate aftershock sequences.
@@
 Thus, immediately after a large event it is less likely that a small event will be included in the catalog, not because it did not occur, but rather because it was not detected.
-Indeed, our model consistently allocates less probability mass to low magnitudes shortly after large events (see the Tohoku aftershock in Fig.~\ref{fig:intro_fig})b, and some of the information gain may be attributed to this effect.
+Indeed, our model consistently allocates less probability mass to low magnitudes shortly after large events (see the Tohoku aftershock in Fig.~\ref{fig:intro_fig}b), and some of the information gain may be attributed to this effect.
 
@@
 Indeed, factoring out the temporal incompleteness reduces the information gain, though it is still significant: evidently, the conditioned CIG curves show a steadily increasing trend.
-Table \ref{tab:mean_ll_benchmarks_main_text} displays the conditioned scores for all benchmarks tested in this study under the columns title \textit{$\mathcal{L}$ temporally conditioned}.
+the horizontal dashed lines on the bars in Fig \ref{fig:metrics}a-c indicate the mean conditioned scores, also presented in Table \ref{tab:mean_ll_all_benchmarks}.
+EM Table \ref{tab:mean_ll_benchmarks_main_text} displays the conditioned scores for all benchmarks tested in this study.
 
@@
 To examine this, we preform two additional tests.
-First, we construct a spatially dependent GR distributions where $m_c$ is held constant ($=m_c^(train\ set)$).
+First, we construct a spatially dependent GR distributions where $m_c$ is held constant ($=m_c^{\mbox{\footnotesize (train set)}}$).
 MAGNET outperforms the spatially varying GR benchmark for all regions in the unconditioned $\mathcal{L}$ score, as can be seen in EM Table \ref{tab:mean_ll_all_benchmarks}.
 The spatially varying GR shows an advantage over MAGNET for solely the New Zealand data set for the temporally conditioned $\mathcal{L}$.
-This advantage may arise from a the measurement artifact of spatial variation of $m_c$ originating the stricture of the seismic network.
+This advantage may arise from a the measurement artifact of spatial variation of $m_c$ originating from the properties of the seismic network.
 
-In order to acount for such spatially dependent artifacts we preform a second calculation.
-We once again condition MAGNET's results on a threshold according to \ref{eq:conditioned_likelihood}, this time choosing $\Tilde{m}=m_c(x,y)$, the spatial incompleteness.
-We calculate $m_c(x,y)$ for a grid of coordinates spaced every $0.1^\circ$.
-The calculation is performed for all events within a square centered at each grid point, with a side length of $0.2^\circ$.
+In order to account for such spatially dependent artifacts we preform a simialrecond calculation.
+We once again condition MAGNET's results on a threshold according to Eq.
+\ref{eq:conditioned_likelihood}, this time choosing a spatially varying completeness $\Tilde{m}=m_c(x,y)$.
+We estimate $m_c(x,y)$ for a grid of coordinates spaced every $0.1^\circ$, for all events within a square centered at each grid point, with a side length of $0.2^\circ$.
+
 The spatially conditioned CIG curves are shown as the red dashed curve in Fig.
 \ref{fig:info_gain_over_time} for all three regions.
-Evidently, it is observed to be a steadily increasing function for the Southern California and Japan data sets.
-In the New Zealand data set the curve shows an increase, then plateaus after the first sequence of major events (the 2016 Te Araroa and Kaikōura earthquakes).
+It is a steadily increasing function for the Southern California and Japan data sets.
+In the New Zealand data set the curve shows an initial increase followed by a plateau after the first sequence of major events (the 2016 Te Araroa and Kaikōura earthquakes).
 This may indicate that a good portion of this information gain followed by these earthquakes originate in the spatial incompleteness artifact.
-Nevertheless, the overall score shows an advantage for our model for the test set.
+Nevertheless, the overall score still shows an advantage for our model over the test set.
 Fig.
-\ref{fig:info_gain_over_time}e-g show the spatial distribution of our MAGNET's advantage over the GR benchmark.
-All three regions exhibit a spatially diffuse pattern.
-This suggests that information is not primarily sourced from any one specific area.
+\ref{fig:info_gain_over_time}e-g show the spatial distribution of MAGNET's advantage over the GR benchmark.
+All three regions exhibit a spatially diffuse pattern, suggesting that information is not gained solely from any one specific area.
 
-\section{Discussion}
-\neri{YOHAI LET'S MEET TO WRITE CMON}
+\section{Implications and Outlook}
+In this work we have used neural based models to find statistical dependencies between earthquake magnitudes and seismic history.
+Since our model is provided with the known location and timing of an event and is only tasked with predicting its magnitude, we have separated out the question of rate and location of earthquakes, and directly probe whether their magnitude is history dependent.
+MAGNET shows robust and significant information gain, even after accounting for possible measurement artifacts such as temporal or spatially dependent incompleteness, compared against a host of benchmarks.
+Our results indicate that some information about the magnitude of a specific event can be extracted, on average, from the regional seismic history in the form of a simple hypo-center catalog.
+
+%A corollary follows this finding: for some earthquakes, the magnitude might be partially predetermined before the earthquake is initiated.
+
+If true, this means that the seismic patterns preceding large events are distinguishable, at least to some extent, from those preceding ``background'' activity.
+A similar claim was recently raised by Ben~-~Zion and Zaliapin\cite{ben-zion_localization_2020}, which show trends in statistical properties of the seismicity prior to fault-size events, though a direct linkage between the two observation requires further investigation.
+
+In the future, it will also be interesting to examine MAGNET's prediction on query coordinates in the vicinity (either in space or time) of true events, producing a space-time hazard map, which may be used to inform existing forecasting and early-warning systems.
+This can also be done by constructing a different model, trained to estimate the magnitude of an upcoming earthquake ahead of the origin time, examining how it depends on the lead time and distance from the event location.
+It would also be interesting to understand what are the statistical features of seismicity that our model identifies, which may be done using various explainability analyses~\cite{sturmfels_visualizing_2020, zhang_survey_2021, liu_interpretable_2023}.
+Potentially, ablation tests \cite{meyes_ablation_2019, sturmfels_visualizing_2020} can be applied to extract heuristic, more simplified dependencies between statistical trends in the seismic history to trends in the resulting PDF.
+
+\section{Possible affects on operational earthquake forecasting and prediction}
+Identifying precursory patterns in the seismic history can advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}.
+Nevertheless, discovering such signals with enough accuracy to prove useful for operational systems may still be further down the road.
+However, an immediate application of the model may be a quick estimation of magnitude for early warning systems.
+Such a system does not require additional data sources or assumptions on the model as the timing and location of the earthquake to be estimated are known or estimated.
+Another immediate application is to incorporate such a history-dependent magnitude distribution into point-process rate forecasting models, such as the epidemic type aftershock sequence model (ETAS) or similar \cite{ogata_statistical_1988, ogata_statistics_2017, dascher-cousineau_using_2023}.
+As discussed above, such models currently assume no statistical dependence of the magnitude.
+Since larger earthquakes produce more aftershock, a more accurate estimation of magnitudes will have a positive compounding effect on the resulting statistics.
+
+% \neri{I bet the reviewer will want us to talk about limitations and improvements to the model but maybe we can wait and see}
 
@@
     % The number here (34 is an example) is the number of references in the main bibliography, and thus defines the starting number of the first reference in the Methods and Extended Data bibliography.
-    \setcounter{enumiv}{34}
+    \setcounter{enumiv}{62}
 }{\oldendthebibliography}
@@
 Outputs of all encoders are concatenated and fed to a Fully Connected Neural Network (FCNN).
-The sizes of layers in the FCNN, as the number of layers themselves, are are optimized by cross validation.
-A $tanh$ activation is used, and a $\texttt{softplus}$ as an activation on the output layer.
+The sizes of layers in the FCNN, as the number of layers themselves, are optimized by cross validation.
+A $tanh$ activation is used for the hidden layers, and a $\texttt{softplus}$ as an activation on the output layer.
 
@@
 
-\subparagraph{Spatial incompleteness} Temporal incompleteness is calculated for a grid of coordinates, spaced by $\theta^\circ$ degrees, covering the relevant region.
-$\theta^\circ$ is set to 0.1 for Southern California and 0.5 for New Zealand and Japan data sets.
+\subparagraph{Spatial incompleteness} Spatial incompleteness is calculated for a grid of coordinates, spaced by $\theta^\circ$ degrees, covering the relevant region.
+$\theta$ is set to 0.1 for Southern California and 0.5 for New Zealand and Japan data sets.
 Calculation is done using train set's events.
@@
 The datasets analysed during the current study are available in the relevant references mentioned throught the article.
-The code for analysing the data is available in \textcolor{red}{\textbf{**link TBD**}}
+The code for analysing the data is available in \textcolor{red}{\textbf{**link TBA**}}
 
@@
 \section*{Acknowledgements}
-We thank Yehuda Ben-Zion, Assaf Inbal, Kelian Dascher-Cousineau, David Marsan and Eugenio Lippiello \neri{Add Zhuang and Ogata?
-I did discuss this with them...} for thoughtful discussions.
+We thank Yehuda Ben-Zion, Assaf Inbal, Kelian Dascher-Cousineau, David Marsan and Eugenio Lippiello, Jiancang Zhuang and Yosihiko Ogata.
+for thoughtful discussions.
 YBS is supported by ISF grant 1907/22 and by Google Gift grant.

### 2024-07-05 article.tex
small changes mostly typos

@@
 Our model is trained on seismic catalogs containing information about past earthquakes, including location, time, and magnitude.
-In some experiments, we also incorporate focal mechanism information, demonstrating improved performance.
+In some experiments, we also incorporate focal mechanism information.
 Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided.
@@
 MAGNET is composed of two main components.
-First, the catalog up to time $t$ (not including) is encoded by a long-short term memory (LSTM) units, which produces a latent representation of the seismic history \cite{hochreiter_long_1997}.
+First, the catalog up to time $t$ (not including) is encoded by long-short term memory (LSTM) units, which produces a latent representation of the seismic history \cite{hochreiter_long_1997}.
 Second, this representation is combined with a space-time query specifying the coordinates, $\bf{x}_i$, $t_i$, of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, $p_{\textbf{x}_i, t_i}(m)$, see Fig.
@@
 The combined information is fed into a fully-connected neural network that outputs a probability density function (PDF) of the predicted magnitude at that time and location.
-\textbf{b} Performance on Major Earthquakes: MAGNET's resulting PDFs for a selection of well-known major earthquakes worldwide.
+\textbf{b} Performance on major earthquakes: MAGNET's resulting PDFs for a selection of well-known major earthquakes worldwide.
 The red curve represents the model's predicted PDF and the dashed black curve represents the Gutenberg-Richter (GR) magnitude-frequency distribution, a common naive benchmark.
@@
 
-All metrics reported in this work were evaluated over a time span that is not included in training period.
+All metrics reported in this work were evaluated over a time span that is not included in the training period.
 The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating a significant information gain in forecasting earthquake magnitudes prior to their occurrence.
@@
  $\mathcal{L}$ is the average minus log-likelihood, which the model is trained to minimize.
-The values of $\ell_i$ for MAGNET and the leading benchmark are presented in Extended Material Fig.
+The values of $\ell_i$ for MAGNET and the leading benchmark are presented in Extended Material (EM) Fig.
 \ref{fig:labels_likelihood}.
@@
 This can be easily calculated from the prediction, as the probability that $m_i>m_t$ is simply $\int_{m_t}^{\infty}p(m')dm'$.
-This task, asking whether a given event will be a large earthquake, is more interpretable and can be evaluated using standard binary classification metrics such as area under receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010} which suit better an imbalanced data set such as ours.
-Fig.~\ref{fig:metrics}d-f and g-i show these metrics for all three regions in examined and compare against the GR-variant models.
+This task, asking whether a given event will be a large earthquake, is more interpretable and can be evaluated using standard binary classification metrics such as area under the receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010} which suit better an imbalanced data set such as ours.
+Fig.~\ref{fig:metrics}d-f and g-i show these metrics for all three regions examined and compare against the GR-variant models.
 Again, MAGNET outperforms all benchmarks.
@@
 Scores conditioned above the temporal incompleteness are marked by the horizontal dashed line on each corresponding bar.
-\textbf{d, e, f}, Receiver Operating Characteristic (ROC) and \textbf{g, h, i} the interpolated precision-recall (PR) curve for a binary classifier determining the existence of a large ($m>=4$) event.
+\textbf{d, e, f}, Receiver Operating Characteristic (ROC) and \textbf{g, h, i} the interpolated precision-recall (PR) curve for a binary classifier determining the next event will be large ($m>=4$).
 The performance of such a classifier can be quantified by the area under the curve (AUC), noted in each frame, color coded identically to the bar plots.
@@
 a}, Information gain of individual events in the Southern California test set.
-Scattered dots indicate magnitude at event index, colored by the information gain per event (scale indicated by colorbar), here warm is an advantage for MAGNET, while cool is an advantage for the benchmark.
-Secondary horizontal axis (grey) indicates corresponds origin time.
+Scattered dots indicate magnitude at event index, colored by the information gain per event (scale is indicated by colorbar), here warm is an advantage for MAGNET, while cool is an advantage for the benchmark.
+Secondary horizontal axis (grey) indicates the corresponding origin time.
 Cumulative Information gain (CIG) (dashed blue), temporally conditioned CIG (dashed yellow) and spatially conditioned CIG (dashed red) curves are superimposed on the scatter, demonstrating a constantly increasing function.
-\textbf{b-d}, Information gain (dashed blue) temporally conditioned CIG (dashed yellow) and spatially conditioned CIG (dashed red) curves for Southern California, New Zealand and Japan test sets, respectively, all three regions examined in this study.
+\textbf{b-d}, Information gain (dashed blue) and temporally and spatially conditioned CIG (dashed yellow and red respectively) curves for Southern California, New Zealand and Japan test sets, respectively, all three regions examined in this study.
 An exception from the increasing trend can be observed for the spatially conditioned CIG of New Zealand (red curve in c).
@@
 Specifically, we show an average information gain of $>0.1$ nats per earthquake in the same region in Japan where Ogata et.~al.~\cite{ogata_exploring_2018} found no information gain.
-Similar plots for the other studied regions are shown in Extended Material Fig.~\ref{fig:nz_japan_info_gain}.
+Similar plots for the other studied regions are shown in EM Fig.~\ref{fig:nz_japan_info_gain}.
 
@@
 
-An obvious such artifact is ``short-time incompleteness'' (STI): the fact that immediately following a large event the signal of subsequent small events might be buried in the coda of the mainshock~\cite{kagan_short-term_2004, stockman_forecasting_2023}.
+One such artifact is ``short-time incompleteness'' (STI): the fact that immediately following a large event the signal of subsequent small events might be buried in the coda of the mainshock~\cite{kagan_short-term_2004, stockman_forecasting_2023}.
 Thus, immediately after a large event it is less likely that a small event will be included in the catalog, not because it did not occur, but rather because it was not detected.
@@
 
-The instantaneous completeness magnitude, $m_c(t)$, is defined as the minimal magnitude above which all earhtquakes can be detected.
+The instantaneous completeness magnitude, $m_c(t)$, is defined as the minimal magnitude above which all earthquakes can be detected.
 If we choose $\tilde m \geq m_c$ then, by definition, the model cannot present an information gain due to the STI effect.
@@
 We thus take $\Tilde{m}=m_c(t)$, which we estimate dynamically using the maximum curvature method \cite{wiemer_minimum_2000} within a window of 150 past and 150 current and future events.
-The resulting $m_c(t)$ curves are presented in the Extended Material Fig.
+The resulting $m_c(t)$ curves are presented in the EM Fig.
 \ref{fig:temp_incompleteness}.
@@
 Indeed, factoring out the temporal incompleteness reduces the information gain, though it is still significant: evidently, the conditioned CIG curves show a steadily increasing trend.
-the horizontal dashed lines on the bars in Fig \ref{fig:metrics}a-c indicate the mean conditioned scores, also presented in Table \ref{tab:mean_ll_all_benchmarks}.
+The horizontal dashed lines on the bars in Fig \ref{fig:metrics}a-c indicate the mean conditioned scores, also presented in Table \ref{tab:mean_ll_all_benchmarks}.
 EM Table \ref{tab:mean_ll_benchmarks_main_text} displays the conditioned scores for all benchmarks tested in this study.
@@
 Similarly to the temporal domain, a spurious information gain may also be caused by temporally constant but spatially varying properties.
-For example, fault stresses, geometry and lithology may change in space resulting in spatially-dependent GR distribution \cite{amitrano_brittle-ductile_2003, scholz_stress_2015, herrmann_revealing_2022, taroni_earthquake_2023}, and the local completeness magnitudes is also generally spatially dependent.
+For example, fault stresses, geometry and lithology may change in space resulting in a spatially-dependent GR distribution \cite{amitrano_brittle-ductile_2003, scholz_stress_2015, herrmann_revealing_2022, taroni_earthquake_2023}, and the local completeness magnitude is also generally spatially dependent.
 In such a case, the information gain of MAGNET does not imply that the magnitudes are correlated in time.
@@
 
-In order to account for such spatially dependent artifacts we preform a simialrecond calculation.
+In order to account for such spatially dependent artifacts we preform a simialr calculation.
 We once again condition MAGNET's results on a threshold according to Eq.
@@
 
-\section{Possible affects on operational earthquake forecasting and prediction}
+\section{Possible effects on operational earthquake forecasting and prediction}
 Identifying precursory patterns in the seismic history can advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}.
@@
 The sizes of layers in the FCNN, as the number of layers themselves, are optimized by cross validation.
-A $tanh$ activation is used for the hidden layers, and a $\texttt{softplus}$ as an activation on the output layer.
+A $\texttt{tanh}$ activation is used for the hidden layers, and a $\texttt{softplus}$ as an activation on the output layer.
 
@@
 \subparagraph{Last $n$ events} The Gutenberg- Richter (GR) distribution fitted for the past $n$ events.
-This method follows the method presented in Gulia \& Wiemer\cite{gulia_real-time_2019} for constructing a b-value time-series.
+This method follows Gulia \& Wiemer\cite{gulia_real-time_2019} for constructing a b-value time-series.
 If in the last $n$ events there are less than 10 earthquakes of magnitudes above $m_c^{train}$, then more events from the past are included, until this condition is met.
@@
 $b$ for a specified location $x_i$ is calculated by the average of all $b$ values in bins within 30km around $x_i$.
-Estimations which consider less than 150 events are discarded.
+Estimates which consider less than 150 events are discarded.
 
@@
 \section*{Acknowledgements}
-We thank Yehuda Ben-Zion, Assaf Inbal, Kelian Dascher-Cousineau, David Marsan and Eugenio Lippiello, Jiancang Zhuang and Yosihiko Ogata.
+We thank Yehuda Ben-Zion, Assaf Inbal, Kelian Dascher-Cousineau, David Marsan, Eugenio Lippiello, Jiancang Zhuang and Yosihiko Ogata.
 for thoughtful discussions.
@@
     \centering
-    \includegraphics[width=1\textwidth]{figures/detailed_architecture.pdf}
+    \includegraphics[width=1\textwidth]{figures/detailed_architecture.png}
     \caption{\textbf{MAGNET architecture.} Described in detail in the methods section.}
@@
 \textbf{c} The entire Japan data set and a focused time span \textbf{d} of the 2011 Tōhoku earthquake and some of its aftershocks.
-Low magnitude data points in a, c and d are subsampled to ease on image rendering.
          }
@@
     \caption{
-    \textbf{Temporal Incompleteness for test sets of all regions examined.} Calculated temporal incompleteness for test sets is presented in red, superimposed over the events above (black) and below (grey) the completeness magnitude of the train set, $m_c^{(train)}$.
+    \textbf{Temporal Incompleteness for test sets of all regions examined.} Calculated temporal incompleteness for test sets is presented in red, superimposed over the events above (black, included in model evaluation) and below (grey) the completeness magnitude of the train set, $m_c^{(train)}$.
 Events are plotted by the index of their appearance in the test set.
-Events of low magnitude, $m_c^{(train)}$ (grey), that are not included in the evaluation, are plotted at their interpolated location along the horizontal axis.
 Data for Japan (c) is displayed only near the 2011 Tōhoku earthquake to reduce data density and enhance clarity of the figure.
-Low magnitude data points in b and c are subsampled to ease on image rendering.
     }
@@
 The blue dashed line indicates the values that the GR benchmark would obtain.
-Low magnitude data points in b and c are subsampled to ease on image rendering.
+In this representation, a higher value indicates a better score for an individual event.
+Both MAGNET and 300 last events benchmark to be well above the GR performance.
+It can be seen that for all regions MAGNET spreads the likelihoods wider, and specifically clearly outperforms the 300 last events benchmark in a significant portion of events.
     }

### 2024-07-09 article.tex
changes in phrasing, update intro fig

@@
 \usepackage{caption}
+\usepackage[left]{lineno}
+\linenumbers
 \bibliographystyle{unsrt}
@@
 This is due both to the lack of information about fault conditions and to the inherent complexity of rupture dynamics.
-Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter distribution, which describes the marginal distribution over large regions and long times.
+Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter (GR) distribution, which describes the marginal distribution over large regions and long times.
 This approach implicitly assumes that earthquake magnitudes are independent of seismic history and are identically distributed.
 In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history.
-We present a neural network-based model for probabilistic forecasting of future magnitudes based on cataloged properties: hypocenter locations, occurrence times and magnitudes of past earthquakes.
+We present MAGNET - MAGnitude Neural EsTimation model, an open-source, geophysically-inspired neural-network model for probabilistic forecasting of future magnitudes from cataloged properties: hypocenter locations, occurrence times and magnitudes of past earthquakes.
 Our history-dependent model outperforms stationary and quasi-stationary state of the art GR-based benchmarks, in real catalogs in Southern California, Japan and New-Zealand.
@@
 This philosophy also inspired physical models which describe earthquake dynamics with a statistical approach \cite{olami_self-organized_1992, sornette_self-organized_1989, bak_earthquakes_1989, de_geus_scaling_2022}.
-These models reproduce many of the phenomenological statistical relations by positing that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition.
+These models reproduce many of the phenomenological statistical relations by positing that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically.
+The magnitudes then follow a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition.
 Under this paradigm, determining the magnitude of an event is indeed impossible using only far-field measurements.
@@
 Some studies have shown circumstantial evidence that earthquake magnitudes may not follow a strictly stationary distribution, e.g.~by showing that statistics are not invariant to permutation of the order, or by showing that the maximal magnitude in a given time frame can be predicted with a better-than-random chance~\cite{xiong_seismic_2023, corral_comment_2005, spassiani_exploring_2016, lippiello_positive_2024, lippiello_positive_2024-1, lippiello_influence_2008, shcherbakov_forecasting_2019, panakkat_neural_2007}.
-However, a significant information gain over the GR benchmark in predicting the magnitude of a given earthquake was not demonstrated, and to the best of our knowledge was not attempted.
+However, a significant information gain over the GR benchmark in predicting the magnitude of a given earthquake was not demonstrated.
 The debate regarding the dependence of magnitude on seismic history remains open.
@@
 
-% \neri{Stress more: this will imply we have information relevant to prediction tasks. Zhuang stated: "you have touched one of the most important problems on earthquake predictability"}
-
-To this end, we construct a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence.
+To this end, we construct MAGnitude Neural EsTimation model, \nobreak MAGNET, a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence.
 Our model is trained on seismic catalogs containing information about past earthquakes, including location, time, and magnitude.
@@
 
-% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice.
-On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff).
-These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition.
-Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics.
-In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).
-
 \section{Methodology}
-We construct the MAGnitude Neural EsTimation model, MAGNET, a generative neural network (NN) that receives as input a hypocentral catalog of past regional seismicity, and the time and location of a future event (a ``query'').
+MAGNET is a generative neural network (NN) that takes as input a hypocentral catalog of regional seismicity, and the time and location of a future event (a ``query'').
 The network produces a probability density function (PDF) estimating its magnitude.
@@
 MAGNET is composed of two main components.
-First, the catalog up to time $t$ (not including) is encoded by long-short term memory (LSTM) units, which produces a latent representation of the seismic history \cite{hochreiter_long_1997}.
+First, the catalog up to time $t$ (excluding $t$) is encoded by a series of geophysically inspired functions and long-short term memory (LSTM) units, which produces a latent representation of the seismic history \cite{hochreiter_long_1997}.
 Second, this representation is combined with a space-time query specifying the coordinates, $\bf{x}_i$, $t_i$, of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, $p_{\textbf{x}_i, t_i}(m)$, see Fig.
@@
 
-We deployed MAGNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog \cite{hauksson_waveform_2012} for Southern California, GeoNet \cite{gns_geonet_1970} for New Zealand, and the JMA catalog \cite{noauthor_japan_nodate} for Japan.
+We applied MAGNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog \cite{hauksson_waveform_2012} for Southern California, GeoNet \cite{gns_geonet_1970} for New Zealand, and the JMA catalog \cite{noauthor_japan_nodate} for Japan.
 While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies and exhibit varying data quality. 
 A separate model is trained for each region, with identical loss function and parameterization of the PDF.
-The resulting PDF is presented in Fig.
-\ref{fig:intro_fig}b for a few examples of major events, superimposed on the stationary GR distribution (fitted on the train set), which is the naive benchmark.
+All metrics reported in this work were evaluated over a time span that is not included in the training period.
+
+The resulting PDF is presented in Fig. \ref{fig:intro_fig}b for a few examples of major events, superimposed on the stationary GR distribution (fitted on the train set), which is the naive benchmark. 
+It can be seen that for most presented earthquakes, the likelihood of the observed magnitude is higher for MAGNET than for the GR benchmark, and in some cases MAGNET's prediction differs qualitatively as well as quantitatively.
+The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating a significant information gain in forecasting earthquake magnitudes prior to their occurrence.
 
@@
 Seismic history is continuously encoded using a Long Short-Term Memory neural network.
-For each earthquake, the encoded history up to its occurrence time ($t_i$) is concatenated with the time and location of the earthquake (``spacetime query'').
+For each earthquake, the encoded history up to its occurrence time ($t_i$) is concatenated with the time and location of the earthquake (``space~-~time query'').
 The combined information is fed into a fully-connected neural network that outputs a probability density function (PDF) of the predicted magnitude at that time and location.
-\textbf{b} Performance on major earthquakes: MAGNET's resulting PDFs for a selection of well-known major earthquakes worldwide.
+In turn, the output is compared to a benchmark model (e.g.
+GR in dashed line) \textbf{b} Performance on major earthquakes: MAGNET's resulting PDFs for a selection of well-known major earthquakes worldwide.
 The red curve represents the model's predicted PDF and the dashed black curve represents the Gutenberg-Richter (GR) magnitude-frequency distribution, a common naive benchmark.
@@
 \end{figure}
-
-All metrics reported in this work were evaluated over a time span that is not included in the training period.
-The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating a significant information gain in forecasting earthquake magnitudes prior to their occurrence.
-%This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation.
-Examining these results across multiple catalogs and regions highlights the robustness of our methodology and the generalizability of our conclusion, paving the way for further exploration of earthquake predictability.
 
@@
 This is a qualitative demonstration that the model works as expected.
-For a quantitative comparison, we calculate the $\mathcal{L}=-\langle \ell_i \rangle$ over all events in the test set, and benchmark it against state-of-the-art models:
+For a quantitative comparison, we calculate $\mathcal{L}=-\langle \ell_i \rangle$ over all events in the test set, and benchmark it against state-of-the-art models:
 the stationary Gutenberg-Richter (GR) distribution (fitted over the training set), and a few variants of a moving window GR approach~\cite{gulia_real-time_2019}.
@@
 We also observe that MAGNET's average prediction aligns well with the \emph{test set} GR distribution, rather than the training set, suggesting its ability to generalize beyond the training data.
-This observation is noticeable even more so in the New Zealand and Japan data sets presented in Extended Data Fig.
+This observation is more pronounced in the New Zealand and Japan data sets presented in Extended Data Fig.
 \ref{fig:model_output_em}.
@@
 
-\section{Interpretation and caveats}
-At this point, it is clear that MAGNET is able to extract information about the magnitude of specific earthquakes, prior to their occurrence.
-However, such an information gain may potentially result not from the fact that the underlying physical process is predictable (at least to some extent), but from various measurement and modeling artifacts.
+\section{Interpretation and potential caveats}
+At this point, it is clear that MAGNET achieves an information gain about the magnitude of specific earthquakes, prior to their occurrence.
+However, such a gain may potentially result not from the fact that the underlying physical process is predictable (at least to some extent), but from various measurement and modeling artifacts.
 
@@
 
-Similarly to the temporal domain, a spurious information gain may also be caused by temporally constant but spatially varying properties.
+Similarly to the temporal domain, a spurious information gain may also be caused by temporally constant but spatially varying statistical properties.
 For example, fault stresses, geometry and lithology may change in space resulting in a spatially-dependent GR distribution \cite{amitrano_brittle-ductile_2003, scholz_stress_2015, herrmann_revealing_2022, taroni_earthquake_2023}, and the local completeness magnitude is also generally spatially dependent.
 In such a case, the information gain of MAGNET does not imply that the magnitudes are correlated in time.
+
 To examine this, we preform two additional tests.
-First, we construct a spatially dependent GR distributions where $m_c$ is held constant ($=m_c^{\mbox{\footnotesize (train set)}}$).
-MAGNET outperforms the spatially varying GR benchmark for all regions in the unconditioned $\mathcal{L}$ score, as can be seen in EM Table \ref{tab:mean_ll_all_benchmarks}.
-The spatially varying GR shows an advantage over MAGNET for solely the New Zealand data set for the temporally conditioned $\mathcal{L}$.
-This advantage may arise from a the measurement artifact of spatial variation of $m_c$ originating from the properties of the seismic network.
+First, we construct a spatially dependent GR distribution which accounts for the local varation in seismic statistics.
+We use the method proposed by \cite{taroni_highdefinition_2021}, as detailed in the Methods section.
+The spatially dependent GR shows some information gain over plain-vanilla GR in New Zealand and Japan, but not in California, see EM Table \ref{tab:mean_ll_all_benchmarks}.
+In any case, this information gain is always smaller than MAGNET's, with the sole exception of the temporally-conditioned test in New Zealand.
 
-In order to account for such spatially dependent artifacts we preform a simialr calculation.
-We once again condition MAGNET's results on a threshold according to Eq.
-\ref{eq:conditioned_likelihood}, this time choosing a spatially varying completeness $\Tilde{m}=m_c(x,y)$.
-We estimate $m_c(x,y)$ for a grid of coordinates spaced every $0.1^\circ$, for all events within a square centered at each grid point, with a side length of $0.2^\circ$.
+Second, simliar to the temporal analysis, we calculate the information gain conditioned on the locally-varying completness magnitude.
+Specifically, we use Eq.
+\ref{eq:conditioned_likelihood}, with a spatially varying completeness $\Tilde{m}=m_c(x,y)$, estimated for a coordinate grid of $0.1^\circ$, see Methods section for details.
 
-The spatially conditioned CIG curves are shown as the red dashed curve in Fig.
-\ref{fig:info_gain_over_time} for all three regions.
+The spatially conditioned CIG are shown by the red dashed curves in Fig.
+\ref{fig:info_gain_over_time}.
 It is a steadily increasing function for the Southern California and Japan data sets.
 In the New Zealand data set the curve shows an initial increase followed by a plateau after the first sequence of major events (the 2016 Te Araroa and Kaikōura earthquakes).
-This may indicate that a good portion of this information gain followed by these earthquakes originate in the spatial incompleteness artifact.
+This may indicate that a good portion of this information gain followed by these earthquakes does originate from spatially inhomogeneity, as also hinted by the spatially-varying GR benchmark.
 Nevertheless, the overall score still shows an advantage for our model over the test set.
 Fig.
-\ref{fig:info_gain_over_time}e-g show the spatial distribution of MAGNET's advantage over the GR benchmark.
+\ref{fig:info_gain_over_time}e-g shows the spatial distribution of MAGNET's advantage over the GR benchmark.
 All three regions exhibit a spatially diffuse pattern, suggesting that information is not gained solely from any one specific area.
@@
 Since our model is provided with the known location and timing of an event and is only tasked with predicting its magnitude, we have separated out the question of rate and location of earthquakes, and directly probe whether their magnitude is history dependent.
-MAGNET shows robust and significant information gain, even after accounting for possible measurement artifacts such as temporal or spatially dependent incompleteness, compared against a host of benchmarks.
+MAGNET shows robust and significant information gain, even after accounting for possible measurement artifacts such as temporal or spatially dependent properties compared against a host of benchmarks, thereby answering an open question in earthquake predictability.
 Our results indicate that some information about the magnitude of a specific event can be extracted, on average, from the regional seismic history in the form of a simple hypo-center catalog.
 
-%A corollary follows this finding: for some earthquakes, the magnitude might be partially predetermined before the earthquake is initiated.
+This means that the seismic patterns preceding large events are distinguishable, at least to some extent, from those preceding ``background'' activity.
+A similar claim was recently raised by Ben~-~Zion and Zaliapin\cite{ben-zion_localization_2020}, which show trends in statistical properties of the seismicity prior to fault-size events, though a direct linkage between the two observation requires further investigation.
+In general, it would also be interesting to understand what are the statistical features of seismicity that our model identifies, which may be done using various explainability analyses~\cite{sturmfels_visualizing_2020, zhang_survey_2021, liu_interpretable_2023}.
 
-If true, this means that the seismic patterns preceding large events are distinguishable, at least to some extent, from those preceding ``background'' activity.
-A similar claim was recently raised by Ben~-~Zion and Zaliapin\cite{ben-zion_localization_2020}, which show trends in statistical properties of the seismicity prior to fault-size events, though a direct linkage between the two observation requires further investigation.
+Understanding such patterns will advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}, though it is still further down the road.
+However, even as a ``black box'', such a model may be integrated into operational systems such early warning systems, which already provide a quick estimation of the source time and location. 
+Another possible application is incorporating a history-dependent magnitude distribution into point-process rate forecasting models, such as the epidemic type aftershock sequence model (ETAS) or similar \cite{ogata_statistical_1988, ogata_statistics_2017, dascher-cousineau_using_2023}.
+As discussed above, such models currently assume no statistical dependence of the magnitude.
+Since larger earthquakes produce more aftershocks, a more accurate estimation of magnitudes will have a positive compounding effect on the resulting statistics.
 
-In the future, it will also be interesting to examine MAGNET's prediction on query coordinates in the vicinity (either in space or time) of true events, producing a space-time hazard map, which may be used to inform existing forecasting and early-warning systems.
-This can also be done by constructing a different model, trained to estimate the magnitude of an upcoming earthquake ahead of the origin time, examining how it depends on the lead time and distance from the event location.
-It would also be interesting to understand what are the statistical features of seismicity that our model identifies, which may be done using various explainability analyses~\cite{sturmfels_visualizing_2020, zhang_survey_2021, liu_interpretable_2023}.
-Potentially, ablation tests \cite{meyes_ablation_2019, sturmfels_visualizing_2020} can be applied to extract heuristic, more simplified dependencies between statistical trends in the seismic history to trends in the resulting PDF.
-
-\section{Possible effects on operational earthquake forecasting and prediction}
-Identifying precursory patterns in the seismic history can advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}.
-Nevertheless, discovering such signals with enough accuracy to prove useful for operational systems may still be further down the road.
-However, an immediate application of the model may be a quick estimation of magnitude for early warning systems.
-Such a system does not require additional data sources or assumptions on the model as the timing and location of the earthquake to be estimated are known or estimated.
-Another immediate application is to incorporate such a history-dependent magnitude distribution into point-process rate forecasting models, such as the epidemic type aftershock sequence model (ETAS) or similar \cite{ogata_statistical_1988, ogata_statistics_2017, dascher-cousineau_using_2023}.
-As discussed above, such models currently assume no statistical dependence of the magnitude.
-Since larger earthquakes produce more aftershock, a more accurate estimation of magnitudes will have a positive compounding effect on the resulting statistics.
-
-% \neri{I bet the reviewer will want us to talk about limitations and improvements to the model but maybe we can wait and see}
+% In light of the foregoing, our study demonstrates the possibility of extracting information about the magnitude of an upcoming earthquake from seismic history, thereby answering an open question in earthquake predictability.
+Our work has been put to test against multiple benchmark models and across multiple regions worldwide with various seismogenic properties, demonstrating our conclusions as general.
+Our model is openly available.
+To this end, our model has the potential not only to be used for scientific inquiry, but also to be put to use in operational means.
+We believe it may further advance our understanding and monitoring of earthquakes.
 

### 2024-07-30 article.tex
Final version before submission

@@
 
-\title[Running Title]{Does an earthquake “know” how big it will be? A neural-net aided study} % 2nd option: Using past seismicity to predict the magnitude of future earthquakes
+\title[Running Title]{Does an earthquake “know” how big it will be? A neural-net aided study}
 
@@
 \author[2]{\fnm{Oren} \sur{Gilon}}
+\author[2]{\fnm{Yossi} \sur{Matias}}
 \author[1]{\fnm{Yohai} \sur{Bar-Sinai}}\email{ybarsinai@gmail.com}
@@
 Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter (GR) distribution, which describes the marginal distribution over large regions and long times.
-This approach implicitly assumes that earthquake magnitudes are independent of seismic history and are identically distributed.
+This approach implicitly assumes that earthquake magnitudes are independent of previous seismicity and are identically distributed.
 In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history.
@@
 Thus, we separate out the task of modeling nucleation statistics and isolate the question of magnitude predictability.
-If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of a specific earthquake is extractable from cataloged properties alone.
+If our model performs better than a draw from the GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of a specific earthquake is extractable from cataloged properties alone.
+An example of MAGNET's performance on a few select earthquakes is shown in Fig \ref{fig:intro_fig}a.
+For instance, MAGNET improved the likelihood of estimating the true magnitude for the Ridgecrest mainshock in Southern California and the Kaikōura event in New Zealand by factors of approximately 13 and 4, respectively, over the naive GR predictor.
 
@@
 This can be easily calculated from the prediction, as the probability that $m_i>m_t$ is simply $\int_{m_t}^{\infty}p(m')dm'$.
-This task, asking whether a given event will be a large earthquake, is more interpretable and can be evaluated using standard binary classification metrics such as area under the receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010} which suit better an imbalanced data set such as ours.
+This task, asking whether a given event will be a large earthquake, is more interpretable and can be evaluated using standard binary classification metrics such as area under the receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010}.
 Fig.~\ref{fig:metrics}d-f and g-i show these metrics for all three regions examined and compare against the GR-variant models.
@@
 One such artifact is ``short-time incompleteness'' (STI): the fact that immediately following a large event the signal of subsequent small events might be buried in the coda of the mainshock~\cite{kagan_short-term_2004, stockman_forecasting_2023}.
-Thus, immediately after a large event it is less likely that a small event will be included in the catalog, not because it did not occur, but rather because it was not detected.
+Thus, immediately after a large event, smaller events are less likely to be included in the catalog, not because they did not occur, but rather because they were not detected.
 Indeed, our model consistently allocates less probability mass to low magnitudes shortly after large events (see the Tohoku aftershock in Fig.~\ref{fig:intro_fig}b), and some of the information gain may be attributed to this effect.
@@
 \ref{fig:info_gain_over_time}e-g shows the spatial distribution of MAGNET's advantage over the GR benchmark.
-All three regions exhibit a spatially diffuse pattern, suggesting that information is not gained solely from any one specific area.
+All three regions exhibit a spatially diffuse pattern, suggesting that information is not gained solely from any one specific area (compare EM Fig.~\ref{fig:seismicity_grid} for the distribution of seismicity).
 
@@
 MAGNET shows robust and significant information gain, even after accounting for possible measurement artifacts such as temporal or spatially dependent properties compared against a host of benchmarks, thereby answering an open question in earthquake predictability.
-Our results indicate that some information about the magnitude of a specific event can be extracted, on average, from the regional seismic history in the form of a simple hypo-center catalog.
+Our results indicate that some information about the magnitude of a specific event can be extracted, on average, from the regional seismic history in the form of a hypo-center catalog.
 
@@
 Understanding such patterns will advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}, though it is still further down the road.
-However, even as a ``black box'', such a model may be integrated into operational systems such early warning systems, which already provide a quick estimation of the source time and location. 
+Additionally, such models may be incorporated in early warning systems, which already provide a quick (though not accurate) estimation of the source time and location. 
 Another possible application is incorporating a history-dependent magnitude distribution into point-process rate forecasting models, such as the epidemic type aftershock sequence model (ETAS) or similar \cite{ogata_statistical_1988, ogata_statistics_2017, dascher-cousineau_using_2023}.
@@
 \section*{Author Contributions}
-N.B., O.Z., O.G.
+N.B., O.Z., O.G., Y.M.
 and Y.B.S.
@@
 
+\newpage
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/seismicity_maps.pdf}
+    \caption{
+    \textbf{Event count in examined regions}.
+Number of events per bin is indicated in the adjacent color bar per inset.
+Bin divisions are identical to those presented in Figures \ref{fig:info_gain_over_time}e-g.
+    }
+    \label{fig:seismicity_grid}
+\end{figure}
+
 \end{document}
